---
title: "CRC-850K-Pipeline"
author: "YC"
date: "2026-06-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  results = "hide",  
  message = FALSE,   
  warning = FALSE   
)
```

## data preprocessing
```{r standard beta value matrix and pd group information-ChAMP filtering}
rm(list = ls())
options(stringsAsFactors = F)
require(GEOquery)
require(Biobase)
library("impute")
library(ChAMP)
library(data.table)

eset <- getGEO("GSE199057",destdir = './',AnnotGPL = T,getGPL = F)
## Modify this part when applying the pipeline to another dataset
pD.all <- pData(eset[[1]])
pD <- pD.all[,c(1,2)]
names(pD) <- c("group", "ID")
pD$group1 <- ifelse(
  grepl("normal colon tissue from CRC patient",pD$group,ignore.case = T),
  "NAT",
  ifelse(grepl("normal colon tissue from non-CRC patient", pD$group,ignore.case = T),"healthy colon",
         "Tumor"))
pD <- pD[,-1]
pD$group <-pD$group1
pD <- pD[,-2]
#⭐Check if the extracted groupings are correct
check_df <- data.frame(Title = pD.all$title, Group = pD$group)
#head(check_df, 10)
#tail(check_df,10)
table(pD$group)
## Prepare beta value matrix
#beta.m <- exprs(eset[[1]])#the easiest way
beta.m <- fread("GSE199057_processed_methlyation_matrix.txt.gz",data.table = F)
dim(beta.m)
beta_cols <- grep("\\.AVG_Beta$", colnames(beta.m), value = TRUE)
beta_sample_id <- sub("\\.AVG_Beta$", "", beta_cols)
head(beta_sample_id)
sample_map <- data.frame(
  GSM = pD.all$geo_accession,
  title = pD.all$title,
  stringsAsFactors = FALSE
)
sample_map$SampleID <- sub(".*\\[([^]]+)\\].*", "\\1", sample_map$title)
head(sample_map)
idx <- match(beta_sample_id, sample_map$SampleID)

beta_to_GSM <- data.frame(
  beta_column = beta_cols,
  SampleID = beta_sample_id,
  GSM = sample_map$GSM[idx],
  title = sample_map$title[idx],
  stringsAsFactors = FALSE
)

head(beta_to_GSM)
sum(is.na(beta_to_GSM$GSM))
beta_to_GSM[is.na(beta_to_GSM$GSM), ]

## In GSE199057, two technical replicates use hyphens in the processed matrix:
##   T1003588-rep1
##   T1003588-rep2
## But GEO metadata title uses spaces:
##   T1003588 rep1
##   T1003588 rep2
## Therefore, manually assign their GSM IDs.
sample_map[grepl("T1003588", sample_map$title, ignore.case = TRUE),
                       c("GSM", "SampleID", "title")]
beta_to_GSM$GSM[beta_to_GSM$SampleID == "T1003588-rep1"] <- "GSM5963372"
beta_to_GSM$title[beta_to_GSM$SampleID == "T1003588-rep1"] <- 
  "colon tumor sample from CRC patient [T1003588 rep1]"

beta_to_GSM$GSM[beta_to_GSM$SampleID == "T1003588-rep2"] <- "GSM5963373"
beta_to_GSM$title[beta_to_GSM$SampleID == "T1003588-rep2"] <- 
  "colon tumor sample from CRC patient [T1003588 rep2]"
## Recheck unmatched samples after manual correction.
sum(is.na(beta_to_GSM$GSM))
beta_to_GSM[is.na(beta_to_GSM$GSM), ]

## Plate-control samples are technical controls.
## They should not be included in Tumor vs NAT / healthy_colon analysis.
beta_to_GSM_clean <- beta_to_GSM[
  !grepl("Plate-Control", beta_to_GSM$SampleID, ignore.case = TRUE),
]
## Confirm that all retained samples have GSM IDs.
sum(is.na(beta_to_GSM_clean))

## Extract only AVG_Beta columns and convert to a numeric matrix.
beta <- as.matrix(beta.m[, beta_to_GSM_clean$beta_column, drop = FALSE])
mode(beta) <- "numeric"
rownames(beta) <- beta.m$TargetID## Set CpG probe IDs as row names.
colnames(beta) <- beta_to_GSM_clean$GSM## Set GSM IDs as column names.
dim(beta)#224 samples
nrow(beta_to_GSM_clean)
## Check which beta columns were removed.
## These should be plate-control columns only.
setdiff(beta_cols, beta_to_GSM_clean$beta_column)
# It returns all unique elements that are present in x but not in y.
range(beta, na.rm = TRUE)
anyNA(beta)

sum(is.na(beta))## Count missing beta values.
mean(is.na(beta)) * 100## Calculate missing-value percentage.
## Check whether missing values are concentrated in specific samples.
na_per_sample <- colSums(is.na(beta))
summary(na_per_sample)
sort(na_per_sample, decreasing = TRUE)[1:10]
na_rate_sample <- colMeans(is.na(beta)) * 100
sort(na_rate_sample, decreasing = TRUE)[1:10]

## Check whether missing values are concentrated in specific probes.
na_per_probe <- rowSums(is.na(beta))
summary(na_per_probe)
sum(na_per_probe > 0)

beta <- beta[rowSums(is.na(beta)) == 0, ]
## Save the cleaned beta matrix as beta.m1 for downstream ChAMP analysis.
beta.m1 <- beta
## Final NA check.
dim(beta.m1)
# The beta signal matrix must not contain any NA values
anyNA(beta.m1)

head(rownames(beta.m1))
head(colnames(beta.m1))

## Identify samples that exist in pD but not in beta.m1.
missing_in_beta <- setdiff(pD$ID, colnames(beta))
missing_in_beta
## Inspect these samples in the original GEO metadata.
## In GSE199057, these should be plate-control samples.
pD.all[pD.all$geo_accession %in% missing_in_beta, c("title", "geo_accession")]
# Back up the data first to avoid recovery issues in case of accidental deletion
pD.raw <- pD
# Remove these 5 plate-control samples from pD-not present in the beta matrix.
pD <- pD[!pD$ID %in% missing_in_beta, , drop = FALSE]
# Check dimensions after deletion
dim(pD)
ncol(beta)
nrow(pD)
## Check whether beta.m1 and pD contain the same sample set.
## setequal() ignores order.
identical(rownames(pD),colnames(beta.m1))
setequal(colnames(beta.m1), pD$ID)
# Confirm that there are no duplicated sample names
sum(duplicated(colnames(beta.m1)))
sum(duplicated(pD$ID))

# Reorder pD to match the column names of beta.m1
pD <- pD[match(colnames(beta.m1), pD$ID), , drop = FALSE]

# Set pD row names as GSM IDs.
rownames(pD) <- pD$ID
## Final strict alignment check.
dim(beta.m1)
dim(pD)
## This must be TRUE before running champ.filter().
identical(rownames(pD), colnames(beta.m1))
## Check final group numbers.
table(pD$group, useNA = "ifany")
# Set group order
pD$group <- factor(
  pD$group,
  levels = c("healthy colon", "NAT", "Tumor")
)

# Sort pD by group
sample_order <- order(pD$group)

pD <- pD[sample_order, , drop = FALSE]

# Set rownames as GSM ID
rownames(pD) <- pD$ID

# Reorder beta matrix columns according to pD
beta.m1 <- beta.m1[, rownames(pD), drop = FALSE]

# Final check
identical(rownames(pD), colnames(beta.m1))
table(pD$group)
## Run ChAMP filtering using the cleaned beta matrix and aligned pD.
myLoad <- champ.filter(beta = beta.m1, pd = pD, arraytype = "EPIC")
## Save the ChAMP object for downstream analysis.
save(myLoad,file = 'step1-output.Rdata')

# Both methods are designed to generate a ChAMP object
# For all downstream analyses, simply use this myLoad variable
#champ.filter() performs initial quality control:
#(1).Filters out probes with low detection p-value.
#(2).Removes probes with missing values.
#(3).Filters sex chromosomes or SNP-associated probes if needed.
#(4).The output myLoad is an object used for all downstream ChAMP analyses 
##(normalization, DMP, DMR, pathway analysis).
```

## Quality control using PCA heatmap correlation heatmap
```{r ChAMP normalisation & quality control}
rm(list = ls())
library("ChAMP")
library("minfi")

#load(file = 'step1-output.Rdata')
# This is a time-consuming step.
# Run it once, save the result, and then keep this block disabled.
if(F){
  load(file = 'step1-output.Rdata')
  myNorm <- champ.norm(beta=myLoad$beta,arraytype="850K",cores=5)
  dim(myNorm) 
  pD=myLoad$pd
  save(myNorm,pD,file = 'step2_champ_myNorm_2groups.Rdata')
}
load("step2_champ_myNorm_2groups.Rdata")

## Step 2. Quality control plot 1: PCA plot
# Convert beta matrix into a sample-by-CpG data frame.
# Original myLoad$beta: rows = CpGs, columns = samples.
# After transpose: rows = samples, columns = CpGs.
dat=as.data.frame(t(beta_new))
# Preview the first four rows and columns
dat[1:4,1:4]
# Load packages for PCA and visualization
library(FactoMineR)
library(factoextra) 
# Perform PCA without plotting automatically
dat.pca <- PCA(dat, graph = FALSE)
Group <- pd_new$group
# Generate PCA plot colored by group
p_pca <- fviz_pca_ind(dat.pca,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = Group, # color by groups
             palette = c("#00AFBB", "#E7B800","#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Groups"
)
p_pca
# Save PCA plot as PDF
ggsave(
  filename = "verificationPlots/Quality_Control_Plots/GSE199057_PCA_plot.pdf",
  plot = p_pca,
  width = 7,
  height = 5
)

##Quality control plot 2: heatmap of top 1000 most variable CpGs
library(pheatmap)
Group <- pd_new$group
# Calculate standard deviation for each CpG across samples,
# sort CpGs by variability, and select the top 1000 most variable CpGs.
# apply(myLoad$beta, 1, sd):
#   1 means applying the function by row, i.e. by CpG.
#   2 would mean applying the function by column, i.e. by sample.
g <- names(tail(sort(apply(beta_new, 1, sd)), 1000)) 
# Subset beta matrix to the top 1000 most variable CpGs
n <- beta_new[g, ]
# Create column annotation data frame for sample groups
annotation_col = data.frame(row.names = colnames(n),
                            Group = Group)

# Draw heatmap of the top 1000 most variable CpGs
pheatmap(n,
         show_colnames =F,
         show_rownames = F,
         annotation_col= annotation_col,
         scale = "row", #Standardize each CpG row to emphasize within-CpG #differences across samples
         breaks = seq(-3,3,length.out = 100),## Limit color scale to -3 to 3; #values outside this range use extreme colors
         cluster_cols = FALSE,
         main = "Top 1000 most variable CpGs",
         filename = "verificationPlots/Quality_Control_Plots/GSE199057_top1000_variable_CpGs_heatmap.png",
         width = 8,
         height = 7
) 

##Quality control plot 3: sample correlation heatmap
# The within-group sample similarity should generally be higher
# than between-group sample similarity.
  beta.m <- beta_new
  group_list <- pd_new$group
# Rename beta matrix as exprSet for compatibility with common expression-analysis #workflow
  exprSet=beta.m
# Draw a simple sample correlation heatmap
  pheatmap::pheatmap(cor(exprSet)) 
  
  colD=data.frame(group_list=group_list)
  rownames(colD)=colnames(exprSet)
# Draw a simple sample correlation heatmap
  pheatmap::pheatmap(cor(exprSet),
                     annotation_col = colD,
                     show_rownames = F,
                     cluster_cols = FALSE,
                     filename = 'verificationPlots/Quality_Control_Plots/GSE199057_cor_all.png')
  
  
# Select the top 500 CpGs with the highest median absolute deviation,
# which represent highly variable CpGs across samples
  exprSet=exprSet[names(sort(apply(exprSet, 1,mad),decreasing = T)[1:500]),]
  dim(exprSet)
  # M=cor(log2(exprSet+1)) 
  M=cor(exprSet)
  pheatmap::pheatmap(M,annotation_col = colD)
  pheatmap::pheatmap(M,
                     show_rownames = F,
                     annotation_col = colD,
                     cluster_cols = F,
                     filename = 'verificationPlots/Quality_Control_Plots/GSE199057_cor_top500.png')

## Additional QC plots using minfi-beta value density plot by sample
  library(ggplot2)
  
  beta_for_plot <- beta_new
  group_list <- pd_new$group
  
  names(group_list) <- rownames(pd_new)
  group_list <- group_list[colnames(beta_for_plot)]
# Set random seed for reproducibility
  set.seed(123)
  
  sample_cpgs <- sample(
    rownames(beta_for_plot),
    size = min(50000, nrow(beta_for_plot))
  )
  
  beta_sub <- beta_for_plot[sample_cpgs, ]
  
  density_df <- do.call(
    rbind,
    lapply(colnames(beta_sub), function(s) {
      d <- density(beta_sub[, s], na.rm = TRUE, from = 0, to = 1)
      data.frame(
        Sample = s,
        Beta = d$x,
        Density = d$y,
        Group = group_list[s]
      )
    })
  )
  
  p_density <- ggplot(
    density_df,
    aes(x = Beta, y = Density, group = Sample, color = Group)
  ) +
    geom_line(alpha = 0.25) +
    theme_bw() +
    labs(
      title = "Beta value density plot by sample",
      x = "Beta value",
      y = "Density"
    )
  
  ggsave(
    filename = "verificationPlots/Quality_Control_Plots/GSE199057_beta_density_by_sample_group.png",
    plot = p_density,
    width = 8,
    height = 5,
    dpi = 300
  )

```

### Additional QC plots using minfi-mds plot and densitybean plot
```{r save_mdsPlot, results='hide', fig.show='hide'}
library(minfi)
## mds plot
graphics.off()

beta.m <- as.matrix(beta_new)
storage.mode(beta.m) <- "numeric"

pd_new <- pd_new[colnames(beta.m), , drop = FALSE]
group_list <- factor(pd_new$group)

pdf_file <- "verificationPlots/Quality_Control_Plots/GSE199057_mdsPlot.pdf"

pdf(pdf_file, width = 8, height = 6)

tryCatch({
  mdsPlot(
    beta.m,
    numPositions = 1000,
    sampGroups = group_list,
    main = "MDS plot using top 1000 variable CpGs"
  )
}, finally = {
  if (dev.cur() > 1) dev.off()
})

```

### Additional QC plots using density bean plot
```{r save_densityBeanPlot, results='hide', fig.show='hide'}
library(minfi)

graphics.off()

pdf_file <- "verificationPlots/Quality_Control_Plots/GSE199057_densityBeanPlot.pdf"

pdf(pdf_file, width = 8, height = 6)

tryCatch({
  densityBeanPlot(
    beta.m,
    sampGroups = group_list,
    main = "Beta value density bean plot"
  )
}, finally = {
  if (dev.cur() > 1) dev.off()
})

```

## Hypermethylated DMRs from tumor vs Normal

```{r Hypermethylated DMRs from tumor vs Normal}
rm(list = ls())   
options(stringsAsFactors = F)
library("ChAMP")

load(file = 'step2_champ_myNorm_2groups.Rdata')
group_list <- pd_new$group
table(group_list)

group_list <- factor(group_list, levels = c("Normal", "Tumor"))
myDMR <- champ.DMR(beta = beta_new, 
                   pheno = group_list, 
                   method = "Bumphunter")

save(myDMR,file = 'myDMR-output-GSE199057.Rdata')
load(file = 'myDMR-output-GSE199057.Rdata')
# Extract DMR results
dmr_results <- myDMR$BumphunterDMR
# Keep only hypermethylated DMRs (CRC > normal)
k1 <- dmr_results$value > 0.2#value = βCRC-βnormal
k2 <- dmr_results$p.value <0.05
k3 <- dmr_results$L >=5
hyper_dmrs <- dmr_results [k1 & k2 & k3,]

# Save as Rdata file
save(hyper_dmrs, file = "hyper_DMRs_CRC_GSE199057.Rdata")

# Save as CSV (most common)
write.csv(hyper_dmrs, "verificationPlots/hypermethylated_DMRs_CRC_GSE199057.csv", row.names = FALSE)
```

## Independent validation in GSE119526
```{r Independent validation in GSE119526}
############################################################
# Step 4. Validate GSE199057 hyper-DMRs in GSE119526
############################################################
rm(list = ls())
options(stringsAsFactors = FALSE)

library(GenomicRanges)
library(IRanges)
library(minfi)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(ggplot2)
library(pheatmap)

############################################################
# 1. load and Check input objects
############################################################

# dD: 620 hyper-DMRs from GSE199057
# myNorm: normalized beta matrix from GSE119526
# pD: group vector for GSE119526 samples
dD <- read.csv("hypermethylated_DMRs_CRC_GSE199057.csv")
load("step2-GSE119526_champ_myNorm.Rdata")
dim(dD)
dim(myNorm)
length(pD)

head(dD)
head(rownames(myNorm))
head(colnames(myNorm))
table(pD$group)

############################################################
# 2. Prepare GSE119526 beta matrix and group list
############################################################

beta_119526 <- myNorm

# Make sure values are beta values between 0 and 1
range(beta_119526, na.rm = TRUE)

############################################################
# 2.1 Strictly align pD with beta_119526
############################################################

# pD must contain sample ID and group
pD$ID <- as.character(pD$ID)
pD$group <- as.character(pD$group)

# reorder beta matrix according to pD
beta_119526 <- beta_119526[, pD$ID, drop = FALSE]

# Critical check
identical(colnames(beta_119526), pD$ID)

############################################################
# 2.2 Build group_list safely
############################################################

group_list <- pD$group
names(group_list) <- pD$ID

group_list <- factor(group_list, levels = c("normal", "CRC"))

table(group_list)

# Final strict check
identical(names(group_list), colnames(beta_119526))

############################################################
# 3. Prepare discovery DMR coordinates
############################################################

# ChAMP Bumphunter output may use either chr or seqnames
if (!"chr" %in% colnames(dD) && "seqnames" %in% colnames(dD)) {
  dD$chr <- dD$seqnames
}

if (!all(c("chr", "start", "end") %in% colnames(dD))) {
  stop("dD must contain chr, start, and end columns.")
}

# Function to standardise chromosome names
fix_chr <- function(x) {
  x <- as.character(x)
  x <- gsub("^chr", "", x, ignore.case = TRUE)
  x[x == "23"] <- "X"
  x[x == "24"] <- "Y"
  x[x %in% c("M", "MT")] <- "M"
  paste0("chr", x)
}

dD$chr <- fix_chr(dD$chr)
dD$start <- as.numeric(dD$start)
dD$end <- as.numeric(dD$end)

dD <- dD[!is.na(dD$chr) & !is.na(dD$start) & !is.na(dD$end), ]

# Add DMR ID
dD$DMR_ID <- paste0("DMR_", seq_len(nrow(dD)))

# Keep discovery delta beta if available
if ("value" %in% colnames(dD)) {
  dD$deltaBeta_GSE199057 <- dD$value
}

############################################################
# 4. Load EPIC annotation and map CpGs to genomic coordinates
############################################################

anno <- minfi::getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
anno <- as.data.frame(anno)
anno$CpG <- rownames(anno)

# Keep only CpGs present in GSE119526 beta matrix
anno <- anno[anno$CpG %in% rownames(beta_119526), ]

anno$chr <- fix_chr(anno$chr)
anno$pos <- as.numeric(anno$pos)

anno <- anno[!is.na(anno$chr) & !is.na(anno$pos), ]

# Reorder beta matrix according to annotation
common_cpgs <- intersect(rownames(beta_119526), anno$CpG)

beta_119526 <- beta_119526[common_cpgs, , drop = FALSE]
anno <- anno[match(common_cpgs, anno$CpG), ]

identical(rownames(beta_119526), anno$CpG)

# CpG genomic ranges
cpg_gr <- GRanges(
  seqnames = anno$chr,
  ranges = IRanges(start = anno$pos, end = anno$pos),
  CpG = anno$CpG
)

############################################################
# 5. Convert discovery DMRs into GRanges
############################################################

dmr_gr <- GRanges(
  seqnames = dD$chr,
  ranges = IRanges(start = dD$start, end = dD$end),
  DMR_ID = dD$DMR_ID
)

############################################################
# 6. Find CpGs inside each discovery DMR
############################################################

hits <- findOverlaps(dmr_gr, cpg_gr)

cat("Number of DMR-CpG overlaps:", length(hits), "\n")

dmr_cpg_list <- split(
  mcols(cpg_gr)$CpG[subjectHits(hits)],
  queryHits(hits)
)

############################################################
# 7. Calculate DMR-level beta in GSE119526
############################################################

validation_results <- dD

validation_results$nCpG_GSE119526 <- 0
validation_results$meanBeta_normal_GSE119526 <- NA_real_
validation_results$meanBeta_CRC_GSE119526 <- NA_real_
validation_results$deltaBeta_GSE119526 <- NA_real_
validation_results$p.value_GSE119526 <- NA_real_

# DMR-level beta matrix: rows = DMRs, columns = samples
dmr_beta_119526 <- matrix(
  NA_real_,
  nrow = nrow(dD),
  ncol = ncol(beta_119526),
  dimnames = list(dD$DMR_ID, colnames(beta_119526))
)

normal_samples <- names(group_list)[group_list == "normal"]
crc_samples <- names(group_list)[group_list == "CRC"]

for (i in seq_len(nrow(dD))) {
  
  cpgs <- dmr_cpg_list[[as.character(i)]]
  
  if (is.null(cpgs)) {
    next
  }
  
  cpgs <- unique(intersect(cpgs, rownames(beta_119526)))
  
  validation_results$nCpG_GSE119526[i] <- length(cpgs)
  
  if (length(cpgs) < 1) {
    next
  }
  
  # Mean beta of all CpGs inside this DMR for each sample
  dmr_value <- colMeans(beta_119526[cpgs, , drop = FALSE], na.rm = TRUE)
  
  dmr_beta_119526[i, ] <- dmr_value
  
  normal_value <- dmr_value[normal_samples]
  crc_value <- dmr_value[crc_samples]
  
  validation_results$meanBeta_normal_GSE119526[i] <- mean(normal_value, na.rm = TRUE)
  validation_results$meanBeta_CRC_GSE119526[i] <- mean(crc_value, na.rm = TRUE)
  
  # Key validation value: CRC - normal
  validation_results$deltaBeta_GSE119526[i] <-
    mean(crc_value, na.rm = TRUE) - mean(normal_value, na.rm = TRUE)
  
  # One-sided Wilcoxon test: CRC beta > normal beta
  validation_results$p.value_GSE119526[i] <- tryCatch(
    wilcox.test(
      crc_value,
      normal_value,
      alternative = "greater",
      exact = FALSE
    )$p.value,
    error = function(e) NA_real_
  )
}

############################################################
# 8. Multiple testing correction and validation labels
############################################################

validation_results$FDR_GSE119526 <- p.adjust(
  validation_results$p.value_GSE119526,
  method = "BH"
)

# Same direction: CRC higher than normal
validation_results$same_trend <- 
  validation_results$deltaBeta_GSE119526 > 0

# Direction + moderate effect
validation_results$same_trend_delta_0.10 <- 
  validation_results$deltaBeta_GSE119526 > 0.10

# Direction + strong effect, comparable to your discovery threshold
validation_results$same_trend_delta_0.20 <- 
  validation_results$deltaBeta_GSE119526 > 0.20

# Direction + nominal significance
validation_results$validated_p005 <-
  validation_results$same_trend & 
  validation_results$p.value_GSE119526 < 0.05

# Direction + FDR significance
validation_results$validated_FDR005 <- 
  validation_results$same_trend & 
  validation_results$FDR_GSE119526 < 0.05

# Recommended strict validation:
validation_results$validated_strict <-
  validation_results$deltaBeta_GSE119526 > 0.20 &
  validation_results$p.value_GSE119526 < 0.05

############################################################
# 9. Summary
############################################################

summary_validation <- data.frame(
  Total_discovery_hyper_DMRs = nrow(validation_results),
  DMRs_with_CpG_overlap = sum(validation_results$nCpG_GSE119526 > 0, na.rm = TRUE),
  Same_trend_DMRs = sum(validation_results$same_trend, na.rm = TRUE),
  Same_trend_percent = round(
    100 * mean(validation_results$same_trend, na.rm = TRUE),
    2
  ),
  Same_trend_delta_0.10 = sum(validation_results$same_trend_delta_0.10, na.rm = TRUE),
  Same_trend_delta_0.20 = sum(validation_results$same_trend_delta_0.20, na.rm = TRUE),
  Validated_p005 = sum(validation_results$validated_p005, na.rm = TRUE),
  Validated_FDR005 = sum(validation_results$validated_FDR005, na.rm = TRUE),
  Validated_strict = sum(validation_results$validated_strict, na.rm = TRUE)
)

print(summary_validation)

############################################################
# 10. Save validation results
############################################################

write.csv(
  validation_results,
  file = "verificationPlots/GSE119526_validation_results_all_620_DMRs.csv",
  row.names = FALSE
)

write.csv(
  summary_validation,
  file = "GSE119526_validation_summary.csv",
  row.names = FALSE
)

validated_hyper_dmrs <- validation_results[
  validation_results$validated_strict == TRUE,
]

write.csv(
  validated_hyper_dmrs,
  file = "GSE119526_validated_hyper_DMRs_strict.csv",
  row.names = FALSE
)

save(
  dD,
  beta_119526,
  group_list,
  dmr_beta_119526,
  validation_results,
  summary_validation,
  validated_hyper_dmrs,
  file = "step4-GSE119526-validation-output.Rdata"
)

############################################################
# 11. Plot: discovery delta beta vs validation delta beta
############################################################

if ("deltaBeta_GSE199057" %in% colnames(validation_results)) {
  
  p_scatter <- ggplot(
    validation_results,
    aes(
      x = deltaBeta_GSE199057,
      y = deltaBeta_GSE119526
    )
  ) +
    geom_point(alpha = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    theme_bw() +
    labs(
      title = "Discovery vs validation delta beta",
      x = "GSE199057 delta beta: CRC - normal",
      y = "GSE119526 delta beta: CRC - normal"
    )
  
  ggsave(
    filename = "verificationPlots/ValidationPlots_in_GSE119526/GSE199057_vs_GSE119526_deltaBeta_scatter.png",
    plot = p_scatter,
    width = 6,
    height = 5,
    dpi = 300
  )
}

############################################################
# 12. Plot: heatmap of top validated DMRs
############################################################

top_dmr_ids <- validation_results$DMR_ID[
  validation_results$validated_strict == TRUE
]

top_dmr_ids <- top_dmr_ids[!is.na(top_dmr_ids)]

if (length(top_dmr_ids) >= 2) {
  
  top_dmr_ids <- top_dmr_ids[seq_len(min(50, length(top_dmr_ids)))]
  
  heat_mat <- dmr_beta_119526[top_dmr_ids, , drop = FALSE]
  
  annotation_col <- data.frame(
    Group = group_list,
    row.names = names(group_list)
  )
  
  pheatmap(
    heat_mat,
    scale = "row",
    show_colnames = FALSE,
    show_rownames = TRUE,
    annotation_col = annotation_col,
    cluster_cols = FALSE,
    filename = "verificationPlots/ValidationPlots_in_GSE119526/GSE119526_top_validated_DMRs_heatmap.png",
    width = 8,
    height = 8
  )
}

############################################################
# 12.Separate effect-size and significance criteria
############################################################

validation_count_df2 <- data.frame(
  Category = c(
    "Coverage / direction",
    "Coverage / direction",
    "Effect size",
    "Effect size",
    "Significance",
    "Significance",
    "Combined strict"
  ),
  
  Criterion = c(
    "CpG overlap",
    "Same direction\nΔβ > 0",
    "Δβ > 0.10",
    "Δβ > 0.20",
    "Direction + P < 0.05",
    "Direction + FDR < 0.05",
    "Δβ > 0.20 & P < 0.05"
  ),
  
  Count = c(
    sum(validation_results$nCpG_GSE119526 > 0, na.rm = TRUE),
    sum(validation_results$same_trend, na.rm = TRUE),
    sum(validation_results$same_trend_delta_0.10, na.rm = TRUE),
    sum(validation_results$same_trend_delta_0.20, na.rm = TRUE),
    sum(validation_results$validated_p005, na.rm = TRUE),
    sum(validation_results$validated_FDR005, na.rm = TRUE),
    sum(validation_results$validated_strict, na.rm = TRUE)
  ),
  
  stringsAsFactors = FALSE
)

validation_count_df2$Percent <- round(
  100 * validation_count_df2$Count / nrow(validation_results),
  2
)

validation_count_df2$Label <- paste0(
  validation_count_df2$Count,
  " (",
  validation_count_df2$Percent,
  "%)"
)

validation_count_df2$Criterion <- factor(
  validation_count_df2$Criterion,
  levels = validation_count_df2$Criterion
)

p_count_facet <- ggplot(
  validation_count_df2,
  aes(x = Criterion, y = Count)
) +
  geom_col(width = 0.7, fill = "grey70", color = "black") +
  geom_text(
    aes(label = Label),
    vjust = -0.25,
    size = 3.3
  ) +
  facet_wrap(~ Category, scales = "free_x") +
  ylim(0, max(validation_count_df2$Count) * 1.18) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Validation of GSE199057 hyper-DMRs in GSE119526",
    x = "Validation criterion",
    y = "Number of DMRs"
  )

ggsave(
  filename = "verificationPlots/ValidationPlots_in_GSE119526/GSE119526_validation_DMR_counts_by_category.png",
  plot = p_count_facet,
  width = 10,
  height = 5.5,
  dpi = 300
)
```

## low-background-filtering-using-GSE123478 (9 normal tissues)
```{r Low-background filtering in other normal tissues}
############################################################
# Step 5. Low-background filtering using GSE213478
#
# Goal:
#   Start from the 48 validated DMRs.
#   Remove DMRs with high background methylation in GSE213478.
############################################################

rm(list = ls())
options(stringsAsFactors = FALSE)

library(GEOquery)
library(Biobase)
library(minfi)
library(GenomicRanges)
library(IRanges)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(ggplot2)
library(pheatmap)

############################################################
# 1. Load the 48 validated DMRs from previous step
############################################################

load("step4-GSE119526-validation-output.Rdata")

# Use the 48 strict validated DMRs
dmr_48 <- validation_results[
  validation_results$validated_strict == TRUE,
]

dim(dmr_48)

# Check
table(dmr_48$validated_strict)
summary(dmr_48$deltaBeta_GSE119526)

############################################################
# 2. Prepare DMR coordinates
############################################################

if (!"chr" %in% colnames(dmr_48) && "seqnames" %in% colnames(dmr_48)) {
  dmr_48$chr <- dmr_48$seqnames
}

if (!all(c("chr", "start", "end", "DMR_ID") %in% colnames(dmr_48))) {
  stop("dmr_48 must contain chr/start/end/DMR_ID columns.")
}

fix_chr <- function(x) {
  x <- as.character(x)
  x <- gsub("^chr", "", x, ignore.case = TRUE)
  x[x == "23"] <- "X"
  x[x == "24"] <- "Y"
  x[x %in% c("M", "MT")] <- "M"
  paste0("chr", x)
}

dmr_48$chr <- fix_chr(dmr_48$chr)
dmr_48$start <- as.numeric(dmr_48$start)
dmr_48$end <- as.numeric(dmr_48$end)

dmr_48 <- dmr_48[
  !is.na(dmr_48$chr) &
    !is.na(dmr_48$start) &
    !is.na(dmr_48$end),
]

############################################################
# 3. Load GSE213478 beta matrix
############################################################

# If you already have a processed GSE213478 beta matrix,
# you can skip this getGEO part and directly assign:
# beta_213478 <- your_GSE213478_beta_matrix
#eset2 <- getGEO("GSE213478", destdir = "./", AnnotGPL = TRUE, getGPL = FALSE)
#pdata_all <- pData(eset2[[1]])

#beta_213478 <- fread("GSE213478_methylation_DNAm_noob_final_BMIQ_all_tissues_987.txt.gz",data.table = F)
#rownames(beta_213478) <- beta_213478[, 1]
#beta_213478 <- beta_213478[, -1]
#save(pdata_all,beta_213478,file = "step5.1-GSE213478-normal_tissuesoutput.Rdata")

load("step5.0-GSE213478-normal_tissuesoutput.Rdata")

dim(beta_213478)
#range(beta_213478, na.rm = TRUE)

############################################################
# 4. Remove probes with too many missing values
############################################################

keep_probe <- rowMeans(!is.na(beta_213478)) >= 0.95
beta_213478 <- beta_213478[keep_probe, , drop = FALSE]

dim(beta_213478)

############################################################
# 5. Load EPIC annotation and map CpGs to genomic coordinates
############################################################

anno <- minfi::getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
anno <- as.data.frame(anno)
anno$CpG <- rownames(anno)

anno <- anno[anno$CpG %in% rownames(beta_213478), ]

anno$chr <- fix_chr(anno$chr)
anno$pos <- as.numeric(anno$pos)

anno <- anno[
  !is.na(anno$chr) &
    !is.na(anno$pos),
]

# Align beta matrix and annotation
common_cpgs <- intersect(rownames(beta_213478), anno$CpG)

beta_213478 <- beta_213478[common_cpgs, , drop = FALSE]
anno <- anno[match(common_cpgs, anno$CpG), ]

identical(rownames(beta_213478), anno$CpG)

cpg_gr <- GRanges(
  seqnames = anno$chr,
  ranges = IRanges(start = anno$pos, end = anno$pos),
  CpG = anno$CpG
)

############################################################
# 6. Convert 48 DMRs into GRanges
############################################################

dmr_gr <- GRanges(
  seqnames = dmr_48$chr,
  ranges = IRanges(start = dmr_48$start, end = dmr_48$end),
  DMR_ID = dmr_48$DMR_ID
)

############################################################
# 7. Find CpGs inside each DMR in GSE213478
############################################################

hits <- findOverlaps(dmr_gr, cpg_gr)

cat("Number of DMR-CpG overlaps in GSE213478:", length(hits), "\n")

dmr_cpg_list <- split(
  mcols(cpg_gr)$CpG[subjectHits(hits)],
  queryHits(hits)
)

############################################################
# 8. Calculate DMR-level beta in GSE213478
############################################################

dmr_beta_213478 <- matrix(
  NA_real_,
  nrow = nrow(dmr_48),
  ncol = ncol(beta_213478),
  dimnames = list(dmr_48$DMR_ID, colnames(beta_213478))
)

background_results <- dmr_48

background_results$nCpG_GSE213478 <- 0
background_results$meanBeta_GSE213478 <- NA_real_
background_results$medianBeta_GSE213478 <- NA_real_
background_results$maxBeta_GSE213478 <- NA_real_
background_results$p95Beta_GSE213478 <- NA_real_
background_results$propSamples_beta_gt_0.20_GSE213478 <- NA_real_

for (i in seq_len(nrow(dmr_48))) {
  
  cpgs <- dmr_cpg_list[[as.character(i)]]
  
  if (is.null(cpgs)) {
    next
  }
  
  cpgs <- unique(intersect(cpgs, rownames(beta_213478)))
  
  background_results$nCpG_GSE213478[i] <- length(cpgs)
  
  if (length(cpgs) < 1) {
    next
  }
  
  # DMR-level beta per sample
  dmr_value <- colMeans(beta_213478[cpgs, , drop = FALSE], na.rm = TRUE)
  
  dmr_beta_213478[i, ] <- dmr_value
  
  background_results$meanBeta_GSE213478[i] <- mean(dmr_value, na.rm = TRUE)
  background_results$medianBeta_GSE213478[i] <- median(dmr_value, na.rm = TRUE)
  background_results$maxBeta_GSE213478[i] <- max(dmr_value, na.rm = TRUE)
  background_results$p95Beta_GSE213478[i] <- as.numeric(
    quantile(dmr_value, probs = 0.95, na.rm = TRUE)
  )
  background_results$propSamples_beta_gt_0.20_GSE213478[i] <- mean(
    dmr_value > 0.20,
    na.rm = TRUE
  )
}

############################################################
# 9. Define high-background exclusion criteria
############################################################

# Criterion A: strictest
# Exclude if any GSE213478 sample has DMR beta > 0.20
background_results$exclude_by_max_beta_gt_0.20 <-
  background_results$maxBeta_GSE213478 > 0.20

# Criterion B: recommended for large multi-tissue background
# Exclude if 95th percentile beta > 0.20
background_results$exclude_by_p95_beta_gt_0.20 <-
  background_results$p95Beta_GSE213478 > 0.20

# Criterion C: more relaxed
# Exclude only if mean background beta > 0.20
background_results$exclude_by_mean_beta_gt_0.20 <-
  background_results$meanBeta_GSE213478 > 0.20

############################################################
# 10. Choose final low-background rule
############################################################

# Recommended default:
# Use p95 <= 0.20 as low-background criterion
# This avoids one extreme outlier sample removing a DMR.

low_background_dmrs <- background_results[
  background_results$nCpG_GSE213478 > 0 &
    background_results$exclude_by_p95_beta_gt_0.20 == FALSE,
]

removed_high_background_dmrs <- background_results[
  background_results$exclude_by_p95_beta_gt_0.20 == TRUE,
]

############################################################
# 11. Summary
############################################################

background_summary <- data.frame(
  Input_validated_DMRs = nrow(background_results),
  DMRs_with_CpG_overlap_GSE213478 = sum(background_results$nCpG_GSE213478 > 0, na.rm = TRUE),
  
  Excluded_by_max_beta_gt_0.20 = sum(background_results$exclude_by_max_beta_gt_0.20, na.rm = TRUE),
  Retained_by_max_beta_le_0.20 = sum(!background_results$exclude_by_max_beta_gt_0.20, na.rm = TRUE),
  
  Excluded_by_p95_beta_gt_0.20 = sum(background_results$exclude_by_p95_beta_gt_0.20, na.rm = TRUE),
  Retained_by_p95_beta_le_0.20 = nrow(low_background_dmrs),
  
  Excluded_by_mean_beta_gt_0.20 = sum(background_results$exclude_by_mean_beta_gt_0.20, na.rm = TRUE),
  Retained_by_mean_beta_le_0.20 = sum(!background_results$exclude_by_mean_beta_gt_0.20, na.rm = TRUE)
)

print(background_summary)

############################################################
# Extract DMRs retained by mean beta <= 0.20
############################################################

low_background_dmrs_mean <- background_results[
  background_results$nCpG_GSE213478 > 0 &
    background_results$meanBeta_GSE213478 <= 0.20,
]

removed_high_background_dmrs_mean <- background_results[
  background_results$nCpG_GSE213478 > 0 &
    background_results$meanBeta_GSE213478 > 0.20,
]

dim(low_background_dmrs_mean)
dim(removed_high_background_dmrs_mean)

write.csv(
  low_background_dmrs_mean,
  file = "verificationPlots/GSE213478_retained_low_background_DMRs_meanBeta_le_0.20.csv",
  row.names = FALSE
)

write.csv(
  removed_high_background_dmrs_mean,
  file = "verificationPlots/GSE213478_removed_high_background_DMRs_meanBeta_gt_0.20.csv",
  row.names = FALSE
)

save(
  low_background_dmrs_mean,
  removed_high_background_dmrs_mean,
  file = "step5-GSE213478-low-background-filter-meanBeta-output.Rdata"
)
############################################################
# Plot: retained vs removed DMRs after mean beta filtering
############################################################
n_input <- nrow(background_results)
n_retained <- nrow(low_background_dmrs_mean)
n_removed <- nrow(removed_high_background_dmrs_mean)
plot_df <- data.frame(
  Status = c(
    "Retained\nmean beta ≤ 0.20",
    "Removed\nmean beta > 0.20"
  ),
  Count = c(
    n_retained,
    n_removed
  ),
  stringsAsFactors = FALSE
)

plot_df$Percent <- round(
  100 * plot_df$Count / n_input,
  2
)

plot_df$Label <- paste0(
  plot_df$Count,
  " DMRs\n(",
  plot_df$Percent,
  "%)"
)

p_retained_removed <- ggplot(
  plot_df,
  aes(x = "GSE213478\nmean beta filter", y = Count, fill = Status)
) +
  geom_col(
    width = 0.45,
    color = "black"
  ) +
  geom_text(
    aes(label = Label),
    position = position_stack(vjust = 0.5),
    size = 4
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 13, hjust = 0.5),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Retained and removed DMRs after GSE213478 filtering",
    x = "",
    y = "Number of DMRs",
    fill = "Filtering result"
  )

print(p_retained_removed)

ggsave(
  filename = "verificationPlots/background_filtering_in_GSE123478/GSE213478_48_to_42_retained_removed_stacked_bar.png",
  plot = p_retained_removed,
  width = 6,
  height = 5,
  dpi = 300
)

```

## low-background-filtering-using-GSE196696(whole blood cells)
```{r low background filtering in whole blood cells}
############################################################
# Step 6. Continue low-background filtering in GSE196696
#
# Input:
#   42 DMRs retained after GSE213478 mean beta <= 0.20
#
# Goal:
#   Further filter candidate DMRs using GSE196696
#   Criterion: mean beta in GSE196696 <= 0.20
############################################################

rm(list = ls())
options(stringsAsFactors = FALSE)

library(GEOquery)
library(Biobase)
library(minfi)
library(GenomicRanges)
library(IRanges)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(ggplot2)
library(data.table)

############################################################
# 1. Load previous GSE213478 mean-beta filtering result
############################################################

load("step5-GSE213478-low-background-filter-meanBeta-output.Rdata")

dim(low_background_dmrs_mean)
dim(removed_high_background_dmrs_mean)

# Use the 42 retained DMRs as input for GSE196696 filtering
candidate_dmrs <- low_background_dmrs_mean

cat("Input DMRs for GSE196696 filtering:", nrow(candidate_dmrs), "\n")

############################################################
# 2. Prepare DMR coordinates
############################################################

candidate_dmrs <- as.data.frame(candidate_dmrs)

# If chr column is absent but seqnames exists, create chr
if (!"chr" %in% colnames(candidate_dmrs) && "seqnames" %in% colnames(candidate_dmrs)) {
  candidate_dmrs$chr <- candidate_dmrs$seqnames
}

if (!all(c("chr", "start", "end", "DMR_ID") %in% colnames(candidate_dmrs))) {
  stop("candidate_dmrs must contain chr/start/end/DMR_ID columns.")
}

fix_chr <- function(x) {
  x <- as.character(x)
  x <- gsub("^chr", "", x, ignore.case = TRUE)
  x[x == "23"] <- "X"
  x[x == "24"] <- "Y"
  x[x %in% c("M", "MT")] <- "M"
  paste0("chr", x)
}

candidate_dmrs$chr <- fix_chr(candidate_dmrs$chr)
candidate_dmrs$start <- as.numeric(candidate_dmrs$start)
candidate_dmrs$end <- as.numeric(candidate_dmrs$end)

candidate_dmrs <- candidate_dmrs[
  !is.na(candidate_dmrs$chr) &
    !is.na(candidate_dmrs$start) &
    !is.na(candidate_dmrs$end),
]

dim(candidate_dmrs)

############################################################
# 3. Load GSE196696 beta matrix
############################################################

if(F){eset196 <- getGEO(
  "GSE196696",
  destdir = "./",
  GSEMatrix = TRUE,
  AnnotGPL = FALSE,
  getGPL = FALSE
)

eset196 <- eset196[[1]]
pd_196696 <- pData(eset196)

file_196696 <- "GSE196696_processed_data.tsv.gz"

# Read only first 5 rows to check format
#test_196696 <- fread(file_196696,nrows = 5,data.table = FALSE,check.names = FALSE)

#dim(test_196696)
#colnames(test_196696)[1:10]
#test_196696[1:5, 1:5]

#Read GSE196696 processed data and extract beta values

# read processed data
raw_196696 <- fread(file_196696,data.table = FALSE,check.names = FALSE)
save(raw_196696,file = 'raw_196696.Rdata')

load("raw_196696.Rdata")

raw_df <- as.data.frame(raw_196696)
pd_df  <- as.data.frame(pd_196696)

# Extract array_id from title
pd_df$array_id <- sub(".* -\\s*", "", pd_df$title)

# Check extraction result
head(pd_df[, c("geo_accession", "title", "array_id")])

#  Extract beta value columns
beta_cols <- colnames(raw_df)[
  colnames(raw_df) != "ID_REF" &
    !grepl("^Detection Pval", colnames(raw_df))
]

length(beta_cols)
head(beta_cols)

# Build mapping from array_id to GSM
array_to_gsm <- setNames(pd_df$geo_accession, pd_df$array_id)

# Check whether any raw data column cannot be mapped to GSM
unmatched <- setdiff(beta_cols, names(array_to_gsm))

if (length(unmatched) > 0) {
  print(head(unmatched))
  stop("Some beta columns cannot be matched to GSM using array_id extracted from pd_196696$title. Please check title format or column name format.")
}

# Get GSM names in the order of beta_cols
gsm_names <- unname(array_to_gsm[beta_cols])

#  Check mapping correctness
head(data.frame(raw_column = beta_cols, GSM = gsm_names))

# Extract beta dataframe
beta_df <- raw_df[, c("ID_REF", beta_cols), drop = FALSE]

# Check for duplicate ID_REF
if (anyDuplicated(beta_df$ID_REF) > 0) {
  stop("Duplicate probes found in ID_REF, please handle duplicates first.")
}

#  Build standard beta value matrix
beta_196696 <- beta_df[, beta_cols, drop = FALSE]
rownames(beta_196696) <- beta_df$ID_REF
colnames(beta_196696) <- gsm_names

# Convert to numeric matrix
beta_196696 <- as.matrix(beta_196696)
storage.mode(beta_196696) <- "numeric"

# Basic checks
dim(beta_196696)
range(beta_196696, na.rm = TRUE)
head(rownames(beta_196696))
head(colnames(beta_196696))

# Further check if it conforms to beta matrix standards
sum(is.na(colnames(beta_196696)))
sum(is.na(rownames(beta_196696)))
anyDuplicated(colnames(beta_196696))
anyDuplicated(rownames(beta_196696))

save(beta_196696,pd_196696,file = 'beta_pd_196696_output.Rdata')
}

load("beta_pd_196696_output.Rdata")

############################################################
# 4. Basic CpG filtering
############################################################

# Keep CpG probes only
beta_196696 <- beta_196696[
  grepl("^cg", rownames(beta_196696)),
  ,
  drop = FALSE
]

# Remove probes with too many missing values
keep_probe <- rowMeans(!is.na(beta_196696)) >= 0.95
beta_196696 <- beta_196696[keep_probe, , drop = FALSE]

dim(beta_196696)

############################################################
# 5. Load EPIC annotation and map CpGs to genomic coordinates
############################################################

anno <- minfi::getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
anno <- as.data.frame(anno)
anno$CpG <- rownames(anno)

anno <- anno[anno$CpG %in% rownames(beta_196696), ]

anno$chr <- fix_chr(anno$chr)
anno$pos <- as.numeric(anno$pos)

anno <- anno[
  !is.na(anno$chr) &
    !is.na(anno$pos),
]

common_cpgs <- intersect(rownames(beta_196696), anno$CpG)

beta_196696 <- beta_196696[common_cpgs, , drop = FALSE]
anno <- anno[match(common_cpgs, anno$CpG), ]

stopifnot(identical(rownames(beta_196696), anno$CpG))

cpg_gr <- GRanges(
  seqnames = anno$chr,
  ranges = IRanges(start = anno$pos, end = anno$pos),
  CpG = anno$CpG
)

############################################################
# 6. Convert candidate DMRs to GRanges
############################################################

dmr_gr <- GRanges(
  seqnames = candidate_dmrs$chr,
  ranges = IRanges(
    start = candidate_dmrs$start,
    end = candidate_dmrs$end
  ),
  DMR_ID = candidate_dmrs$DMR_ID
)

############################################################
# 7. Find CpGs inside each DMR in GSE196696
############################################################

hits <- findOverlaps(dmr_gr, cpg_gr)

cat("Number of DMR-CpG overlaps in GSE196696:", length(hits), "\n")

dmr_cpg_list <- split(
  mcols(cpg_gr)$CpG[subjectHits(hits)],
  queryHits(hits)
)

############################################################
# 8. Calculate DMR-level background beta in GSE196696
############################################################

dmr_beta_196696 <- matrix(
  NA_real_,
  nrow = nrow(candidate_dmrs),
  ncol = ncol(beta_196696),
  dimnames = list(candidate_dmrs$DMR_ID, colnames(beta_196696))
)

background_results_196696 <- candidate_dmrs

background_results_196696$nCpG_GSE196696 <- 0
background_results_196696$meanBeta_GSE196696 <- NA_real_
background_results_196696$medianBeta_GSE196696 <- NA_real_
background_results_196696$maxBeta_GSE196696 <- NA_real_
background_results_196696$p95Beta_GSE196696 <- NA_real_
background_results_196696$propSamples_beta_gt_0.20_GSE196696 <- NA_real_

for (i in seq_len(nrow(candidate_dmrs))) {
  
  cpgs <- dmr_cpg_list[[as.character(i)]]
  
  if (is.null(cpgs)) {
    next
  }
  
  cpgs <- unique(intersect(cpgs, rownames(beta_196696)))
  
  background_results_196696$nCpG_GSE196696[i] <- length(cpgs)
  
  if (length(cpgs) < 1) {
    next
  }
  
  # DMR-level beta per sample
  dmr_value <- colMeans(beta_196696[cpgs, , drop = FALSE], na.rm = TRUE)
  
  dmr_beta_196696[i, ] <- dmr_value
  
  background_results_196696$meanBeta_GSE196696[i] <- mean(dmr_value, na.rm = TRUE)
  background_results_196696$medianBeta_GSE196696[i] <- median(dmr_value, na.rm = TRUE)
  background_results_196696$maxBeta_GSE196696[i] <- max(dmr_value, na.rm = TRUE)
  background_results_196696$p95Beta_GSE196696[i] <- as.numeric(
    quantile(dmr_value, probs = 0.95, na.rm = TRUE)
  )
  background_results_196696$propSamples_beta_gt_0.20_GSE196696[i] <- mean(
    dmr_value > 0.20,
    na.rm = TRUE
  )
}

############################################################
# 9. Filter by mean beta <= 0.20 in GSE196696
############################################################

low_background_dmrs_mean_196696 <- background_results_196696[
  background_results_196696$nCpG_GSE196696 > 0 &
    background_results_196696$meanBeta_GSE196696 <= 0.20,
]

removed_high_background_dmrs_mean_196696 <- background_results_196696[
  background_results_196696$nCpG_GSE196696 > 0 &
    background_results_196696$meanBeta_GSE196696 > 0.20,
]

dim(low_background_dmrs_mean_196696)
dim(removed_high_background_dmrs_mean_196696)

############################################################
# 10. Summary
############################################################

background_summary_196696 <- data.frame(
  Input_DMRs_after_GSE213478_mean_filter = nrow(background_results_196696),
  DMRs_with_CpG_overlap_GSE196696 = sum(background_results_196696$nCpG_GSE196696 > 0, na.rm = TRUE),
  Excluded_by_mean_beta_gt_0.20_GSE196696 = nrow(removed_high_background_dmrs_mean_196696),
  Retained_by_mean_beta_le_0.20_GSE196696 = nrow(low_background_dmrs_mean_196696),
  Retained_percent = round(
    100 * nrow(low_background_dmrs_mean_196696) / nrow(background_results_196696),
    2
  )
)

print(background_summary_196696)

############################################################
# 11. Save outputs
############################################################

write.csv(
  background_results_196696,
  file = "verificationPlots/GSE196696_background_filter_all_42_DMRs.csv",
  row.names = FALSE
)

write.csv(
  low_background_dmrs_mean_196696,
  file = "verificationPlots/GSE196696_retained_low_background_DMRs_meanBeta_le_0.20.csv",
  row.names = FALSE
)

write.csv(
  removed_high_background_dmrs_mean_196696,
  file = "verificationPlots/GSE196696_removed_high_background_DMRs_meanBeta_gt_0.20.csv",
  row.names = FALSE
)

write.csv(
  background_summary_196696,
  file = "verificationPlots/GSE196696_background_filter_summary_meanBeta.csv",
  row.names = FALSE
)

save(
  candidate_dmrs,
  beta_196696,
  pd_196696,
  dmr_beta_196696,
  background_results_196696,
  low_background_dmrs_mean_196696,
  removed_high_background_dmrs_mean_196696,
  background_summary_196696,
  file = "step6-GSE196696-low-background-filter-meanBeta-output.Rdata"
)
#################################################################################
## Visualisation
#################################################################################

load("step6-GSE196696-low-background-filter-meanBeta-output.Rdata")
##Fig1 GSE196696_filtering_flow

library(ggplot2)

flow_df <- data.frame(
  Step = factor(
    c(
      "Input after GSE213478",
      "With CpG overlap in GSE196696",
      "Retained: mean beta <= 0.20",
      "Removed: mean beta > 0.20"
    ),
    levels = c(
      "Input after GSE213478",
      "With CpG overlap in GSE196696",
      "Retained: mean beta <= 0.20",
      "Removed: mean beta > 0.20"
    )
  ),
  Count = c(
    nrow(background_results_196696),
    sum(background_results_196696$nCpG_GSE196696 > 0, na.rm = TRUE),
    nrow(low_background_dmrs_mean_196696),
    nrow(removed_high_background_dmrs_mean_196696)
  )
)

p_flow <- ggplot(flow_df, aes(x = Step, y = Count)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = Count), vjust = -0.4, size = 4) +
  theme_bw(base_size = 13) +
  labs(
    title = "GSE196696 whole-blood background filtering",
    x = NULL,
    y = "Number of DMRs"
  ) +
  ylim(0, max(flow_df$Count) * 1.15) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

p_flow

ggsave(
  "verificationPlots/background_filtering_in_GSE196696/GSE196696_filtering_flow.png",
  p_flow,
  width = 7,
  height = 4.8,
  dpi = 300
)

##  Figure2 Are the 42 candidate DMRs low-methylated in GSE196696 whole blood?
plot_df <- background_results_196696[
  background_results_196696$nCpG_GSE196696 > 0,
]

p_mean_box <- ggplot(plot_df, aes(x = "GSE196696 whole blood", y = meanBeta_GSE196696)) +
  geom_boxplot(width = 0.35, outlier.shape = NA) +
  geom_jitter(width = 0.08, size = 2, alpha = 0.75) +
  geom_hline(
    yintercept = 0.20,
    linetype = "dashed",
    linewidth = 0.8
  ) +
  theme_bw(base_size = 13) +
  labs(
    title = "Low-background methylation of retained DMRs in GSE196696",
    x = NULL,
    y = "DMR-level mean beta"
  ) +
  ylim(0, 0.25) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

p_mean_box

ggsave(
  "verificationPlots/background_filtering_in_GSE196696/GSE196696_meanBeta_boxplot.png",
  p_mean_box,
  width = 4.8,
  height = 4.8,
  dpi = 300
)


##Figure3 DMR × sample heatmap

library(pheatmap)

#  Keep only DMRs with CpG overlap
heat_mat <- dmr_beta_196696[
  rownames(dmr_beta_196696) %in% plot_df$DMR_ID,
  ,
  drop = FALSE
]

# Sort DMRs by mean beta (high to low)
heat_mat <- heat_mat[
  order(rowMeans(heat_mat, na.rm = TRUE), decreasing = TRUE),
  ,
  drop = FALSE
]

# For clarity, do not show 570 sample names
pdf(
  "verificationPlots/background_filtering_in_GSE196696/GSE196696_DMR_sample_heatmap.pdf",
  width = 10,
  height = 7
)

pheatmap(
  heat_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_colnames = FALSE,
  show_rownames = TRUE,
  fontsize_row = 6,
  breaks = seq(0, 0.25, length.out = 101),
  main = "DMR-level beta values in GSE196696 whole blood"
)

dev.off()
```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
