---
title: "Comparison of 116-DMR and Top10-DMR radial SVM models"
author: "YC"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This report examines whether the final top 10 DMR panel retains performance comparable to the full 116-DMR panel.

Both radial SVM models are trained using the same GSE101764 samples, the same repeated cross-validation folds, and the same preprocessing and tuning strategy. They are then evaluated on the same TCGA full cohort.

**Interpretation note:** the top 10 DMRs were selected using TCGA full-cohort SHAP prioritisation. Therefore, this comparison is a descriptive reduced-panel analysis rather than a fully independent external validation of the top 10 panel.

# Initialise analysis

This chunk starts from a clean environment and fixes the random seed.

```{r initialise-analysis}
rm(list = ls())

options(stringsAsFactors = FALSE)
options(scipen = 20)

set.seed(123)
```

The same seed is used throughout the model comparison.

# Load packages

The required packages are loaded below.

```{r load-packages}
library(data.table)
library(caret)
library(pROC)
library(ggplot2)
library(kernlab)
```

`caret::train()` is called explicitly later to avoid conflicts with other functions named `train`.

# Define files

This chunk defines the input files and output directory.

```{r define-inputs}
training_model_file <-
  "GSE101764_ML_training_3models_116DMRs.Rdata"

tcga_full_file <-
  "TCGA_full_testing_results_3models.Rdata"

top10_file <-
  "Final_top10_DMRs_for_wetlab_validation__nsim500.csv"

outdir <-
  "step6.1_full_TCGA_top10_vs_116_performance"

input_files <- c(
  training_model_file,
  tcga_full_file,
  top10_file
)

if (!all(file.exists(input_files))) {
  stop(
    "Missing input file(s): ",
    paste(input_files[!file.exists(input_files)], collapse = ", ")
  )
}

if (!dir.exists(outdir)) {
  dir.create(outdir, recursive = TRUE)
}
```

The analysis stops if any required input file is unavailable.

# Load data and model objects

Only the objects required for this comparison are checked.

```{r load-analysis-objects}
load(training_model_file)

if (!exists("model_svm")) {
  stop("model_svm was not found in the training RData file.")
}

if (!exists("ml_df")) {
  stop("ml_df was not found in the training RData file.")
}

load(tcga_full_file)

if (!exists("test_df_full")) {
  stop("test_df_full was not found in the TCGA RData file.")
}

top10_dmrs <- fread(top10_file)

ml_df <- as.data.frame(ml_df)
test_df_full <- as.data.frame(test_df_full)
```

The original SVM supplies the full 116-feature names, while `ml_df` and `test_df_full` provide training and testing data.

# Prepare outcomes

Tumor is placed first because it is the positive class used by `twoClassSummary()`.

```{r prepare-outcomes}
ml_df$group <- factor(
  as.character(ml_df$group),
  levels = c("Tumor", "Normal")
)

test_df_full$group <- factor(
  as.character(test_df_full$group),
  levels = c("Tumor", "Normal")
)

if (anyNA(ml_df$group) || anyNA(test_df_full$group)) {
  stop("Unexpected or missing group labels were found.")
}

table(ml_df$group)
table(test_df_full$group)
```

Both cohorts must contain valid Tumor and Normal labels.

# Define feature panels

The full feature set comes from the original radial SVM and the reduced panel comes from the corrected Step 5 result.

```{r define-feature-panels}
features_116 <- setdiff(
  colnames(model_svm$trainingData),
  ".outcome"
)

if (!"Feature" %in% names(top10_dmrs)) {
  stop("The top 10 file does not contain a Feature column.")
}

features_top10 <- as.character(top10_dmrs$Feature)

if (length(features_116) != 116) {
  stop("The original radial SVM does not contain exactly 116 DMRs.")
}

if (length(features_top10) != 10) {
  stop("The selected panel does not contain exactly 10 DMRs.")
}

if (anyDuplicated(features_top10) > 0) {
  stop("Duplicated DMRs were found in the top 10 panel.")
}

length(features_116)
features_top10
```

The top 10 panel must contain 10 unique DMR names.

# Check feature availability

This essential check confirms that all required predictors occur in both cohorts.

```{r check-feature-availability}
missing_features <- unique(c(
  setdiff(features_116, colnames(ml_df)),
  setdiff(features_116, colnames(test_df_full)),
  setdiff(features_top10, colnames(ml_df)),
  setdiff(features_top10, colnames(test_df_full))
))

if (length(missing_features) > 0) {
  stop(
    "Missing feature(s): ",
    paste(missing_features, collapse = ", ")
  )
}

if (!all(features_top10 %in% features_116)) {
  stop("Some top 10 DMRs are not included in the 116-DMR panel.")
}
```

The reduced model must be a true subset of the full model.

# Prepare model data

The two models use the same samples and outcome but different predictor panels.

```{r prepare-model-data}
train_116 <- ml_df[
  ,
  c("group", features_116),
  drop = FALSE
]

train_top10 <- ml_df[
  ,
  c("group", features_top10),
  drop = FALSE
]

test_116 <- test_df_full[
  ,
  features_116,
  drop = FALSE
]

test_top10 <- test_df_full[
  ,
  features_top10,
  drop = FALSE
]

if (anyNA(train_116) || anyNA(train_top10) ||
    anyNA(test_116) || anyNA(test_top10)) {
  stop("Missing predictor values were found.")
}
```

Only the predictor count differs between the two training datasets.

# Define identical cross-validation folds

Both models are tuned using the same repeated five-fold cross-validation splits.

```{r define-cross-validation}
set.seed(123)

cv_index <- caret::createMultiFolds(
  y = train_116$group,
  k = 5,
  times = 10
)

ctrl_svm <- caret::trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 10,
  index = cv_index,
  classProbs = TRUE,
  summaryFunction = caret::twoClassSummary,
  savePredictions = "final"
)
```

Using identical folds makes the tuning comparison fairer.

# Train the 116-DMR model

The full radial SVM is retrained using the same procedure as the reduced model.

```{r train-116-model}
set.seed(123)

model_svm_116_retrained <- caret::train(
  group ~ .,
  data = train_116,
  method = "svmRadial",
  metric = "ROC",
  trControl = ctrl_svm,
  tuneLength = 10,
  preProcess = c("center", "scale")
)

print(model_svm_116_retrained$bestTune)
```

The best `sigma` and `C` are selected by repeated cross-validated ROC.

# Train the Top10-DMR model

The reduced radial SVM is trained with the same samples, folds, and tuning strategy.

```{r train-top10-model}
set.seed(123)

model_svm_top10 <- caret::train(
  group ~ .,
  data = train_top10,
  method = "svmRadial",
  metric = "ROC",
  trControl = ctrl_svm,
  tuneLength = 10,
  preProcess = c("center", "scale")
)

print(model_svm_top10$bestTune)
```
# Compare cross-validated training performance

This chunk extracts the best repeated cross-validation performance from the two fitted SVM models.

```{r compare-cross-validated-performance}
cv_116 <- caret::getTrainPerf(
  model_svm_116_retrained
)

cv_top10 <- caret::getTrainPerf(
  model_svm_top10
)

cv_performance_comparison <- data.table(
  Model = c(
    "116-DMR radial SVM",
    "Top10-DMR radial SVM"
  ),
  CV_ROC = c(
    cv_116$TrainROC,
    cv_top10$TrainROC
  ),
  CV_Sensitivity = c(
    cv_116$TrainSens,
    cv_top10$TrainSens
  ),
  CV_Specificity = c(
    cv_116$TrainSpec,
    cv_top10$TrainSpec
  )
)

print(cv_performance_comparison)
```

Similar cross-validated ROC, sensitivity, and specificity values suggest that reducing the panel does not substantially weaken performance in the training cohort.

# Define performance calculation

This helper function avoids division-by-zero errors when calculating classification metrics.

```{r define-safe-division}
safe_div <- function(x, y) {
  ifelse(
    y == 0,
    NA_real_,
    x / y
  )
}
```

A metric is returned as `NA` only when its denominator is zero.

This function calculates the confusion-matrix metrics and AUROC for one fitted model.

```{r define-performance-function}
calc_performance <- function(
    true_group, pred_class, prob_tumor, model_name) {

  true_group <- factor(
    as.character(true_group), levels = c("Tumor", "Normal"))
  pred_class <- factor(
    as.character(pred_class), levels = c("Tumor", "Normal"))
  cm <- table(
    Predicted = pred_class, Reference = true_group)

  TP <- cm["Tumor", "Tumor"]
  TN <- cm["Normal", "Normal"]
  FP <- cm["Tumor", "Normal"]
  FN <- cm["Normal", "Tumor"]

  sensitivity <- safe_div(TP, TP + FN)
  specificity <- safe_div(TN, TN + FP)
  ppv <- safe_div(TP, TP + FP)
  npv <- safe_div(TN, TN + FN)
  accuracy <- safe_div(TP + TN, sum(cm))
  f1 <- safe_div(2 * ppv * sensitivity, ppv + sensitivity)

  roc_obj <- pROC::roc(
    response = true_group,
    predictor = prob_tumor,
    levels = c("Normal", "Tumor"),
    direction = "<",
    quiet = TRUE)

  performance <- data.table(
    Cohort = "TCGA full",
    Model = model_name,
    n_Normal = sum(true_group == "Normal"),
    n_Tumor = sum(true_group == "Tumor"),
    Accuracy = as.numeric(accuracy),
    AUROC = as.numeric(pROC::auc(roc_obj)),
    Sensitivity = as.numeric(sensitivity),
    Specificity = as.numeric(specificity),
    PPV = as.numeric(ppv),
    NPV = as.numeric(npv),
    F1 = as.numeric(f1),
    Balanced_Accuracy = as.numeric(
      (sensitivity + specificity) / 2))

  list(
    performance = performance,
    roc = roc_obj,
    confusion_matrix = cm)
}
```

The returned ROC and confusion matrix are retained for paired model comparisons.

# Define model prediction

This function applies one fitted model to TCGA and returns summary metrics and sample-level predictions.

```{r define-prediction-function}
predict_and_score <- function(
    model, feature_set, test_df, model_name) {

  newdata <- test_df[
    ,
    feature_set,
    drop = FALSE
  ]

  pred_class <- predict(
    model,
    newdata = newdata
  )

  prob_tumor <- predict(
    model,
    newdata = newdata,
    type = "prob"
  )[, "Tumor"]

  score <- calc_performance(
    true_group = test_df$group,
    pred_class = pred_class,
    prob_tumor = prob_tumor,
    model_name = model_name
  )

  sample_ids <- rownames(test_df)
  if (is.null(sample_ids)) {
    sample_ids <- paste0(
      "Sample_",
      seq_len(nrow(test_df))
    )
  }

  predictions <- data.table(
    Sample = sample_ids,
    True_group = as.character(test_df$group),
    Predicted_group = as.character(pred_class),
    Tumor_probability = as.numeric(prob_tumor)
  )

  list(
    performance = score$performance,
    predictions = predictions,
    roc = score$roc,
    confusion_matrix = score$confusion_matrix
  )
}
```

Both models will be evaluated using the same function and the same TCGA samples.

# Evaluate the two models

The retrained 116-DMR and Top10-DMR models are now applied to the TCGA full cohort.

```{r evaluate-models}
res_116_full <- predict_and_score(
  model = model_svm_116_retrained,
  feature_set = features_116,
  test_df = test_df_full,
  model_name = "116-DMR radial SVM"
)

res_top10_full <- predict_and_score(
  model = model_svm_top10,
  feature_set = features_top10,
  test_df = test_df_full,
  model_name = "Top10-DMR radial SVM"
)

performance_full_comparison <- rbindlist(
  list(
    res_116_full$performance,
    res_top10_full$performance
  )
)

print(performance_full_comparison)
```

The resulting table directly compares performance in the same TCGA cohort.

# Compare metric differences

This chunk calculates the difference between the Top10-DMR and 116-DMR model for each metric.

```{r calculate-metric-differences}
metric_cols <- c(
  "Accuracy",
  "AUROC",
  "Sensitivity",
  "Specificity",
  "PPV",
  "NPV",
  "F1",
  "Balanced_Accuracy"
)

values_116 <- unlist(
  performance_full_comparison[
    Model == "116-DMR radial SVM",
    ..metric_cols
  ],
  use.names = FALSE
)

values_top10 <- unlist(
  performance_full_comparison[
    Model == "Top10-DMR radial SVM",
    ..metric_cols
  ],
  use.names = FALSE
)

metric_difference_table <- data.table(
  Metric = metric_cols,
  Model_116 = as.numeric(values_116),
  Model_Top10 = as.numeric(values_top10)
)

metric_difference_table[
  ,
  Top10_minus_116 :=
    Model_Top10 - Model_116
]

print(metric_difference_table)
```

Differences close to zero indicate that the reduced panel retains similar predictive performance.

# Compare AUROC values

Because both ROC curves are based on the same TCGA samples, a paired DeLong test is used.

```{r compare-auroc}
delong_test <- tryCatch(
  pROC::roc.test(
    res_116_full$roc,
    res_top10_full$roc,
    paired = TRUE,
    method = "delong"
  ),
  error = function(e) NULL
)

delong_p_value <- if (is.null(delong_test)) {
  NA_real_
} else {
  delong_test$p.value
}

auc_comparison <- data.table(
  AUROC_116 = as.numeric(
    pROC::auc(res_116_full$roc)
  ),
  AUROC_Top10 = as.numeric(
    pROC::auc(res_top10_full$roc)
  ),
  AUROC_difference =
    as.numeric(pROC::auc(res_top10_full$roc)) -
    as.numeric(pROC::auc(res_116_full$roc)),
  DeLong_p_value = delong_p_value
)

print(auc_comparison)
```

A non-significant result means that the AUROC difference is not detected statistically in this cohort, but it does not establish formal equivalence.

# Compare sample-level predictions

This chunk aligns the two models' predictions for every TCGA sample.

```{r prepare-paired-predictions}
prediction_pair <- merge(
  res_116_full$predictions[
    ,
    .(
      Sample,
      True_group,
      Predicted_116 = Predicted_group,
      Probability_116 = Tumor_probability
    )
  ],
  res_top10_full$predictions[
    ,
    .(
      Sample,
      True_group,
      Predicted_Top10 = Predicted_group,
      Probability_Top10 = Tumor_probability
    )
  ],
  by = c("Sample", "True_group")
)

if (nrow(prediction_pair) != nrow(test_df_full)) {
  stop("The paired prediction table has an unexpected number of samples.")
}
```

Every TCGA sample should have one prediction from each model.

This chunk calculates class agreement and the correlation between predicted Tumor probabilities.

```{r assess-prediction-agreement}
prediction_pair[
  ,
  `:=`(
    Correct_116 =
      Predicted_116 == True_group,
    Correct_Top10 =
      Predicted_Top10 == True_group,
    Same_prediction =
      Predicted_116 == Predicted_Top10
  )
]

prediction_agreement <- data.table(
  Class_agreement = mean(
    prediction_pair$Same_prediction
  ),
  Probability_correlation = cor(
    prediction_pair$Probability_116,
    prediction_pair$Probability_Top10,
    method = "spearman"
  )
)

print(prediction_agreement)
```

High class agreement and probability correlation indicate that the Top10 model behaves similarly to the full model.

# Compare classification errors

McNemar's test compares paired correct and incorrect classifications from the two models.

```{r compare-classification-errors}
mcnemar_table <- table(
  Model_116_correct =
    prediction_pair$Correct_116,
  Model_Top10_correct =
    prediction_pair$Correct_Top10
)

mcnemar_result <- tryCatch(
  stats::mcnemar.test(mcnemar_table),
  error = function(e) NULL
)

mcnemar_p_value <- if (is.null(mcnemar_result)) {
  NA_real_
} else {
  mcnemar_result$p.value
}

mcnemar_summary <- data.table(
  McNemar_p_value = mcnemar_p_value
)

print(mcnemar_table)
print(mcnemar_summary)
```

This test assesses whether one model makes significantly more classification errors than the other on the same samples.

# Export performance results

The full-precision and rounded performance tables are saved below.

```{r export-performance-tables}
performance_rounded <- copy(
  performance_full_comparison
)

performance_rounded[
  ,
  (metric_cols) := lapply(
    .SD,
    round,
    digits = 3
  ),
  .SDcols = metric_cols
]

fwrite(
  performance_full_comparison,
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_performance.csv"
  )
)

fwrite(
  performance_rounded,
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_performance_rounded.csv"
  )
)
```

The rounded table is suitable for presentation, while the full-precision table should be retained for analysis.

This chunk saves the metric differences and paired statistical comparisons.

```{r export-comparison-tables}
fwrite(
  cv_performance_comparison,
  file.path(
    outdir,
    "GSE101764_CV_top10_vs_116_performance.csv"
  )
)

fwrite(
  metric_difference_table,
  file.path(
    outdir,
    "TCGA_full_top10_minus_116_metric_differences.csv"
  )
)

fwrite(
  auc_comparison,
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_AUROC_comparison.csv"
  )
)

fwrite(
  mcnemar_summary,
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_McNemar_test.csv"
  )
)
```

These outputs quantify the reduction in performance and the paired statistical comparisons.

This chunk saves the sample-level paired predictions and agreement summary.

```{r export-prediction-tables}
fwrite(
  prediction_pair,
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_paired_predictions.csv"
  )
)

fwrite(
  prediction_agreement,
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_prediction_agreement.csv"
  )
)
```

The paired table identifies samples for which the two models disagree.

# Plot performance metrics

The main performance metrics are converted to long format for plotting.

```{r prepare-performance-plot}
plot_metrics <- c(
  "Accuracy",
  "AUROC",
  "F1",
  "PPV",
  "Sensitivity",
  "Specificity",
  "Balanced_Accuracy"
)

performance_long <- melt(
  performance_full_comparison,
  id.vars = c("Cohort", "Model"),
  measure.vars = plot_metrics,
  variable.name = "Metric",
  value.name = "Value"
)
```

Each metric will be displayed side by side for the two models.

```{r build-performance-plot}
p_performance_full <- ggplot(
  performance_long,
  aes(
    x = Metric,
    y = Value,
    fill = Model
  )
) +
  geom_col(
    position = position_dodge(width = 0.75),
    width = 0.65
  ) +
  geom_text(
    aes(label = sprintf("%.3f", Value)),
    position = position_dodge(width = 0.75),
    vjust = -0.35,
    size = 3.4
  ) +
  coord_cartesian(ylim = c(0, 1.08)) +
  labs(
    title = "TCGA full-cohort performance comparison",
    subtitle = "116-DMR versus Top10-DMR radial SVM",
    x = NULL,
    y = "Metric value",
    fill = "Model"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    legend.position = "top"
  )
```

Similar bar heights indicate that the reduced panel retains performance close to the full panel.

```{r save-performance-plot}
print(p_performance_full)

ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_performance_barplot.png"
  ),
  plot = p_performance_full,
  width = 10,
  height = 6,
  dpi = 300
)

ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_performance_barplot.pdf"
  ),
  plot = p_performance_full,
  width = 10,
  height = 6
)
```

The comparison plot is saved in both PNG and PDF formats.

# Plot ROC curves

The two ROC curves are converted into one plotting table.

```{r prepare-roc-plot}
roc_plot_df <- rbindlist(
  list(
    data.table(
      False_positive_rate =
        1 - res_116_full$roc$specificities,
      Sensitivity =
        res_116_full$roc$sensitivities,
      Model = "116-DMR radial SVM"
    ),
    data.table(
      False_positive_rate =
        1 - res_top10_full$roc$specificities,
      Sensitivity =
        res_top10_full$roc$sensitivities,
      Model = "Top10-DMR radial SVM"
    )
  )
)
```

The two models can now be compared across all classification thresholds.
```{r build-roc-plot}
roc_plot_df$Model <- factor(
  roc_plot_df$Model,
  levels = c(
    "116-DMR radial SVM",
    "Top10-DMR radial SVM"
  )
)

p_roc_comparison <- ggplot(
  roc_plot_df,
  aes(
    x = False_positive_rate,
    y = Sensitivity,
    color = Model,
    linetype = Model,
    linewidth = Model
  )
) +
  geom_step(
    direction = "vh",
    alpha = 0.9
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "grey60",
    linewidth = 0.7
  ) +
  scale_color_manual(
    values = c(
      "116-DMR radial SVM" = "#F8766D",
      "Top10-DMR radial SVM" = "#00BFC4"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "116-DMR radial SVM" = "solid",
      "Top10-DMR radial SVM" = "longdash"
    )
  ) +
  scale_linewidth_manual(
    values = c(
      "116-DMR radial SVM" = 1.8,
      "Top10-DMR radial SVM" = 1.0
    )
  ) +
  coord_equal() +
  labs(
    title = "TCGA full-cohort ROC comparison",
    subtitle = paste0(
      "Paired DeLong p = ",
      signif(delong_p_value, 3)
    ),
    x = "1 - Specificity",
    y = "Sensitivity",
    color = "Model",
    linetype = "Model"
  ) +
  guides(
    linewidth = "none"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0.5
    ),
    legend.position = "top",
    legend.title = element_text(
      face = "bold"
    )
  )
```

Substantial overlap between the curves indicates similar discrimination.

```{r save-roc-plot}
print(p_roc_comparison)

ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_ROC_comparison.png"
  ),
  plot = p_roc_comparison,
  width = 7,
  height = 7,
  dpi = 300
)

ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_ROC_comparison.pdf"
  ),
  plot = p_roc_comparison,
  width = 7,
  height = 7
)
```

The ROC comparison is saved in both PNG and PDF formats.
## Assess and visualise sample-level prediction agreement
```{r}
# Assess and visualise sample-level prediction agreement

## Confirm that the paired prediction table is complete.
stopifnot(
  nrow(prediction_pair) == nrow(test_df_full),
  !anyNA(prediction_pair$Predicted_116),
  !anyNA(prediction_pair$Predicted_Top10),
  !anyNA(prediction_pair$Probability_116),
  !anyNA(prediction_pair$Probability_Top10)
)

## Use identical factor levels for the two model predictions.
pred_116_factor <- factor(
  prediction_pair$Predicted_116,
  levels = c("Tumor", "Normal")
)

pred_top10_factor <- factor(
  prediction_pair$Predicted_Top10,
  levels = c("Tumor", "Normal")
)
```
## Calculate the proportion of samples receiving the same class prediction.
```{r}
class_agreement <- mean(
  pred_116_factor == pred_top10_factor
)

## Calculate Cohen's kappa between the two models.
## Here, the 116-DMR prediction is treated as the reference prediction,
## not as the biological ground truth.
prediction_kappa_cm <- caret::confusionMatrix(
  data = pred_top10_factor,
  reference = pred_116_factor
)

cohen_kappa <- unname(
  prediction_kappa_cm$overall["Kappa"]
)

## Calculate the Spearman correlation between predicted
## Tumor probabilities from the two models.
probability_correlation <- cor(
  prediction_pair$Probability_116,
  prediction_pair$Probability_Top10,
  method = "spearman",
  use = "complete.obs"
)

## Calculate the average absolute difference in predicted probability.
mean_absolute_probability_difference <- mean(
  abs(
    prediction_pair$Probability_116 -
      prediction_pair$Probability_Top10
  ),
  na.rm = TRUE
)

## Build the complete agreement summary.
prediction_agreement <- data.table(
  Class_agreement = class_agreement,
  Cohen_kappa = cohen_kappa,
  Probability_correlation = probability_correlation,
  Mean_absolute_probability_difference =
    mean_absolute_probability_difference
)

cat("\nSample-level prediction agreement:\n")
print(prediction_agreement)

cat("\nPrediction agreement confusion matrix:\n")
print(prediction_kappa_cm$table)
```

## Build a model-to-model prediction agreement table.
```{r}
## Build a model-to-model prediction agreement table.
prediction_agreement_matrix <- table(
  Prediction_116 = pred_116_factor,
  Prediction_Top10 = pred_top10_factor
)

agreement_heatmap_df <- as.data.table(
  as.data.frame(prediction_agreement_matrix)
)

setnames(
  agreement_heatmap_df,
  c(
    "Prediction_116",
    "Prediction_Top10",
    "Count"
  )
)

agreement_heatmap_df[
  ,
  Percent := Count / sum(Count)
]

print(agreement_heatmap_df)
```
## plot showing class agreement,Cohen’s kappa and predicted probability correlation
```{r}
## Prepare agreement metrics for plotting.
agreement_metric_plot_df <- data.table(
  Metric = factor(
    c(
      "Class agreement",
      "Cohen's kappa",
      "Probability correlation"
    ),
    levels = c(
      "Class agreement",
      "Cohen's kappa",
      "Probability correlation"
    )
  ),
  Value = c(
    class_agreement,
    cohen_kappa,
    probability_correlation
  )
)

p_agreement_summary <- ggplot(
  agreement_metric_plot_df,
  aes(
    x = Metric,
    y = Value
  )
) +
  geom_col(
    width = 0.65,
    fill = "grey50"
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.3f",
        Value
      )
    ),
    vjust = -0.4,
    size = 4
  ) +
  coord_cartesian(
    ylim = c(0, 1.08)
  ) +
  labs(
    title = "Agreement between the 116-DMR and Top10-DMR models",
    subtitle = paste0(
      "Mean absolute probability difference = ",
      sprintf(
        "%.3f",
        mean_absolute_probability_difference
      )
    ),
    x = NULL,
    y = "Agreement value"
  ) +
  theme_bw(
    base_size = 11
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0.5
    ),
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    ),
    panel.grid.major.x = element_blank()
  )

print(p_agreement_summary)
```

## Label whether the two models produced the same class prediction.
```{r}
## Label whether the two models produced the same class prediction.
prediction_pair[
  ,
  Prediction_agreement_status := factor(
    ifelse(
      Same_prediction,
      "Same predicted class",
      "Different predicted class"
    ),
    levels = c(
      "Same predicted class",
      "Different predicted class"
    )
  )
]

p_probability_agreement <- ggplot(
  prediction_pair,
  aes(
    x = Probability_116,
    y = Probability_Top10,
    color = True_group,
    shape = Prediction_agreement_status
  )
) +
  geom_point(
    alpha = 0.70,
    size = 2.2
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    linewidth = 0.8
  ) +
  coord_equal(
    xlim = c(0, 1),
    ylim = c(0, 1)
  ) +
  labs(
    title = "Agreement in predicted Tumor probabilities",
    subtitle = paste0(
      "Spearman correlation = ",
      sprintf(
        "%.3f",
        probability_correlation
      ),
      "; class agreement = ",
      sprintf(
        "%.1f%%",
        100 * class_agreement
      ),
      "; Cohen's kappa = ",
      sprintf(
        "%.3f",
        cohen_kappa
      )
    ),
    x = "Tumor probability from 116-DMR model",
    y = "Tumor probability from Top10-DMR model",
    color = "True group",
    shape = "Model agreement"
  ) +
  theme_bw(
    base_size = 11
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 11
    ),
    axis.title = element_text(
      size = 11
    ),
    axis.text = element_text(
      size = 9
    ),
    legend.position = "top",
    legend.box = "vertical"
  )

print(p_probability_agreement)
```
## heatmap
```{r}
p_class_agreement_heatmap <- ggplot(
  agreement_heatmap_df,
  aes(
    x = Prediction_116,
    y = Prediction_Top10,
    fill = Count
  )
) +
  geom_tile(
    color = "white",
    linewidth = 1
  ) +
  geom_text(
    aes(
      label = paste0(
        Count,
        "\n",
        sprintf(
          "%.1f%%",
          100 * Percent
        )
      )
    ),
    size = 5
  ) +
  labs(
    title = "Sample-level class agreement between the two models",
    subtitle = paste0(
      "Overall class agreement = ",
      sprintf(
        "%.1f%%",
        100 * class_agreement
      )
    ),
    x = "116-DMR model prediction",
    y = "Top10-DMR model prediction",
    fill = "Number of samples"
  ) +
  theme_bw(
    base_size = 11
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0.5
    ),
    panel.grid = element_blank()
  )

print(p_class_agreement_heatmap)
```
## save plot results
```{r}
## Export the expanded agreement summary.
fwrite(
  prediction_agreement,
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_prediction_agreement.csv"
  )
)

## Export the prediction-agreement matrix.
fwrite(
  agreement_heatmap_df,
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_prediction_agreement_matrix.csv"
  )
)

## Export the three plotted agreement metrics.
fwrite(
  agreement_metric_plot_df,
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_agreement_metrics.csv"
  )
)
```
## save plots
```{r}
## Save the agreement summary plot.
ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_agreement_summary.png"
  ),
  plot = p_agreement_summary,
  width = 8,
  height = 6,
  dpi = 300
)

ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_agreement_summary.pdf"
  ),
  plot = p_agreement_summary,
  width = 8,
  height = 6
)

## Save the probability agreement plot.
ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_probability_agreement.png"
  ),
  plot = p_probability_agreement,
  width = 8,
  height = 7,
  dpi = 300
)

ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_probability_agreement.pdf"
  ),
  plot = p_probability_agreement,
  width = 8,
  height = 7
)

## Save the class-agreement heatmap.
ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_class_agreement_heatmap.png"
  ),
  plot = p_class_agreement_heatmap,
  width = 7,
  height = 6,
  dpi = 300
)

ggsave(
  file.path(
    outdir,
    "TCGA_full_top10_vs_116_class_agreement_heatmap.pdf"
  ),
  plot = p_class_agreement_heatmap,
  width = 7,
  height = 6
)
```

# Save analysis objects

The fitted models and all comparison objects are saved for later review.

```{r save-analysis-objects}
save(
  features_116,
  features_top10,
  cv_index,
  model_svm_116_retrained,
  model_svm_top10,
  cv_performance_comparison,
  performance_full_comparison,
  metric_difference_table,
  prediction_pair,
  prediction_agreement,
  prediction_kappa_cm,
  prediction_agreement_matrix,
  agreement_heatmap_df,
  agreement_metric_plot_df,
  auc_comparison,
  delong_test,
  mcnemar_table,
  mcnemar_result,
  p_performance_full,
  p_roc_comparison,
  p_agreement_summary,
  p_probability_agreement,
  p_class_agreement_heatmap,
  file = file.path(
    outdir,
    "TCGA_full_top10_vs_116_radial_SVM_results.Rdata"
  )
)
```

This RData file allows the tables and figures to be reopened without retraining the models.

# Session information

The R version and package dependencies are listed below.

```{r session-information}
sessionInfo()
```
