---
title: "Step 7.1 - DMR CpG layout visualisation"
author: "Yanni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 7.1 Purpose of DMR CpG layout visualisation
This step generates publication-style CpG genomic layout plots for the prioritised DMR candidates.

The visualisation integrates:

Consensus representative CpG ranking results
Original DMRcate EPIC probe coordinates
Genomic positions of constituent CpG probes

Each DMR will be visualised separately, showing:

DMR genomic interval
All constituent CpG probes
Consensus representative rank of each CpG
Future primer/amplicon annotation space for assay design

### Clear environment and define output directory
```{r }
rm(list = ls())
outdir <- "step7.1_probes_visualisation"


if(!dir.exists(outdir)){
  
  dir.create(
    outdir,
    recursive = TRUE
  )
}

```
### Load required packages
```{r}
library(data.table)
library(ggplot2)
library(svglite)

```
### Load consensus probe ranking table and EPIC probe coordinate mapping table and Check required columns
```{r}
probe_rank <- fread(
  "Top20_DMR_probe_consensus_ranking_across_discovery_datasets.csv")
probe_mapping <- fread(
  "Top20_priority_DMR_actual_EPIC_probes_used_by_DMRcate_nsim500.csv")


required_rank_columns <- c(
  
  "Current_DMR_ID",
  "CpG",
  "Consensus_representative_rank")

required_mapping_columns <- c(
  
  "DMR_ID",
  "probe_id",
  "probe_chr",
  "probe_position",
  "start",
  "end")

if(
  !all(required_rank_columns %in% colnames(probe_rank))
){
  
  stop("Missing required columns in probe ranking table.")}


if(
  !all(required_mapping_columns %in% colnames(probe_mapping))
){
  
  stop("Missing required columns in probe mapping table.")}
```
### Standardise probe coordinate table
Rename genomic coordinate columns to create a consistent structure for downstream plotting.
The final identifier format:Current_DMR_ID + CpG
```{r}
mapping_for_plot <- probe_mapping[
  
  ,
  
  .(
    
    Current_DMR_ID = DMR_ID,
    
    CpG = probe_id,
    
    chr = probe_chr,
    
    position = probe_position,
    
    DMR_start = start,
    
    DMR_end = end)]
```
### Merge probe ranking with genomic coordinates
Combine:
Consensus representative ranking information
CpG genomic location information
After merging, each CpG contains:
DMR identity
CpG ID
genomic coordinate
consensus rank
```{r}
plot_data <- merge(
  
  probe_rank,
  
  mapping_for_plot,
  
  by = c(
    "Current_DMR_ID",
    "CpG"
  ),
  
  all.x = TRUE)
cat("Merged plotting dataset:",
  nrow(plot_data),
  "rows x",
  ncol(plot_data),
  "columns\n")
```
### Validate genomic coordinate matching
Check whether all ranked CpGs have successfully matched EPIC probe coordinates.
```{r}
missing_position <- sum(is.na(plot_data$position))
cat("CpGs without genomic position:",
  missing_position,
  "\n")
if(missing_position > 0){
  
  warning("Some CpGs could not be matched with genomic coordinates.")}
```
### Arrange plotting dataset
Order CpGs by:DMR identity ; Genomic position

This ordering will be used for genomic layout visualisation.
```{r}
setorder(plot_data,
  
  Current_DMR_ID,
  
  position)

plot_data[
  ,
  Consensus_representative_rank :=
    as.numeric(Consensus_representative_rank)]
```
### Generate DMR summary table
Generate a DMR-level summary containing:
number of CpGs
chromosome
genomic interval
```{r}
dmr_summary <- plot_data[
  
  ,
  
  .(n_CpGs = .N,
    
    chr = unique(chr),
    
    DMR_start = unique(DMR_start),
    
    DMR_end = unique(DMR_end)),
  by = Current_DMR_ID]
dmr_summary
```
### save plotting datasets
```{r}
fwrite(plot_data,
  file.path(outdir,
    "DMR_CpG_layout_plotting_dataset.csv"))

fwrite(dmr_summary,
  file.path(outdir,
    "DMR_CpG_layout_summary.csv"))
```
### Prepare plotting data for individual DMR
This function extracts CpG probes belonging to one selected DMR and assigns vertical positions for the lollipop-style genomic layout.

The plotting design includes:

- two-level CpG heights to reduce overlap
- compact CpG track layout
- reserved upper space for future primer/amplicon annotation
```{r prepare-single-DMR-function}
## Prepare plotting data for individual DMR
## Purpose:
## Extract CpG probes belonging to one selected DMR and
## prepare lollipop plot coordinates.
## Visualisation strategy:
## 1. CpG number controls vertical layers:
##    <=15 CpGs  : 2 layers
##    >15 CpGs   : 3 layers
## 2. Genomic coordinates are retained.
## 3. Closely located CpGs (<50 bp) receive small horizontal
##    offsets only for visualisation to reduce overlap.
prepare_single_DMR_plot_data <- function(
  plot_data,
  selected_DMR
){
  ## Extract CpG information for selected DMR
  dmr_plot_data <- copy(
    
    plot_data[
      Current_DMR_ID == selected_DMR])
  
  if(
    nrow(dmr_plot_data) == 0
  ){
    
    stop(
      paste("No data found for",
        selected_DMR))}
  ## Arrange CpGs according to genomic position
  setorder(
    dmr_plot_data,
    position
  )
  ## Count CpGs within selected DMR
  n_CpGs <- nrow(
    dmr_plot_data
  )
  ## Determine number of vertical layers
  if(
    n_CpGs <= 15
  ){
    
    n_layers <- 2
    
    
  }else{
    
    n_layers <- 3
    
  }
  ## Assign CpGs into vertical layers
  ## CpGs are sequentially distributed according to genomic
  ## order.
  dmr_plot_data[
    ,
    plot_layer :=
      seq_len(.N) %% n_layers
  ]
  
  ## Convert layer number into plotting height
  
  if(
    n_layers == 2
  ){
    
    height_levels <- c(
      0.12,
      0.22
    )
    
    
  }else{
    
    height_levels <- c(
      0.10,
      0.18,
      0.26
    )
    
  }
  
  dmr_plot_data[
    ,
    plot_y :=
      height_levels[
        plot_layer + 1
      ]
  ]
  
  ## Add horizontal jitter for closely located CpGs
  ## Genomic position is preserved.
  ## plot_x is only used for visualisation.
  dmr_plot_data[
    ,
    plot_x := position
  ]
  
  
  
  close_threshold <- 50
  
  
  for(i in seq_len(nrow(dmr_plot_data)-1)){
    
    
    if(
      abs(
        dmr_plot_data$position[i+1] -
        dmr_plot_data$position[i]
      ) < close_threshold
    ){
      
      
      dmr_plot_data$plot_x[i+1] <-
        
        dmr_plot_data$plot_x[i] + 20
      
    }
    
    
  }
  
  ## Extract DMR genomic information
  dmr_start <- unique(
    dmr_plot_data$DMR_start
  )
  
  
  dmr_end <- unique(
    dmr_plot_data$DMR_end
  )
  
  
  dmr_chr <- unique(dmr_plot_data$chr)
  priority_rank <- unique(
    dmr_plot_data$priority_rank
  )
  ## Generate plot title
  ## Example:
  ## Rank 1: DMR_33 | chr6:62995523-62997123
  coordinate <- paste0(
    
    dmr_chr,
    
    ":",
    
    dmr_start,
    
    "-",
    
    dmr_end)
  plot_title <- paste0(
    
    "Rank ",
    
    priority_rank,
    
    ": ",
    
    selected_DMR,
    
    " | ",
    
    coordinate)
  ## Return plotting objects
  return(
    
    list(
      
      dmr_plot_data = dmr_plot_data,
      
      dmr_start = dmr_start,
      
      dmr_end = dmr_end,
      
      plot_title = plot_title))}
```

### Generate single DMR lollipop plot function
This function generates one CpG layout plot for a selected DMR.
The plot contains:
DMR genomic region baseline
CpG genomic positions
lollipop stems
white CpG circles
consensus rank displayed inside each circle
```{r}
plot_single_DMR_layout <- function(
  plot_data,
  selected_DMR
){
  ## Prepare plotting data
  prepared <- prepare_single_DMR_plot_data(
    
    plot_data = plot_data,
    
    selected_DMR = selected_DMR)
  dmr_plot_data <- prepared$dmr_plot_data
  
  dmr_start <- prepared$dmr_start
  
  dmr_end <- prepared$dmr_end
  
  plot_title <- prepared$plot_title
  ## Add x-axis padding
  ## Prevent CpGs close to DMR boundaries from touching
  ## the edge of the figure.
  x_padding <- max(
    
    20,
    
    round((dmr_end - dmr_start) * 0.05))
  ## Generate lollipop plot
  p <- ggplot() +
    ## DMR genomic baseline
    geom_segment(
      
      aes(
        
        x = dmr_start,
        
        xend = dmr_end,
        
        y = 0,
        
        yend = 0
        
      ),
      
      linewidth = 1.2,
      
      colour = "black"
      
    ) +
    ## CpG vertical stems
    geom_segment(
      
      data = dmr_plot_data,
      
      aes(
        
        x = position,
        
        xend = position,
        
        y = 0,
        
        yend = plot_y
        
      ),
      
      linewidth = 0.9,
      
      colour = "black"
      
    ) +
    ## CpG circles
    geom_point(
      
      data = dmr_plot_data,
      
      aes(
        
        x = position,
        
        y = plot_y
        
      ),
      
      shape = 21,
      
      fill = "white",
      
      colour = "black",
      
      size = 4.6,
      
      stroke = 0.8
      
    ) +
    ## Consensus rank labels inside CpG circles
    geom_text(
      
      data = dmr_plot_data,
      
      aes(
        
        x = position,
        
        y = plot_y,
        
        label = Consensus_representative_rank
        
      ),
      
      size = 2.8,
      
      colour = "black"
      
    ) +
    ## Axis scaling
    ## Upper space is reserved for future primer annotation.
    scale_x_continuous(
      
      limits = c(
        
        dmr_start - x_padding,
        
        dmr_end + x_padding
        
      ),
      
      expand = c(
        0,
        0
      )
      
    ) +
    scale_y_continuous(
      
      limits = c(
        
        -0.03,
        
        0.50
        
      ),
      
      expand = c(
        0,
        0
      )
      
    ) +
    ## Figure labels
    labs(
      
      title = plot_title,
      
      x = "Genomic position",
      
      y = NULL
      
    ) +
    ## Figure theme
    theme_classic(
      base_size = 14
    ) +
    
    theme(
      
      plot.title = element_text(
        
        face = "bold",
        
        size = 16
        
      ),
      
      axis.title.x = element_text(
        
        size = 14
        
      ),
      
      axis.text.x = element_text(
        
        size = 11
        
      ),
      
      axis.text.y = element_blank(),
      
      axis.ticks.y = element_blank(),
      
      axis.line.y = element_blank(),
      
      plot.margin = margin(
        
        t = 12,
        
        r = 15,
        
        b = 10,
        
        l = 15))
return(p)
}
```
### Test one example DMR
Before generating all Top 20 DMR figures, visually confirm:
title format
CpG spacing
rank labels
reserved primer space
```{r}
## Test plotting function using one example DMR
example_DMR <- "DMR_61"
p_DMR61 <- plot_single_DMR_layout(
  
  plot_data = plot_data,
  
  selected_DMR = example_DMR)
p_DMR61
```
### Create figure output directories
Create separate folders for PDF and SVG output formats.
SVG format will be used for:
Illustrator editing
poster figure assembly
manual primer annotation
PDF format will be used for:
manuscript supplementary figures
high-quality vector export
```{r}
pdf_dir <- file.path(outdir,"PDF")

svg_dir <- file.path(outdir,"SVG")

if(
  !dir.exists(pdf_dir)){dir.create(
    pdf_dir,
    recursive = TRUE)}

if(!dir.exists(svg_dir)){dir.create(
    svg_dir,
    recursive = TRUE)}
```
### Define DMR list for plotting
Extract all candidate DMR IDs included in the probe prioritisation analysis.
Each DMR will generate one independent genomic layout figure.
```{r}
DMR_to_plot <- unique(plot_data$Current_DMR_ID)
cat("Number of DMRs to plot:",
  length(DMR_to_plot),
  "\n")
print(DMR_to_plot)
```
### Generate and save all DMR plots
Loop through all prioritised DMRs and save:PDF figures & SVG figures

Each figure contains:DMR genomic region；CpG positions；Consensus representative ranks
```{r}
figure_index <- data.table()
for(
  this_DMR in DMR_to_plot
){
  ## Generate plot
  p_current <- plot_single_DMR_layout(
    
    plot_data = plot_data,
    
    selected_DMR = this_DMR
    
  )
  ## Extract metadata for file naming
  current_metadata <- unique(
    
    plot_data[
      
      Current_DMR_ID == this_DMR,
      
      .(
        
        priority_rank,
        
        chr,
        
        DMR_start,
        
        DMR_end
        
      )
      
    ]
    
  )
  current_rank <- current_metadata$priority_rank[1]
  
  current_chr <- current_metadata$chr[1]
  
  current_start <- current_metadata$DMR_start[1]
  
  current_end <- current_metadata$DMR_end[1]
  ## Create figure filename
  figure_name <- paste0(
    
    "Rank",
    
    current_rank,
    
    "_",
    
    this_DMR,
    
    "_chr",
    
    current_chr,
    
    "_",
    
    current_start,
    
    "-",
    
    current_end,
    
    "_CpG_layout")
  ## Save PDF
  ggsave(filename = file.path(pdf_dir,
      paste0(
        figure_name,
        ".pdf")),
    
    plot = p_current,
    
    width = 8.5,
    
    height = 4.5,
    
    units = "in")
  ## Save SVG
  ggsave(filename = file.path(svg_dir,
  paste0(figure_name,".svg")
    ),
    
    plot = p_current,
    
    width = 8.5,
    
    height = 4.5,
    
    units = "in")
  ## Record figure information
  figure_index <- rbind(
    
    figure_index,
    
    data.table(
      
      priority_rank = current_rank,
      
      DMR_ID = this_DMR,
      
      chromosome = current_chr,
      
      start = current_start,
      
      end = current_end,
      
      figure_name = figure_name))
  cat("Saved:",figure_name,"\n")}
```
### Save figure index table
Generate a summary table linking:
DMR identity
genomic coordinates
priority ranking
exported figure names
This table will facilitate:
manual figure assembly
primer annotation
manuscript supplementary figure preparation
```{r}
setorder(figure_index,priority_rank)
fwrite(figure_index,
  file.path(outdir,"DMR_layout_figure_index.csv"))
figure_index
```

### Record R session information
```{r}
sessionInfo()
```

