---
title: "Step 7. Probe prioritisation across discovery datasets"
author: "YC"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 8
    fig_height: 7
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This report prioritises representative CpG probes within the selected Top 20 DMRs.

For each DMR, the analysis:

1. uses the actual CpG probes contributing to the Step 1 DMRcate region;
2. calculates a sample-level DMR mean beta value separately in GSE199057, GSE149282, and GSE119526;
3. compares each CpG with the corresponding DMR mean across samples;
4. calculates both the direct correlation with the full DMR mean and a leave-one-CpG-out correlation;
5. combines the three dataset-specific results into a cross-dataset consensus ranking;
6. selects one primary representative CpG and two backup CpGs per DMR.

The direct full-mean correlation follows the simple approach proposed by Mark. The leave-one-CpG-out Spearman correlation is used as the primary ranking criterion because it avoids including the tested CpG in its own reference mean.

**Interpretation note:** the selected probes represent the DMR methylation pattern across the three discovery cohorts. Final wet-lab selection should additionally consider assay design, surrounding sequence, SNPs, cross-reactivity, and other technical constraints.

# 1. Initialise analysis

This chunk clears the current environment and defines general display options.

```{r initialise-analysis}
rm(list = ls())

options(stringsAsFactors = FALSE)
options(scipen = 20)
```

Starting from a clean environment reduces the risk of using objects left from an earlier analysis.

# 2. Load packages

The required packages provide data handling and visualisation functions.

```{r load-packages}
library(data.table)
library(ggplot2)
```

Successful package loading confirms that the analysis environment is ready.

# 3. Define input and output files

The three processed discovery datasets provide sample-level CpG beta values. The Top 20 files define the selected DMRs and their actual DMRcate constituent probes.

```{r define-input-files}
gse199057_file <-
  "step0.1-GSE199057_output.Rdata"

gse149282_file <-
  "step0.2-GSE149282_output.Rdata"

gse119526_file <-
  "step0.3-GSE119526_output.Rdata"

top20_dmr_file <-
  "Top20_priority_DMR_matched_to_original_DMRcate_nsim500.csv"

dmr_probe_mapping_file <-
  "Top20_priority_DMR_actual_EPIC_probes_used_by_DMRcate_nsim500.csv"

probe_audit_file <-
  "Top20_priority_DMR_probe_summary_and_audit_nsim500.csv"

outdir <-
  "step7.probe prioritisation"
```

Replace these names with full paths only when the files are stored in different directories.

# 4. Check input files and create output directories

The analysis stops before loading data if any required file is unavailable.

```{r check-input-files}
input_files <- c(
  gse199057_file,
  gse149282_file,
  gse119526_file,
  top20_dmr_file,
  dmr_probe_mapping_file,
  probe_audit_file
)

if (!all(file.exists(input_files))) {
  stop(
    "Missing input file(s): ",
    paste(
      input_files[!file.exists(input_files)],
      collapse = ", "
    )
  )
}

if (!dir.exists(outdir)) {
  dir.create(
    outdir,
    recursive = TRUE
  )
}

scatter_plot_dir <- file.path(
  outdir,
  "selected_probe_scatterplots"
)

if (!dir.exists(scatter_plot_dir)) {
  dir.create(
    scatter_plot_dir,
    recursive = TRUE
  )
}
```

Successful completion confirms that all inputs are present and the result directories are ready.

# 5. Define the RData loading function

This helper function loads `myLoad` into a temporary environment so that repeated object names do not overwrite one another.

```{r define-load-myload-function}
load_myload <- function(file) {

  if (!file.exists(file)) {
    stop(
      "Input file does not exist: ",
      file
    )
  }

  temp_env <- new.env()

  loaded_objects <- load(
    file = file,
    envir = temp_env
  )

  if (!"myLoad" %in% loaded_objects) {
    stop(
      "The file ",
      file,
      " does not contain an object named 'myLoad'."
    )
  }

  temp_env$myLoad
}
```

The function returns only the expected `myLoad` object from each processed dataset.

# 6. Load the processed discovery datasets

The three discovery datasets are loaded independently.

```{r load-discovery-datasets}
obj_199057 <- load_myload(
  gse199057_file
)

obj_149282 <- load_myload(
  gse149282_file
)

obj_119526 <- load_myload(
  gse119526_file
)
```

Each loaded object should contain a filtered beta matrix and its corresponding phenotype table.

# 7. Extract beta matrices and phenotype data

The original object names used in Step 1 are retained.

```{r extract-beta-and-phenotype-data}
g1 <- obj_199057$beta
pd1 <- obj_199057$pd

g2 <- obj_149282$beta
pd2 <- obj_149282$pd

g3 <- obj_119526$beta
pd3 <- obj_119526$pd
```

`g1`, `g2`, and `g3` contain CpGs in rows and samples in columns. The corresponding `pd` objects contain sample annotations.

# 8. Define a sample-alignment function

This function verifies and restores exact correspondence between beta-matrix columns and phenotype rows.

```{r define-alignment-function}
align_beta_and_pd <- function(
    beta,
    pd,
    dataset_name) {

  beta <- as.matrix(beta)
  storage.mode(beta) <- "numeric"
  pd <- as.data.frame(pd)

  if (!"group" %in% colnames(pd)) {
    stop(
      dataset_name,
      ": column 'group' was not found in pd."
    )
  }

  if ("ID" %in% colnames(pd)) {
    if (!setequal(
      colnames(beta),
      as.character(pd$ID)
    )) {
      stop(
        dataset_name,
        ": beta columns and pd$ID do not contain the same samples."
      )
    }

    pd <- pd[
      match(
        colnames(beta),
        as.character(pd$ID)
      ),
      ,
      drop = FALSE
    ]

    rownames(pd) <- as.character(
      pd$ID
    )
  }

  if (!identical(
    colnames(beta),
    rownames(pd)
  )) {
    stop(
      dataset_name,
      ": sample alignment failed."
    )
  }

  pd$group <- factor(
    as.character(pd$group),
    levels = c(
      "Normal",
      "Tumor"
    )
  )

  if (anyNA(pd$group)) {
    stop(
      dataset_name,
      ": unexpected or missing group labels were found."
    )
  }

  list(
    beta = beta,
    pd = pd
  )
}
```

The function also standardises the group order as Normal followed by Tumor.

# 9. Align the three discovery datasets

The alignment function is applied separately to each beta matrix and phenotype table.

```{r align-discovery-datasets}
aligned_199057 <- align_beta_and_pd(
  g1,
  pd1,
  "GSE199057"
)

aligned_149282 <- align_beta_and_pd(
  g2,
  pd2,
  "GSE149282"
)

aligned_119526 <- align_beta_and_pd(
  g3,
  pd3,
  "GSE119526"
)

g1 <- aligned_199057$beta
pd1 <- aligned_199057$pd

g2 <- aligned_149282$beta
pd2 <- aligned_149282$pd

g3 <- aligned_119526$beta
pd3 <- aligned_119526$pd
```

All three phenotype tables now follow the exact sample order of their corresponding beta matrices.

# 10. Summarise the discovery datasets

This table documents the CpG, sample, Normal, and Tumor counts available in each discovery dataset.

```{r summarise-discovery-datasets}
discovery_dataset_summary <- data.table(
  Dataset = c(
    "GSE199057",
    "GSE149282",
    "GSE119526"
  ),
  CpGs_before_common_filter = c(
    nrow(g1),
    nrow(g2),
    nrow(g3)
  ),
  Samples = c(
    ncol(g1),
    ncol(g2),
    ncol(g3)
  ),
  Normal = c(
    sum(pd1$group == "Normal"),
    sum(pd2$group == "Normal"),
    sum(pd3$group == "Normal")
  ),
  Tumor = c(
    sum(pd1$group == "Tumor"),
    sum(pd2$group == "Tumor"),
    sum(pd3$group == "Tumor")
  )
)

print(
  discovery_dataset_summary
)
```

The group counts should be checked against the corrected phenotype assignments used in Step 1.

# 11. Identify CpGs shared across all discovery datasets

The common CpG universe is reconstructed exactly as in Step 1.

```{r identify-common-cpgs}
common_cpg <- Reduce(
  intersect,
  list(
    rownames(g1),
    rownames(g2),
    rownames(g3)
  )
)

cat(
  "Number of CpG probes shared across all three datasets:",
  length(common_cpg),
  "\n"
)

if (length(common_cpg) == 0L) {
  stop(
    "No common CpG probes were found across the three datasets."
  )
}
```

A non-zero result confirms that the datasets share probes that can be compared consistently.

# 12. Restrict beta matrices to common CpGs

Using the same `common_cpg` vector places the CpGs in an identical order in all three matrices.

```{r subset-to-common-cpgs}
g1 <- g1[
  common_cpg,
  ,
  drop = FALSE
]

g2 <- g2[
  common_cpg,
  ,
  drop = FALSE
]

g3 <- g3[
  common_cpg,
  ,
  drop = FALSE
]

stopifnot(
  identical(
    rownames(g1),
    rownames(g2)
  ),
  identical(
    rownames(g1),
    rownames(g3)
  )
)

cat(
  "All three matrices contain",
  nrow(g1),
  "common CpGs in the same order.\n"
)
```

The resulting matrices use the same CpG universe as the discovery meta-analysis.

# 13. Load Top 20 DMR and probe-mapping tables

The Top 20 region table, actual DMRcate probe mapping, and existing probe audit are imported.

```{r load-top20-input-tables}
top20_dmrs_raw <- fread(
  top20_dmr_file
)

dmr_probe_mapping_raw <- fread(
  dmr_probe_mapping_file
)

probe_audit_raw <- fread(
  probe_audit_file
)

cat("\nTop 20 DMR table columns:\n")
print(names(top20_dmrs_raw))

cat("\nDMR-probe mapping table columns:\n")
print(names(dmr_probe_mapping_raw))

cat("\nProbe audit table columns:\n")
print(names(probe_audit_raw))
```

The printed column names provide an audit trail of the imported source-table structures.

# 14. Define table-standardisation helpers

These helpers identify common column-name variants without modifying the original imported tables.

```{r define-table-standardisation-helpers}
find_column <- function(
    data,
    candidates,
    description,
    required = TRUE) {

  matched <- candidates[
    candidates %in% names(data)
  ]

  if (length(matched) == 0) {
    if (required) {
      stop(
        "Could not identify ",
        description,
        ". Available columns: ",
        paste(
          names(data),
          collapse = ", "
        )
      )
    }

    return(NULL)
  }

  matched[1]
}

normalise_coordinate <- function(x) {
  x <- as.character(x)
  x <- gsub(
    ",",
    "",
    x
  )
  gsub(
    "\\s+",
    "",
    x
  )
}
```

Standardised working copies are constructed below, while the original tables are retained for auditing.

# 15. Standardise the Top 20 DMR table

This chunk identifies the current DMR identifier, stable coordinate, and priority-rank columns.

```{r identify-top20-dmr-columns}
top20_id_column <- find_column(
  top20_dmrs_raw,
  c(
    "Current_DMR_ID",
    "current_DMR_ID",
    "DMR_ID",
    "DMR"
  ),
  "the current DMR identifier"
)

top20_coordinate_column <- find_column(
  top20_dmrs_raw,
  c(
    "coordinate_key",
    "DMR_coordinate",
    "discovery_coord",
    "coord",
    "DMR_id"
  ),
  "the DMR coordinate"
)

top20_rank_column <- find_column(
  top20_dmrs_raw,
  c(
    "priority_rank",
    "Priority_rank",
    "Rank",
    "rank"
  ),
  "the priority rank",
  required = FALSE
)
```

When no explicit rank column is present, the source-table row order is retained.

```{r build-standard-top20-dmr-table}
top20_dmrs <- data.table(
  Current_DMR_ID = as.character(
    top20_dmrs_raw[[
      top20_id_column
    ]]
  ),
  coordinate_key = normalise_coordinate(
    top20_dmrs_raw[[
      top20_coordinate_column
    ]]
  )
)

if (is.null(top20_rank_column)) {
  top20_dmrs[
    ,
    priority_rank := seq_len(.N)
  ]
} else {
  top20_dmrs[
    ,
    priority_rank := as.integer(
      top20_dmrs_raw[[
        top20_rank_column
      ]]
    )
  ]
}

top20_dmrs <- unique(
  top20_dmrs
)

setorder(
  top20_dmrs,
  priority_rank
)

top20_dmrs[
  ,
  DMR_order := seq_len(.N)
]
```

The standardised working table uses stable names required by the subsequent analysis.

# 16. Check the Top 20 DMR table

The selected table must contain 20 unique DMR identifiers and 20 unique coordinates.

```{r check-top20-dmr-table}
if (nrow(top20_dmrs) != 20) {
  stop(
    "The standardised Top 20 table does not contain exactly 20 DMRs."
  )
}

if (anyDuplicated(
  top20_dmrs$Current_DMR_ID
) > 0) {
  stop(
    "Duplicated Current_DMR_ID values were found."
  )
}

if (anyDuplicated(
  top20_dmrs$coordinate_key
) > 0) {
  stop(
    "Duplicated coordinate_key values were found."
  )
}

print(
  top20_dmrs
)
```

The printed table defines the DMR order used in all result tables and figures.

# 17. Identify columns in the actual DMRcate probe mapping

The CpG column is required. The mapping can be linked using either the stable coordinate or the current DMR identifier.

```{r identify-probe-mapping-columns}
mapping_cpg_column <- find_column(
  dmr_probe_mapping_raw,
  c(
    "CpG",
    "cpg",
    "Probe",
    "probe",
    "Probe_ID",
    "probe_id",
    "CpG_ID",
    "cpg_id"
  ),
  "the CpG probe identifier"
)

mapping_coordinate_column <- find_column(
  dmr_probe_mapping_raw,
  c(
    "coordinate_key",
    "DMR_coordinate",
    "discovery_coord",
    "coord",
    "DMR_id"
  ),
  "the DMR coordinate",
  required = FALSE
)

mapping_id_column <- find_column(
  dmr_probe_mapping_raw,
  c(
    "Current_DMR_ID",
    "current_DMR_ID",
    "DMR_ID",
    "DMR"
  ),
  "the DMR identifier",
  required = FALSE
)

if (
  is.null(mapping_coordinate_column) &&
  is.null(mapping_id_column)
) {
  stop(
    "The probe mapping must contain a DMR coordinate or identifier."
  )
}
```

A stable region link is required for every probe.

# 18. Standardise the DMR–CpG mapping

Coordinates are preferred because they remain stable when DMR display numbers change.

```{r build-standard-probe-mapping}
mapping_work <- data.table(
  CpG = as.character(
    dmr_probe_mapping_raw[[
      mapping_cpg_column
    ]]
  )
)

if (!is.null(mapping_coordinate_column)) {
  mapping_work[
    ,
    coordinate_key := normalise_coordinate(
      dmr_probe_mapping_raw[[
        mapping_coordinate_column
      ]]
    )
  ]

  top20_probe_mapping <- merge(
    mapping_work,
    top20_dmrs,
    by = "coordinate_key",
    all.x = TRUE,
    sort = FALSE
  )
} else {
  mapping_work[
    ,
    Current_DMR_ID := as.character(
      dmr_probe_mapping_raw[[
        mapping_id_column
      ]]
    )
  ]

  top20_probe_mapping <- merge(
    mapping_work,
    top20_dmrs,
    by = "Current_DMR_ID",
    all.x = TRUE,
    sort = FALSE
  )
}
```

The resulting table assigns every actual DMRcate probe to one selected Top 20 DMR.

# 19. Clean and validate the DMR–CpG mapping

Blank probes, unmatched DMRs, and duplicated DMR–CpG pairs are checked before analysis.

```{r clean-and-check-probe-mapping}
top20_probe_mapping <- top20_probe_mapping[
  !is.na(CpG) &
    CpG != ""
]

if (anyNA(
  top20_probe_mapping$Current_DMR_ID
)) {
  print(
    top20_probe_mapping[
      is.na(Current_DMR_ID)
    ]
  )

  stop(
    "Some mapping rows could not be linked to the Top 20 DMR table."
  )
}

top20_probe_mapping <- unique(
  top20_probe_mapping[
    ,
    .(
      DMR_order,
      priority_rank,
      Current_DMR_ID,
      coordinate_key,
      CpG
    )
  ]
)

setorder(
  top20_probe_mapping,
  DMR_order,
  CpG
)

if (uniqueN(
  top20_probe_mapping$Current_DMR_ID
) != 20) {
  stop(
    "The probe mapping does not cover all 20 DMRs."
  )
}
```

Only unique and valid DMR–CpG pairs are retained.

# 20. Confirm that mapped probes belong to `common_cpg`

The actual DMRcate probes should all be present in the common CpG universe used during discovery.

```{r check-mapped-probes-against-common-cpgs}
mapped_cpg <- unique(
  top20_probe_mapping$CpG
)

missing_from_common_cpg <- setdiff(
  mapped_cpg,
  common_cpg
)

probe_common_cpg_check <- data.table(
  Total_unique_mapped_CpGs =
    length(mapped_cpg),
  CpGs_present_in_common_set =
    length(
      intersect(
        mapped_cpg,
        common_cpg
      )
    ),
  CpGs_missing_from_common_set =
    length(
      missing_from_common_cpg
    )
)

print(
  probe_common_cpg_check
)

if (length(
  missing_from_common_cpg
) > 0) {
  print(
    missing_from_common_cpg
  )

  stop(
    "Some actual DMRcate probes are absent from common_cpg."
  )
}
```

The expected number of missing probes is zero.

# 21. Summarise Top 20 DMR probe coverage

The number of actual constituent CpGs is counted for every selected DMR.

```{r summarise-top20-probe-coverage}
top20_probe_mapping_summary <-
  top20_probe_mapping[
    ,
    .(
      n_actual_DMRcate_CpGs =
        uniqueN(CpG)
    ),
    by = .(
      DMR_order,
      priority_rank,
      Current_DMR_ID,
      coordinate_key
    )
  ]

setorder(
  top20_probe_mapping_summary,
  DMR_order
)

print(
  top20_probe_mapping_summary
)

if (any(
  top20_probe_mapping_summary$
    n_actual_DMRcate_CpGs < 2
)) {
  stop(
    "At least one DMR contains fewer than two mapped CpGs."
  )
}
```

At least two CpGs are required to calculate a leave-one-CpG-out DMR mean.

# 22. Build discovery dataset lists

The three beta matrices and phenotype tables are grouped into named lists for consistent processing.

```{r build-discovery-data-lists}
discovery_beta_list <- list(
  GSE199057 = g1,
  GSE149282 = g2,
  GSE119526 = g3
)

discovery_pd_list <- list(
  GSE199057 = pd1,
  GSE149282 = pd2,
  GSE119526 = pd3
)

stopifnot(
  identical(
    names(discovery_beta_list),
    names(discovery_pd_list)
  )
)
```

Each discovery cohort will be analysed independently rather than combining samples across datasets.

# 23. Define safe calculation functions

These helpers return `NA` when a correlation or summary statistic cannot be calculated reliably.

```{r define-safe-calculation-functions}
safe_cor <- function(
    x,
    y,
    method = "spearman") {

  keep <- complete.cases(
    x,
    y
  )

  if (sum(keep) < 3) {
    return(NA_real_)
  }

  x_keep <- x[keep]
  y_keep <- y[keep]

  sd_x <- stats::sd(x_keep)
  sd_y <- stats::sd(y_keep)

  if (
    !is.finite(sd_x) ||
    !is.finite(sd_y) ||
    sd_x == 0 ||
    sd_y == 0
  ) {
    return(NA_real_)
  }

  stats::cor(
    x_keep,
    y_keep,
    method = method
  )
}

safe_mean <- function(x) {
  x <- x[is.finite(x)]

  if (length(x) == 0) {
    return(NA_real_)
  }

  mean(x)
}
```

Undefined correlations do not interrupt the complete analysis.

```{r define-safe-summary-functions}
safe_summary <- function(
    x,
    function_name) {

  x <- x[is.finite(x)]

  if (length(x) == 0) {
    return(NA_real_)
  }

  switch(
    function_name,
    median = stats::median(x),
    mean = base::mean(x),
    minimum = base::min(x),
    maximum = base::max(x),
    sd = if (
      length(x) >= 2
    ) stats::sd(x) else NA_real_,
    stop(
      "Unknown summary function."
    )
  )
}

safe_all_positive <- function(x) {
  x <- x[is.finite(x)]

  if (length(x) == 0) {
    return(NA)
  }

  all(x > 0)
}
```

These functions are also used to build cross-dataset consensus statistics.

# 24. Define the dataset-specific probe analysis

For each DMR, this function calculates the full DMR mean, the leave-one-CpG-out mean, correlation statistics, beta differences, and sample-level supporting values.

```{r define-dataset-probe-analysis-function}
calculate_dataset_probe_results <- function(
    beta,
    pd,
    dataset_name,
    top20_dmrs,
    top20_probe_mapping) {

  probe_result_list <- list()
  sample_result_list <- list()
  dmr_mean_result_list <- list()

  probe_index <- 1
  sample_index <- 1
  dmr_index <- 1

  for (i in seq_len(nrow(top20_dmrs))) {

    current_dmr <- top20_dmrs$Current_DMR_ID[i]

    current_cpgs <- top20_probe_mapping[
      Current_DMR_ID == current_dmr,
      CpG
    ]

    current_cpgs <- intersect(
      current_cpgs,
      rownames(beta)
    )

    if (length(current_cpgs) < 2) {
      stop(
        dataset_name,
        ": fewer than two CpGs were available for ",
        current_dmr,
        "."
      )
    }

    beta_dmr <- beta[
      current_cpgs,
      ,
      drop = FALSE
    ]

    dmr_mean_full <- colMeans(
      beta_dmr,
      na.rm = TRUE
    )

    dmr_mean_full[
      is.nan(dmr_mean_full)
    ] <- NA_real_

    dmr_mean_result_list[[dmr_index]] <-
      data.table(
        Dataset = dataset_name,
        DMR_order = top20_dmrs$DMR_order[i],
        priority_rank = top20_dmrs$priority_rank[i],
        Current_DMR_ID = current_dmr,
        coordinate_key = top20_dmrs$coordinate_key[i],
        Sample = colnames(beta_dmr),
        Group = as.character(pd$group),
        DMR_mean_beta = dmr_mean_full
      )

    dmr_index <- dmr_index + 1

    for (j in seq_along(current_cpgs)) {

      current_cpg <- current_cpgs[j]

      probe_beta <- as.numeric(
        beta_dmr[j, ]
      )

      dmr_mean_leave_one_out <- colMeans(
        beta_dmr[
          -j,
          ,
          drop = FALSE
        ],
        na.rm = TRUE
      )

      dmr_mean_leave_one_out[
        is.nan(dmr_mean_leave_one_out)
      ] <- NA_real_

      normal_index <- pd$group == "Normal"
      tumor_index <- pd$group == "Tumor"

      mean_beta_normal <- safe_mean(
        probe_beta[normal_index]
      )

      mean_beta_tumor <- safe_mean(
        probe_beta[tumor_index]
      )

      dmr_mean_normal <- safe_mean(
        dmr_mean_full[normal_index]
      )

      dmr_mean_tumor <- safe_mean(
        dmr_mean_full[tumor_index]
      )

      cpg_delta_beta <-
        mean_beta_tumor -
        mean_beta_normal

      dmr_delta_beta <-
        dmr_mean_tumor -
        dmr_mean_normal

      delta_direction_consistent <- if (
        is.finite(cpg_delta_beta) &&
        is.finite(dmr_delta_beta)
      ) {
        sign(cpg_delta_beta) ==
          sign(dmr_delta_beta)
      } else {
        NA
      }

      probe_result_list[[probe_index]] <-
        data.table(
          Dataset = dataset_name,
          DMR_order = top20_dmrs$DMR_order[i],
          priority_rank = top20_dmrs$priority_rank[i],
          Current_DMR_ID = current_dmr,
          coordinate_key = top20_dmrs$coordinate_key[i],
          CpG = current_cpg,
          nCpG_in_DMR = length(current_cpgs),
          n_complete_all = sum(
            complete.cases(
              probe_beta,
              dmr_mean_leave_one_out
            )
          ),
          Spearman_full_DMR_mean = safe_cor(
            probe_beta,
            dmr_mean_full,
            method = "spearman"
          ),
          Pearson_full_DMR_mean = safe_cor(
            probe_beta,
            dmr_mean_full,
            method = "pearson"
          ),
          Spearman_leave_one_out = safe_cor(
            probe_beta,
            dmr_mean_leave_one_out,
            method = "spearman"
          ),
          Pearson_leave_one_out = safe_cor(
            probe_beta,
            dmr_mean_leave_one_out,
            method = "pearson"
          ),
          Spearman_leave_one_out_Normal = safe_cor(
            probe_beta[normal_index],
            dmr_mean_leave_one_out[normal_index],
            method = "spearman"
          ),
          Spearman_leave_one_out_Tumor = safe_cor(
            probe_beta[tumor_index],
            dmr_mean_leave_one_out[tumor_index],
            method = "spearman"
          ),
          Mean_absolute_difference_full_mean =
            safe_mean(
              abs(
                probe_beta -
                  dmr_mean_full
              )
            ),
          Mean_beta_Normal = mean_beta_normal,
          Mean_beta_Tumor = mean_beta_tumor,
          CpG_deltaBeta = cpg_delta_beta,
          DMR_mean_Normal = dmr_mean_normal,
          DMR_mean_Tumor = dmr_mean_tumor,
          DMR_deltaBeta = dmr_delta_beta,
          Delta_direction_consistent =
            delta_direction_consistent
        )

      sample_result_list[[sample_index]] <-
        data.table(
          Dataset = dataset_name,
          DMR_order = top20_dmrs$DMR_order[i],
          priority_rank = top20_dmrs$priority_rank[i],
          Current_DMR_ID = current_dmr,
          coordinate_key = top20_dmrs$coordinate_key[i],
          CpG = current_cpg,
          Sample = colnames(beta_dmr),
          Group = as.character(pd$group),
          Probe_beta = probe_beta,
          DMR_mean_full = dmr_mean_full,
          DMR_mean_leave_one_out =
            dmr_mean_leave_one_out
        )

      probe_index <- probe_index + 1
      sample_index <- sample_index + 1
    }
  }

  list(
    probe_results = rbindlist(
      probe_result_list,
      use.names = TRUE,
      fill = TRUE
    ),
    sample_results = rbindlist(
      sample_result_list,
      use.names = TRUE,
      fill = TRUE
    ),
    dmr_mean_results = rbindlist(
      dmr_mean_result_list,
      use.names = TRUE,
      fill = TRUE
    )
  )
}
```

The function returns one probe-level table, one sample-level table, and one DMR-mean table for each dataset.

# 25. Run the analysis in all three discovery datasets

The function is applied independently to each discovery cohort.

```{r run-dataset-specific-analyses}
dataset_analysis_list <- lapply(
  names(discovery_beta_list),
  function(dataset_name) {
    calculate_dataset_probe_results(
      beta =
        discovery_beta_list[[
          dataset_name
        ]],
      pd =
        discovery_pd_list[[
          dataset_name
        ]],
      dataset_name =
        dataset_name,
      top20_dmrs =
        top20_dmrs,
      top20_probe_mapping =
        top20_probe_mapping
    )
  }
)

names(dataset_analysis_list) <-
  names(discovery_beta_list)
```

Samples are not directly combined across cohorts, preventing dataset-level differences from driving the correlations.

# 26. Combine dataset-specific results

The three sets of probe statistics and sample-level values are combined into long-format tables.

```{r combine-dataset-specific-results}
dataset_probe_results <- rbindlist(
  lapply(
    dataset_analysis_list,
    function(x) {
      x$probe_results
    }
  ),
  use.names = TRUE,
  fill = TRUE
)

sample_level_probe_results <- rbindlist(
  lapply(
    dataset_analysis_list,
    function(x) {
      x$sample_results
    }
  ),
  use.names = TRUE,
  fill = TRUE
)

dmr_sample_mean_beta <- rbindlist(
  lapply(
    dataset_analysis_list,
    function(x) {
      x$dmr_mean_results
    }
  ),
  use.names = TRUE,
  fill = TRUE
)

print(
  dim(
    dataset_probe_results
  )
)
```

`dataset_probe_results` contains one row per DMR–CpG–dataset combination.

# 27. Rank probes within each dataset

Dataset-specific ranks describe how each CpG performs relative to the other probes in the same DMR and cohort.

```{r rank-probes-within-datasets}
dataset_probe_results <-
  dataset_probe_results[
    order(
      Dataset,
      DMR_order,
      -Spearman_leave_one_out,
      -Pearson_leave_one_out,
      Mean_absolute_difference_full_mean,
      CpG,
      na.last = TRUE
    )
  ]

dataset_probe_results[
  ,
  Dataset_representative_rank :=
    seq_len(.N),
  by = .(
    Dataset,
    Current_DMR_ID
  )
]

print(
  dataset_probe_results[
    Dataset_representative_rank <= 3
  ]
)
```

These ranks are supporting information; final selection is based on consistency across all three cohorts.

# 28. Build the cross-dataset consensus table

For every DMR–CpG pair, correlations and ranks are summarised across the three discovery datasets.

```{r calculate-consensus-statistics}
consensus_probe_results <-
  dataset_probe_results[
    ,
    .(
      n_datasets_available = sum(
        is.finite(
          Spearman_leave_one_out
        )
      ),
      Median_Spearman_leave_one_out =
        safe_summary(
          Spearman_leave_one_out,
          "median"
        ),
      Mean_Spearman_leave_one_out =
        safe_summary(
          Spearman_leave_one_out,
          "mean"
        ),
      Minimum_Spearman_leave_one_out =
        safe_summary(
          Spearman_leave_one_out,
          "minimum"
        ),
      Maximum_Spearman_leave_one_out =
        safe_summary(
          Spearman_leave_one_out,
          "maximum"
        ),
      SD_Spearman_leave_one_out =
        safe_summary(
          Spearman_leave_one_out,
          "sd"
        ),
      Median_Pearson_leave_one_out =
        safe_summary(
          Pearson_leave_one_out,
          "median"
        ),
      Median_Spearman_full_DMR_mean =
        safe_summary(
          Spearman_full_DMR_mean,
          "median"
        ),
      Median_absolute_difference_full_mean =
        safe_summary(
          Mean_absolute_difference_full_mean,
          "median"
        ),
      Median_CpG_deltaBeta =
        safe_summary(
          CpG_deltaBeta,
          "median"
        ),
      Median_DMR_deltaBeta =
        safe_summary(
          DMR_deltaBeta,
          "median"
        ),
      All_correlations_positive =
        safe_all_positive(
          Spearman_leave_one_out
        ),
      Delta_direction_consistent_datasets =
        sum(
          Delta_direction_consistent,
          na.rm = TRUE
        ),
      Median_dataset_rank =
        safe_summary(
          Dataset_representative_rank,
          "median"
        ),
      Mean_dataset_rank =
        safe_summary(
          Dataset_representative_rank,
          "mean"
        ),
      Top1_in_n_datasets =
        sum(
          Dataset_representative_rank == 1,
          na.rm = TRUE
        )
    ),
    by = .(
      DMR_order,
      priority_rank,
      Current_DMR_ID,
      coordinate_key,
      CpG,
      nCpG_in_DMR
    )
  ]
```

The median correlation represents typical performance, while the minimum and standard deviation describe cross-dataset consistency.

# 29. Rank probes by cross-dataset consensus

The ranking prioritises complete, consistently positive, strongly correlated, and stable probes.

```{r rank-probes-by-consensus}
consensus_probe_results[
  ,
  All_correlations_positive_score :=
    as.integer(
      All_correlations_positive
    )
]

consensus_probe_results <-
  consensus_probe_results[
    order(
      DMR_order,
      -n_datasets_available,
      -All_correlations_positive_score,
      -Median_Spearman_leave_one_out,
      -Minimum_Spearman_leave_one_out,
      SD_Spearman_leave_one_out,
      Median_absolute_difference_full_mean,
      CpG,
      na.last = TRUE
    )
  ]

consensus_probe_results[
  ,
  Consensus_representative_rank :=
    seq_len(.N),
  by = Current_DMR_ID
]
```

Rank 1 identifies the CpG that most consistently follows its DMR methylation pattern across discovery cohorts.

# 30. Select primary and backup probes

One primary representative CpG and up to two backup CpGs are retained for each Top 20 DMR.

```{r select-primary-and-backup-probes}
selected_probes <- copy(
  consensus_probe_results[
    Consensus_representative_rank == 1
  ]
)

top3_representative_probes <- copy(
  consensus_probe_results[
    Consensus_representative_rank <= 3
  ]
)

setorder(
  selected_probes,
  DMR_order
)

setorder(
  top3_representative_probes,
  DMR_order,
  Consensus_representative_rank
)

if (nrow(selected_probes) != 20) {
  stop(
    "The final rank-1 table does not contain exactly 20 probes."
  )
}

print(
  selected_probes[
    ,
    .(
      priority_rank,
      Current_DMR_ID,
      coordinate_key,
      CpG,
      Median_Spearman_leave_one_out,
      Minimum_Spearman_leave_one_out,
      SD_Spearman_leave_one_out,
      All_correlations_positive,
      Median_dataset_rank,
      Top1_in_n_datasets,
      Median_CpG_deltaBeta,
      Median_DMR_deltaBeta
    )
  ]
)
```

The Top 3 table provides alternatives when the highest-ranked CpG is unsuitable for assay design.

# 31. Build the selected-probe dataset check

This table shows the selected CpG's performance separately in each discovery dataset.

```{r build-selected-probe-dataset-check}
selected_probe_dataset_check <- merge(
  dataset_probe_results,
  selected_probes[
    ,
    .(
      Current_DMR_ID,
      CpG,
      Consensus_representative_rank
    )
  ],
  by = c(
    "Current_DMR_ID",
    "CpG"
  ),
  all = FALSE
)

setorder(
  selected_probe_dataset_check,
  DMR_order,
  Dataset
)

print(
  selected_probe_dataset_check[
    ,
    .(
      Dataset,
      priority_rank,
      Current_DMR_ID,
      CpG,
      Spearman_leave_one_out,
      Pearson_leave_one_out,
      Spearman_full_DMR_mean,
      Spearman_leave_one_out_Normal,
      Spearman_leave_one_out_Tumor,
      CpG_deltaBeta,
      DMR_deltaBeta,
      Delta_direction_consistent,
      Dataset_representative_rank
    )
  ]
)
```

This table should be reviewed before a probe is accepted based only on its consensus summary.

# 32. Prepare sample-level data for selected probes

Only the rank-1 CpG from each DMR is retained for final scatterplots.

```{r prepare-selected-probe-sample-data}
selected_probe_sample_results <- merge(
  sample_level_probe_results,
  selected_probes[
    ,
    .(
      Current_DMR_ID,
      CpG,
      Consensus_representative_rank
    )
  ],
  by = c(
    "Current_DMR_ID",
    "CpG"
  ),
  all = FALSE
)

selected_probe_sample_results[
  ,
  DMR_label := paste0(
    "Rank ",
    priority_rank,
    ": ",
    Current_DMR_ID,
    "\n",
    coordinate_key,
    "\n",
    CpG
  )
]

print(
  dim(
    selected_probe_sample_results
  )
)
```

Each selected DMR contributes one CpG profile per sample in each discovery dataset.

# 33. Build the consensus-correlation plot

The selected CpG's median correlation is shown together with its minimum-to-maximum range across datasets.

```{r prepare-consensus-correlation-plot}
selected_probes[
  ,
  DMR_label := paste0(
    "Rank ",
    priority_rank,
    ": ",
    Current_DMR_ID,
    "\n",
    CpG
  )
]

selected_probes[
  ,
  DMR_label := factor(
    DMR_label,
    levels = rev(
      DMR_label
    )
  )
]
```

The factor order places the highest-priority DMR at the top.

```{r build-consensus-correlation-plot}
p_consensus_correlation <- ggplot(
  selected_probes,
  aes(
    y = DMR_label
  )
) +
  geom_segment(
    aes(
      x = Minimum_Spearman_leave_one_out,
      xend = Maximum_Spearman_leave_one_out,
      yend = DMR_label
    ),
    linewidth = 0.7
  ) +
  geom_point(
    aes(
      x = Median_Spearman_leave_one_out
    ),
    size = 3
  ) +
  geom_text(
    aes(
      x = Median_Spearman_leave_one_out,
      label = sprintf(
        "%.3f",
        Median_Spearman_leave_one_out
      )
    ),
    hjust = -0.25,
    size = 3
  ) +
  coord_cartesian(
    xlim = c(
      min(
        0,
        selected_probes$
          Minimum_Spearman_leave_one_out,
        na.rm = TRUE
      ),
      1.08
    )
  ) +
  labs(
    title = "Representative CpG selected for each Top 20 DMR",
    subtitle = "Point: median leave-one-out Spearman correlation; line: dataset range",
    x = "Spearman correlation",
    y = "DMR and selected CpG"
  ) +
  theme_bw(
    base_size = 12
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0.5
    ),
    panel.grid.major.y =
      element_blank()
  )

print(
  p_consensus_correlation
)
```

A high median and a narrow, consistently positive range support stable probe representativeness.

# 34. Build the selected-probe heatmap

The dataset-specific leave-one-out correlations are displayed for every selected probe.

```{r prepare-selected-probe-heatmap}
selected_probe_heatmap_df <-
  selected_probe_dataset_check[
    ,
    .(
      Dataset,
      priority_rank,
      Current_DMR_ID,
      CpG,
      Spearman_leave_one_out
    )
  ]

selected_probe_heatmap_df[
  ,
  DMR_label := paste0(
    "Rank ",
    priority_rank,
    ": ",
    Current_DMR_ID,
    "\n",
    CpG
  )
]

selected_probe_heatmap_df[
  ,
  DMR_label := factor(
    DMR_label,
    levels = rev(
      as.character(
        selected_probes$DMR_label
      )
    )
  )
]
```

Each row represents one selected CpG and each column represents one discovery dataset.

```{r build-selected-probe-heatmap}
p_selected_probe_heatmap <- ggplot(
  selected_probe_heatmap_df,
  aes(
    x = Dataset,
    y = DMR_label,
    fill = Spearman_leave_one_out
  )
) +
  geom_tile(
    color = "white",
    linewidth = 0.6
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.2f",
        Spearman_leave_one_out
      )
    ),
    size = 3
  ) +
  scale_fill_gradient2(
    midpoint = 0,
    limits = c(-1, 1),
    na.value = "grey85"
  ) +
  labs(
    title = "Dataset-specific agreement of selected CpGs",
    subtitle = "Leave-one-CpG-out Spearman correlation",
    x = "Discovery dataset",
    y = "DMR and selected CpG",
    fill = "Spearman\ncorrelation"
  ) +
  theme_bw(
    base_size = 12
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0.5
    ),
    panel.grid = element_blank()
  )

print(
  p_selected_probe_heatmap
)
```

Consistently high values across all three datasets indicate robust cross-cohort representativeness.

# 35. Define the selected-probe scatterplot function

One figure is generated per DMR, with separate panels for the three discovery datasets.

```{r define-selected-probe-scatterplot-function}
build_selected_probe_scatterplot <- function(
    plot_data,
    dmr_id) {

  current_data <- plot_data[
    Current_DMR_ID == dmr_id
  ]

  selected_cpg <- unique(
    current_data$CpG
  )

  ggplot(
    current_data,
    aes(
      x = DMR_mean_leave_one_out,
      y = Probe_beta,
      color = Group
    )
  ) +
    geom_point(
      alpha = 0.70,
      size = 1.8
    ) +
    geom_smooth(
      method = "lm",
      formula = y ~ x,
      se = FALSE,
      linewidth = 0.7
    ) +
    facet_wrap(
      ~ Dataset,
      scales = "free"
    ) +
    labs(
      title = paste0(
        dmr_id,
        ": ",
        selected_cpg
      ),
      subtitle = "Selected CpG versus the mean beta of the remaining CpGs",
      x = "Leave-one-CpG-out DMR mean beta",
      y = "Selected CpG beta value",
      color = "Group"
    ) +
    theme_bw(
      base_size = 12
    ) +
    theme(
      plot.title = element_text(
        hjust = 0.5,
        face = "bold"
      ),
      plot.subtitle = element_text(
        hjust = 0.5
      ),
      legend.position = "top"
    )
}
```

A tight increasing relationship in all three panels supports selection of the CpG as a region-level representative.

# 36. Save one scatterplot per Top 20 DMR

The twenty DMR-specific scatterplots are written to a dedicated subdirectory.

```{r save-selected-probe-scatterplots}
for (current_dmr in
     selected_probes$Current_DMR_ID) {

  current_plot <-
    build_selected_probe_scatterplot(
      selected_probe_sample_results,
      current_dmr
    )

  safe_dmr_name <- gsub(
    "[^A-Za-z0-9_-]",
    "_",
    current_dmr
  )

  ggsave(
    file.path(
      scatter_plot_dir,
      paste0(
        safe_dmr_name,
        "_selected_probe_scatter.png"
      )
    ),
    plot = current_plot,
    width = 10,
    height = 4,
    dpi = 300
  )

  ggsave(
    file.path(
      scatter_plot_dir,
      paste0(
        safe_dmr_name,
        "_selected_probe_scatter.pdf"
      )
    ),
    plot = current_plot,
    width = 10,
    height = 4
  )
}
```

These plots allow the final selected probe to be inspected separately in each discovery cohort.

# 37. Export input and mapping audits

The standardised Top 20 table, actual DMRcate probe mapping, and coverage checks are saved.

```{r export-input-and-mapping-audits}
fwrite(
  discovery_dataset_summary,
  file.path(
    outdir,
    "discovery_dataset_summary.csv"
  )
)

fwrite(
  top20_dmrs,
  file.path(
    outdir,
    "Top20_DMRs_standardised.csv"
  )
)

fwrite(
  top20_probe_mapping,
  file.path(
    outdir,
    "Top20_DMR_actual_DMRcate_probe_mapping_standardised.csv"
  )
)

fwrite(
  top20_probe_mapping_summary,
  file.path(
    outdir,
    "Top20_DMR_actual_DMRcate_probe_count_summary.csv"
  )
)

fwrite(
  probe_common_cpg_check,
  file.path(
    outdir,
    "Top20_DMR_probe_common_CpG_check.csv"
  )
)
```

These files document exactly which DMRs and CpGs entered the analysis.

# 38. Export probe-prioritisation results

The full dataset-level calculations and final consensus rankings are saved below.

```{r export-probe-prioritisation-results}
fwrite(
  dataset_probe_results,
  file.path(
    outdir,
    "Top20_DMR_probe_representativeness_by_discovery_dataset.csv"
  )
)

fwrite(
  consensus_probe_results,
  file.path(
    outdir,
    "Top20_DMR_probe_consensus_ranking_across_discovery_datasets.csv"
  )
)

fwrite(
  selected_probes,
  file.path(
    outdir,
    "Top20_DMR_representative_probe_rank1_across_discovery_datasets.csv"
  )
)

fwrite(
  top3_representative_probes,
  file.path(
    outdir,
    "Top20_DMR_representative_probe_top3_across_discovery_datasets.csv"
  )
)

fwrite(
  selected_probe_dataset_check,
  file.path(
    outdir,
    "Top20_DMR_selected_probe_dataset_specific_check.csv"
  )
)
```

The rank-1 file is the main statistical result, while the Top 3 file supplies assay-design alternatives.

# 39. Export sample-level supporting values

The exact sample-level values used in the correlation analysis are retained for reproducibility.

```{r export-sample-level-results}
fwrite(
  dmr_sample_mean_beta,
  file.path(
    outdir,
    "Top20_DMR_sample_level_mean_beta_by_discovery_dataset.csv"
  )
)

fwrite(
  sample_level_probe_results,
  file.path(
    outdir,
    "Top20_DMR_all_probe_sample_level_values.csv"
  )
)

fwrite(
  selected_probe_sample_results,
  file.path(
    outdir,
    "Top20_DMR_selected_probe_sample_level_values.csv"
  )
)
```

These tables allow the correlations and scatterplots to be reconstructed without repeating the complete analysis.

# 40. Save summary plots

The consensus plot and heatmap are saved in both PNG and PDF formats.

```{r save-consensus-correlation-plot}
ggsave(
  file.path(
    outdir,
    "Top20_DMR_selected_probe_consensus_correlations.png"
  ),
  plot = p_consensus_correlation,
  width = 11,
  height = 9,
  dpi = 300
)

ggsave(
  file.path(
    outdir,
    "Top20_DMR_selected_probe_consensus_correlations.pdf"
  ),
  plot = p_consensus_correlation,
  width = 11,
  height = 9
)
```

The consensus plot summarises the magnitude and range of each selected probe's correlations.

```{r save-selected-probe-heatmap}
ggsave(
  file.path(
    outdir,
    "Top20_DMR_selected_probe_dataset_correlation_heatmap.png"
  ),
  plot = p_selected_probe_heatmap,
  width = 9,
  height = 10,
  dpi = 300
)

ggsave(
  file.path(
    outdir,
    "Top20_DMR_selected_probe_dataset_correlation_heatmap.pdf"
  ),
  plot = p_selected_probe_heatmap,
  width = 9,
  height = 10
)
```

The heatmap highlights probes whose representativeness is consistent across all three discovery datasets.

# 41. Save analysis objects

All principal inputs, calculations, selections, and plots are saved for later review.

```{r save-analysis-objects}
save(
  g1,
  g2,
  g3,
  pd1,
  pd2,
  pd3,
  common_cpg,
  top20_dmrs_raw,
  dmr_probe_mapping_raw,
  probe_audit_raw,
  top20_dmrs,
  top20_probe_mapping,
  top20_probe_mapping_summary,
  discovery_dataset_summary,
  probe_common_cpg_check,
  dataset_probe_results,
  sample_level_probe_results,
  dmr_sample_mean_beta,
  consensus_probe_results,
  selected_probes,
  top3_representative_probes,
  selected_probe_dataset_check,
  selected_probe_sample_results,
  p_consensus_correlation,
  p_selected_probe_heatmap,
  file = file.path(
    outdir,
    "step7_Top20_probe_prioritisation_across_discovery_datasets.Rdata"
  )
)
```

The saved RData allows the final tables and summary plots to be reopened without repeating the full workflow.

# 42. Session information

The R version and package dependencies are listed at the end of the report.

```{r session-information}
sessionInfo()
```
