---
title: "step2.independent validation in 5 450k datasets"
author: "YC"
date: "2026-06-25"
output:
 html_document:
  toc: true
  toc_float: true
  code_folding: hide
  fig_width: 7
  fig_height: 7
  theme: cosmo
---

#⭐ Overview

This document preprocesses five independent GEO 450K methylation datasets and then validates discovery hyper-DMRs across those datasets. 

The setup chunk defines the default display behaviour for all subsequent R code chunks.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  results = "hide",
  message = FALSE,
  warning = FALSE)
```

##🍁🍁 step0.data preprocessing
In this step, we performed three main tasks for each dataset:
1.Created a standardized beta-value matrix.
2.Constructed a Normal/Tumor phenotype table and aligned the phenotype information with the corresponding samples in the beta-value matrix.
3.Applied champ.filter() for quality control and probe filtering, and saved the resulting myLoad object, which contains the filtered beta-value matrix and its corresponding phenotype table.

### dataset1: GSE42752 preprocessing

For GSE42752, we performed the following steps:
1. Extracted the processed beta-value matrix directly from the GEO `ExpressionSet` using `exprs(eset[[1]])`.
2. Obtained phenotype information from the GEO metadata and classified the samples as Normal or Tumor.
3. Checked the sample IDs and aligned the phenotype table with the beta-value matrix.
4. Applied `champ.filter()` for quality control and probe filtering.
5. Saved the resulting `myLoad` object for subsequent validation analysis.


#### Clears the current R environment and loads the packages required for GSE42752.
```{r gse42752_setup}
rm(list = ls())
options(timeout = 100000) 
options(scipen = 20)
options(stringsAsFactors = F)

require(GEOquery)
require(Biobase)
library("impute")
library(ChAMP)
```
The required packages for GEO data retrieval and ChAMP filtering were loaded.

#### Download GEO series matrix metadata & Read GEO metadata
```{r gse42752_read_geo}
eset <- getGEO("GSE42752",destdir = './',AnnotGPL = T,getGPL = F)
beta.m <- exprs(eset[[1]])
class(beta.m)
dim(beta.m)
```
`class(beta.m)` and `dim(beta.m)` were used to check the data type and the number of probes and samples. The results were compared with the GEO dataset page to confirm that the data were imported correctly.

#### Extracts the relevant phenotype columns, assigns samples to Normal or Tumor.
```{r gse42752_build_pd}

pD.all <- pData(eset[[1]])
pD <- pD.all[,c(37,2)]
names(pD) <- c("group", "ID")
pD$group <- ifelse(grepl("normal",pD$group,ignore.case = T),
  "Normal","Tumor")

#check if the extracted groupings are correct
check_df <- data.frame(Title = pD.all$`tissue:ch1`, Group = pD$group)
head(check_df, 10)
tail(check_df,10)
table(pD$group)
```
The sample groups were checked to confirm correct classification.

#### Align pD with beta.m
```{r gse42752_align_samples}
## Check whether beta.m and pD contain the same sample set;setequal() ignores order.
identical(rownames(pD),colnames(beta.m))
setequal(colnames(beta.m), pD$ID)
# Confirm that there are no duplicated sample names
sum(duplicated(colnames(beta.m)))
sum(duplicated(pD$ID))
# Set pD row names as GSM IDs.
rownames(pD) <- pD$ID
# Set group order
pD$group <- factor(pD$group,levels = c("Normal", "Tumor"))
# Sort pD by group
sample_order <- order(pD$group)
pD <- pD[sample_order, , drop = FALSE]
# Reorder beta matrix columns according to pD
beta.m <- beta.m[, rownames(pD), drop = FALSE]
## Final strict alignment check.
identical(rownames(pD),colnames(beta.m))##This must be TRUE before champ.filter().
dim(beta.m)
dim(pD)
```
The phenotype table and beta matrix have to be aligned before `champ.filter()` was applied.

#### performs ChAMP probe filtering and saves the resulting `myLoad` object as an `.Rdata` file.

```{r gse42752_champ_filter}
## Run ChAMP filtering using the cleaned beta matrix and aligned pD.
myLoad <- champ.filter(beta = beta.m, pd = pD, arraytype = "450k")
## Save the ChAMP object for downstream analysis.
save(myLoad, file = "step0.1-GSE42752_output.Rdata")
#saveRDS(myLoad, file = "step0.1-GSE42752_output.rds")
#myLoad <- readRDS("step0.1-GSE42752_output.rds")
# For all downstream analyses, simply use this myLoad variable
```
champ.filter() performs initial quality control:
(1).Filters out probes with low detection p-value.
(2).Removes probes with missing values.
(3).Filters sex chromosomes or SNP-associated probes if needed.
(4).The output myLoad is an object used for all downstream ChAMP analyses (normalization, DMP, DMR, pathway analysis).

`myLoad` contains the filtered beta-value matrix and its corresponding phenotype table for subsequent analysis.

### dataset2: GSE48684

For GSE48684, we performed the following steps:
1. Extracted the processed beta-value matrix directly from the GEO `ExpressionSet`.
2. Retained the `normal-C`, `normal-H`, and `cancer` samples and classified them as Normal or Tumor.
3. Checked and aligned the phenotype table with the beta-value matrix.
4. Applied `champ.filter()` and saved the resulting `myLoad` object.

#### Clear environment and load required packages
```{r gse48684_setup2}

rm(list = ls())
options(timeout = 100000) 
options(scipen = 20)
options(stringsAsFactors = F)

require(GEOquery)
require(Biobase)
library("impute")
library(ChAMP)
```

#### Read GEO data and extract the beta-value matrix
```{r}
eset <- getGEO("GSE48684",destdir = './',AnnotGPL = T,getGPL = F)
beta.m <- exprs(eset[[1]])
class(beta.m)
dim(beta.m)
```


#### Build the phenotype table
```{r}
pD.all <- pData(eset[[1]])
pD <- pD.all[,c(22,2)]
pD_nt <- pD[pD$description %in% c("normal-C", "normal-H", "cancer"), ]
pD_nt1 <- pD_nt
pD_nt1$Group <- ifelse(pD_nt1$description %in% c("normal-C", "normal-H"),
  "Normal",
  "Tumor")
pD_nt1 <- pD_nt1[, c("geo_accession", "Group")]
colnames(pD_nt1)[colnames(pD_nt1) == "geo_accession"] <- "ID"
colnames(pD_nt1)[colnames(pD_nt1) == "Group"] <- "group"
pD <- pD_nt1

table(pD$group)
check_df <- data.frame(Title = pD_nt$description, Group = pD$group)
head(check_df, 10)
tail(check_df,10)
```

#### Align the phenotype table with the beta matrix
```{r}
identical(rownames(pD),colnames(beta.m))
setequal(colnames(beta.m), pD$ID)
sum(duplicated(colnames(beta.m)))
sum(duplicated(pD$ID))

# Set group order
pD$group <- factor(pD$group,levels = c("Normal", "Tumor"))
# Sort pD by group
sample_order <- order(pD$group)
pD <- pD[sample_order, , drop = FALSE]

# Reorder beta matrix columns according to pD
beta.m <- beta.m[, rownames(pD), drop = FALSE]

## Final strict alignment check.
identical(rownames(pD),colnames(beta.m))
dim(pD)
dim(beta.m)
```

#### Run ChAMP filtering and save the result
```{r}
## Run ChAMP filtering using the cleaned beta matrix and aligned pD.
myLoad <- champ.filter(beta = beta.m, pd = pD, arraytype = "450k")
## Save the ChAMP object for downstream analysis.
save(myLoad, file = "step0.2-GSE48684_output.Rdata")

```

### dataset3: GSE77718

For GSE77718, we performed the following steps:
1. Extracted the processed beta-value matrix from the GEO `ExpressionSet`.
2. Used the disease-state metadata to classify samples as Normal or Tumor.
3. Checked and aligned the phenotype table with the beta-value matrix.
4. Applied `champ.filter()` and saved the resulting `myLoad` object.

#### Clear the environment and load the required packages
```{r GSE77718}
rm(list = ls())
options(timeout = 100000) 
options(scipen = 20)
options(stringsAsFactors = F)
require(GEOquery)
require(Biobase)
library("impute")
library(ChAMP)
```
#### Read GEO data and extract the beta-value matrix
```{r}
eset <- getGEO("GSE77718",destdir = './',AnnotGPL = T,getGPL = F)
beta.m <- exprs(eset[[1]])
class(beta.m)
dim(beta.m)
```

#### Build the phenotype table

```{r}
pD.all <- pData(eset[[1]])
pD <- pD.all[,c(33,2)]
names(pD) <- c("group", "ID")
pD$group <- ifelse(grepl("Normal",pD$group,ignore.case = T),
                   "Normal","Tumor")
#check if the extracted groupings are correct
check_df <- data.frame(Title = pD.all$`disease state:ch1`, Group = pD$group)
head(check_df, 10)
tail(check_df,10)
table(pD$group)
```

#### Align the phenotype table with the beta matrix

```{r}
identical(rownames(pD),colnames(beta.m))
setequal(colnames(beta.m), pD$ID)
sum(duplicated(colnames(beta.m)))
sum(duplicated(pD$ID))

# Set group order
pD$group <- factor(pD$group,levels = c("Normal", "Tumor"))

# Sort pD by group
sample_order <- order(pD$group)

pD <- pD[sample_order, , drop = FALSE]

# Reorder beta matrix columns according to pD
beta.m <- beta.m[, rownames(pD), drop = FALSE]

## Final strict alignment check.
identical(rownames(pD),colnames(beta.m))
dim(beta.m)
dim(pD)
```
The final alignment check should return `TRUE`.

#### Run ChAMP filtering and save the result
```{r}
## Run ChAMP filtering using the cleaned beta matrix and aligned pD.
myLoad <- champ.filter(beta = beta.m, pd = pD, arraytype = "450k")
## Save the ChAMP object for downstream analysis.
save(myLoad, file = "step0.3-GSE77718_output.Rdata")

```

### Dataset4: GSE101764

For GSE101764, we performed the following steps:
1. Extracted the processed beta-value matrix from the GEO `ExpressionSet`.
2. Classified mucosa samples as Normal and the remaining selected samples as Tumor.
3. Checked and aligned the phenotype table with the beta-value matrix.
4. Applied `champ.filter()` and saved the resulting `myLoad` object.

#### Module0.Clear environment and load required packages
```{r GSE101764}
rm(list = ls())
options(timeout = 100000) 
options(scipen = 20)
options(stringsAsFactors = F)

require(GEOquery)
require(Biobase)
library("impute")
library(ChAMP)
```

#### Module 1. Read GEO metadata
```{r}
eset <- getGEO("GSE101764",destdir = './',AnnotGPL = T,getGPL = F)
beta.m <- exprs(eset[[1]])
class(beta.m)
dim(beta.m)
```

#### Module 2. Build phenotype table pD

```{r}
pD.all <- pData(eset[[1]])
pD <- pD.all[,c(13,2)]
names(pD) <- c("group", "ID")
pD$group <- ifelse(grepl("mucosa",pD$group,ignore.case = T),
                   "Normal","Tumor")

check_df <- data.frame(Title = pD.all$characteristics_ch1.3, Group = pD$group)
head(check_df, 10)
tail(check_df,10)
table(pD$group)
```

#### Module 3. Align pD with beta.m
```{r}
identical(rownames(pD),colnames(beta.m))
setequal(colnames(beta.m), pD$ID)
sum(duplicated(colnames(beta.m)))
sum(duplicated(pD$ID))

# Set group order
pD$group <- factor(pD$group,levels = c("Normal", "Tumor"))

# Sort pD by group
sample_order <- order(pD$group)

pD <- pD[sample_order, , drop = FALSE]

# Reorder beta matrix columns according to pD
beta.m <- beta.m[, rownames(pD), drop = FALSE]

## Final strict alignment check.
identical(rownames(pD),colnames(beta.m))
dim(beta.m)
dim(pD)
```

#### Module 4. Run ChAMP filtering
```{r}
## Run ChAMP filtering using the cleaned beta matrix and aligned pD.
myLoad <- champ.filter(beta = beta.m, pd = pD, arraytype = "450k")
## Save the ChAMP object for downstream analysis.
save(myLoad, file = "step0.4-GSE101764_output.Rdata")
```

### Dataset 5: GSE131013 preprocessing

GSE131013 required additional processing because its beta-value matrix was provided as a separate normalized matrix file.

For GSE131013, we performed the following steps:

1. Downloaded the GEO metadata.
2. Read the normalized beta-value matrix from `GSE131013_normalized_matrix.txt.gz`.
3. Matched the internal matrix sample IDs to the corresponding GEO accession numbers.
4. Constructed a numeric CpG-by-sample beta-value matrix.
5. Built and aligned the Normal/Tumor phenotype table.
6. Applied `champ.filter()` and saved the resulting `myLoad` object.

#### Clear the environment and load the required packages
```{r gse131013_setup}

rm(list = ls())
options(timeout = 100000) 
options(scipen = 20)
options(stringsAsFactors = F)
require(GEOquery)
require(Biobase)
library("impute")
library(ChAMP)
library(data.table)
```

#### Read GEO metadata
```{r}
eset <- getGEO("GSE131013",destdir = './',AnnotGPL = T,getGPL = F)
#beta.m <- exprs(eset[[1]])
#class(beta.m)
#dim(beta.m)
```
The GEO metadata were available for phenotype construction and sample matching.
#### Build the initial phenotype table
```{r}
pD.all <- pData(eset[[1]])
pD <- pD.all[,c(43,2)]
names(pD) <- c("group", "ID")
pD$group <- ifelse(grepl("Tumor",pD$group,ignore.case = T),
                   "Tumor","Normal")

#check if the extracted groupings are correct
check_df <- data.frame(Title = pD.all$characteristics_ch1, Group = pD$group)
head(check_df, 10)
tail(check_df,10)
table(pD$group)
```

#### Read the normalized beta-value matrix
The normalized matrix file was read, and the beta-value columns were selected while detection p-value columns were excluded.
```{r}
# prepare standard beta value matrix
beta.df <- fread("GSE131013_normalized_matrix.txt.gz", data.table = FALSE)
dim(beta.df)
#View(beta.df)
#extract columns including beta value
beta.cols <- colnames(beta.df)[
  colnames(beta.df) != "ID_REF" &
    !grepl("Detection\\.Pval$", colnames(beta.df))]

length(beta.cols)

## Extract the internal sample ID from the GEO sample title
## Example:
## "Paired Adjacent Normal sample from A2004_N patient" -> "A2004_N"
## "Tumor sample from A2004_T patient" -> "A2004_T"
## "Mucosa sample from A6144_M healthy donor" -> "A6144_M"
pD.all$matrix_id <- sub(
  ".*from[[:space:]]+([^[:space:]]+).*",
  "\\1",pD.all$title)
## Check whether matrix_id extraction worked correctly
head(pD.all[, c("title", "geo_accession", "matrix_id")], 20)
## Check how many extracted matrix IDs are found in beta column names
sum(pD.all$matrix_id %in% beta.cols)

```

The initial matching check identified 238 of 240 sample IDs. We therefore investigated the two unmatched samples and standardised their IDs to ensure complete alignment between the GEO metadata and the beta-value matrix.

#### Extract and check internal sample IDs
Internal sample IDs were extracted from the GEO sample titles and compared with the beta-matrix column names.
```{r}
## Check unmatched sample IDs between GEO metadata and beta matrix
## Sample IDs present in GEO metadata but not directly found in beta matrix
setdiff(pD.all$matrix_id, beta.cols)
## Sample IDs present in beta matrix but not directly found in GEO metadata
setdiff(beta.cols, pD.all$matrix_id)
## Check matching after removing extra "_0" suffix
beta.cols.clean <- sub("_0$", "", beta.cols)

sum(pD.all$matrix_id %in% beta.cols.clean)

setdiff(pD.all$matrix_id, beta.cols.clean)
setdiff(beta.cols.clean, pD.all$matrix_id)
```
Both setdiff() checks returned character(0), confirming that all sample IDs were successfully matched.

#### Build the sample-ID mapping table
A mapping table was created, and an extra `_0` suffix was removed when needed for sample matching.
```{r}
## Match GEO metadata to the order of beta matrix columns
idx <- match(beta.cols.clean, pD.all$matrix_id)
## Confirm that all samples were matched
stopifnot(!anyNA(idx))
## Reorder metadata to match beta matrix columns
pD.aligned <- pD.all[idx, , drop = FALSE]
stopifnot(identical(beta.cols.clean, pD.aligned$matrix_id))
```

#### Build the final sample mapping table
```{r}
## Build a sample mapping table
sample.map <- data.frame(
  matrix_id_original = beta.cols,
  matrix_id_for_match = beta.cols.clean,
  GEO_accession = pD.all$geo_accession[idx],
  title = pD.all$title[idx],
  stringsAsFactors = FALSE
)

## Set original beta matrix sample IDs as row names
rownames(sample.map) <- sample.map$matrix_id_original

## Check whether any samples remain unmatched
stopifnot(!anyNA(sample.map$GEO_accession))

## Inspect the mapping table
head(sample.map, 20)
tail(sample.map, 20)
```

#### Construct the standard beta-value matrix
The beta-value columns were converted to a numeric matrix with CpG probe IDs as rows and GEO accession numbers as columns.
```{r}
## Extract beta values
beta.m <- beta.df[, sample.map$matrix_id_original,drop = FALSE]
## Set row names as CpG probe IDs
rownames(beta.m) <- beta.df$ID_REF
## Convert data frame to numeric matrix
beta.m <- as.matrix(beta.m)
mode(beta.m) <- "numeric"

## Set column names as GEO accession numbers
colnames(beta.m) <- sample.map$GEO_accession

## Check the standard beta matrix
dim(beta.m)
head(rownames(beta.m))
head(colnames(beta.m))
```

#### Align the phenotype table with the beta matrix
```{r}
identical(rownames(pD),colnames(beta.m))
setequal(colnames(beta.m), pD$ID)
sum(duplicated(colnames(beta.m)))
sum(duplicated(pD$ID))

# Set group order
pD$group <- factor(pD$group,levels = c("Normal", "Tumor"))
# Sort pD by group
sample_order <- order(pD$group)
pD <- pD[sample_order, , drop = FALSE]

# Reorder beta matrix columns according to pD
beta.m <- beta.m[, rownames(pD), drop = FALSE]
## Final strict alignment check.
identical(rownames(pD),colnames(beta.m))
dim(beta.m)
dim(pD)
```

#### Run ChAMP filtering and save the result
```{r}
## Run ChAMP filtering using the cleaned beta matrix and aligned pD.
myLoad <- champ.filter(beta = beta.m, pd = pD, arraytype = "450k")
## Save the ChAMP object for downstream analysis.
save(myLoad, file = "step0.5-GSE131013_output.Rdata")
```

##🍁🍁 step1.Independent validation in five 450k datasets

In this step, the discovery hyper-DMRs were tested in the five independent GEO datasets.
The main validation workflow included:

1. Loaded the discovery DMR table and the five preprocessed `myLoad` objects.
2. Standardized the DMR coordinates and mapped 450K CpG probes to each DMR.
3. Prepared and aligned each validation dataset.
4. Calculated the Normal and Tumor mean beta values, delta beta values, p-values, and FDR values.
5. Summarized the validation results across datasets.
6. Saved the full results and exported robustly validated hyper-DMRs-DMRs have passed at least 3 datasete will be retained for background filtering.

#### Clear the environment and load the required packages 
```{r validation_setup}

rm(list = ls())
options(timeout = 100000)
options(scipen = 20)
options(stringsAsFactors = FALSE)

library(data.table)
library(minfi)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
```

#### Set the input files and output directory
```{r}
## Discovery hyper-DMR file from DMRcate meta-analysis
discovery_dmr_file <- "dmr_hyper_dmrcate_meta.csv"

## Validation datasets
## Each .Rdata file should contain an object named myLoad with myLoad$beta and myLoad$pd
dataset_files <- c(
  GSE42752  = "step0.1-GSE42752_output.Rdata",
  GSE48684  = "step0.2-GSE48684_output.Rdata",
  GSE77718  = "step0.3-GSE77718_output.Rdata",
  GSE101764 = "step0.4-GSE101764_output.Rdata",
  GSE131013 = "step0.5-GSE131013_output.Rdata")

## Output directory
outdir <- "independent_validation_5datasets_check"

if (!dir.exists(outdir)) {dir.create(outdir)}
```

All validation inputs and outputs were assigned fixed file paths.

#### Set the validation criteria

min_cpg_required <- 5 means at least 6 CpGs are required,which matches my discovery criterion: 
no.cpgs >= 6;
deltaBeta_validation is Tumor - Normal;
For hyper-DMR validation: Tumor - Normal > 0.20.
```{r}
min_cpg_required <- 5
delta_cutoff <- 0.20
p_cutoff <- 0.05
fdr_cutoff <- 0.05
```

#### Load discovery DMRs from CSV
```{r}
dmr_obj <- read.csv(discovery_dmr_file,
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = FALSE)

print(discovery_dmr_file)
dim(dmr_obj)
#View(dmr_obj)
```

#### Define a function to identify column names
This helper function finds the first available column that matches a list of possible names.
```{r}
match_col <- function(df, candidates) {
  
  idx <- match(tolower(candidates), tolower(colnames(df)))
  idx <- idx[!is.na(idx)]
  
  if (length(idx) == 0) {
    return(NA_character_)
  } else {
    return(colnames(df)[idx[1]])
  }
}
```
`match_col()` was created for DMR-table standardization.

#### Define a function to identify the CpG column
This helper function searches the DMR table for a column containing Illumina CpG probe IDs.
```{r}
find_cpg_col <- function(df) {
  
  candidate_cols <- colnames(df)[
    grepl("cpg|probe|probes|name|id", colnames(df), ignore.case = TRUE)
  ]
  
  if (length(candidate_cols) == 0) {
    return(NA_character_)
  }
  
  for (cn in candidate_cols) {
    
    vals <- as.character(df[[cn]])
    vals <- vals[!is.na(vals)]
    vals <- vals[seq_len(min(length(vals), 100))]
    
    if (any(grepl("cg[0-9]{8}", vals))) {
      return(cn)
    }
  }
  
  return(NA_character_)
}
```
`find_cpg_col()` returns the CpG column name when one is available.

#### Define a function to identify the coordinate column
This helper function searches for genomic coordinates written in `chr:start-end` format.
```{r}
find_coord_col <- function(df) {
  
  candidate_cols <- colnames(df)[
    grepl("coord|region|position|location|dmr", colnames(df), ignore.case = TRUE)
  ]
  
  if (length(candidate_cols) == 0) {
    candidate_cols <- colnames(df)
  }
  
  for (cn in candidate_cols) {
    
    vals <- as.character(df[[cn]])
    vals <- vals[!is.na(vals)]
    vals <- vals[seq_len(min(length(vals), 100))]
    
    if (any(grepl("^chr[^:]+:[0-9,]+-[0-9,]+$", vals))) {
      return(cn)
    }
  }
  
  return(NA_character_)
}
```
`find_coord_col()` was created for coordinate detection.

#### Define a function to standardize chromosome names
This function converts chromosome labels to a consistent `chr` format.
```{r}
standardize_chr <- function(x) {
  
  x <- as.character(x)
  x <- gsub("^CHR", "chr", x, ignore.case = TRUE)
  x <- ifelse(grepl("^chr", x, ignore.case = TRUE), x, paste0("chr", x))
  
  return(x)
}
```
Chromosome names were standardized for annotation matching.

#### Define a function to parse genomic coordinates
This function separates a genomic coordinate into chromosome, start, and end values.
```{r}
parse_coord <- function(coord) {
  
  coord <- as.character(coord)
  coord <- gsub(",", "", coord)
  
  ok <- grepl("^chr[^:]+:[0-9]+-[0-9]+$", coord)
  
  chr <- rep(NA_character_, length(coord))
  start <- rep(NA_integer_, length(coord))
  end <- rep(NA_integer_, length(coord))
  
  chr[ok] <- sub(":.*", "", coord[ok])
  start[ok] <- as.integer(sub(".*:([0-9]+)-[0-9]+$", "\\1", coord[ok]))
  end[ok] <- as.integer(sub(".*:[0-9]+-([0-9]+)$", "\\1", coord[ok]))
  
  out <- data.frame(
    chr = standardize_chr(chr),
    start = start,
    end = end,
    stringsAsFactors = FALSE
  )
  
  return(out)
}
```
`parse_coord()` was created for coordinate conversion.

#### Define a function to standardize the DMR table
This function identifies the DMR coordinates and CpG information and creates a standard DMR table.
```{r}
standardize_dmr_table <- function(dmr_obj) {
  
  dmr.df <- as.data.frame(dmr_obj)
  
  chr_col <- match_col(
    dmr.df,
    c("chr", "chrom", "chromosome", "seqnames")
  )
  
  start_col <- match_col(
    dmr.df,
    c("start", "start_position", "pos_start", "dmr_start")
  )
  
  end_col <- match_col(
    dmr.df,
    c("end", "end_position", "pos_end", "dmr_end")
  )
  
  coord_col <- find_coord_col(dmr.df)
  cpg_col <- find_cpg_col(dmr.df)
  
  rowname_coord <- NULL
  
  if (!is.null(rownames(dmr.df))) {
    rn <- rownames(dmr.df)
    
    if (any(grepl("^chr[^:]+:[0-9,]+-[0-9,]+$", rn))) {
      rowname_coord <- parse_coord(rn)
    }
  }
  
  coord_parsed <- NULL
  
  if (!is.na(coord_col)) {
    coord_vals <- as.character(dmr.df[[coord_col]])
    
    if (any(grepl("^chr[^:]+:[0-9,]+-[0-9,]+$", coord_vals))) {
      coord_parsed <- parse_coord(coord_vals)
    }
  }
  
  ## chr
  if (!is.na(chr_col)) {
    chr_out <- standardize_chr(dmr.df[[chr_col]])
  } else if (!is.null(coord_parsed)) {
    chr_out <- coord_parsed$chr
  } else if (!is.null(rowname_coord)) {
    chr_out <- rowname_coord$chr
  } else {
    chr_out <- rep(NA_character_, nrow(dmr.df))
  }
  
  ## start
  if (!is.na(start_col)) {
    start_out <- as.integer(gsub(",", "", dmr.df[[start_col]]))
  } else if (!is.null(coord_parsed)) {
    start_out <- coord_parsed$start
  } else if (!is.null(rowname_coord)) {
    start_out <- rowname_coord$start
  } else {
    start_out <- rep(NA_integer_, nrow(dmr.df))
  }
  
  ## end
  if (!is.na(end_col)) {
    end_out <- as.integer(gsub(",", "", dmr.df[[end_col]]))
  } else if (!is.null(coord_parsed)) {
    end_out <- coord_parsed$end
  } else if (!is.null(rowname_coord)) {
    end_out <- rowname_coord$end
  } else {
    end_out <- rep(NA_integer_, nrow(dmr.df))
  }
  
  ## DMR ID
  if (!all(is.na(chr_out)) && !all(is.na(start_out)) && !all(is.na(end_out))) {
    dmr_id <- paste0(chr_out, ":", start_out, "-", end_out)
  } else if (!is.na(coord_col)) {
    dmr_id <- as.character(dmr.df[[coord_col]])
  } else if (!is.null(rownames(dmr.df)) &&
             !all(rownames(dmr.df) == as.character(seq_len(nrow(dmr.df))))) {
    dmr_id <- rownames(dmr.df)
  } else {
    dmr_id <- paste0("DMR_", seq_len(nrow(dmr.df)))
  }
  
  dmr_id <- make.unique(as.character(dmr_id))
  
  out <- data.frame(
    DMR_id = dmr_id,
    chr = chr_out,
    start = start_out,
    end = end_out,
    cpg_string = if (!is.na(cpg_col)) as.character(dmr.df[[cpg_col]]) else NA_character_,
    stringsAsFactors = FALSE
  )
  
  if (all(is.na(out$chr)) && all(is.na(out$cpg_string))) {
    stop("No chromosome coordinates or CpG probe IDs were found in the discovery DMR table.")
  }
  
  return(out)
}
```
`standardize_dmr_table()` was created to support different possible discovery-table formats.

#### Standardize the discovery DMR table
The discovery DMR table was converted to standard DMR ID, chromosome, start, end, and CpG-string columns.
```{r}
dmr.tbl <- standardize_dmr_table(dmr_obj)
dim(dmr.tbl)
colSums(is.na(dmr.tbl[, c("chr", "start", "end")]))
```

#### Load the Illumina 450K annotation
The hg19 Illumina 450K annotation was loaded and converted into a CpG-coordinate table.
```{r}
anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)

anno.df <- data.frame(
  CpG = rownames(anno),
  chr = standardize_chr(anno$chr),
  pos = as.integer(anno$pos),
  stringsAsFactors = FALSE
)

anno.dt <- as.data.table(anno.df)
```
The annotation table was prepared for mapping CpGs to DMRs.

#### Define a function to map CpGs to each DMR
This function first uses CpG IDs stored in the DMR table and otherwise maps probes using chromosome coordinates.
```{r}
build_dmr_cpg_list <- function(dmr.tbl, anno.dt) {
  
  cpg.list <- vector("list", nrow(dmr.tbl))
  names(cpg.list) <- dmr.tbl$DMR_id
  
  for (i in seq_len(nrow(dmr.tbl))) {
    
    cpgs <- character(0)
    
    ## First, try to extract CpG IDs directly from the discovery DMR table
    if (!is.na(dmr.tbl$cpg_string[i]) &&
        grepl("cg[0-9]{8}", dmr.tbl$cpg_string[i])) {
      
      cpgs <- unique(unlist(regmatches(
        dmr.tbl$cpg_string[i],
        gregexpr("cg[0-9]{8}", dmr.tbl$cpg_string[i])
      )))
    }
    
    ## If no CpG IDs are available, map CpGs by chromosome coordinates
    if (length(cpgs) == 0 &&
        !is.na(dmr.tbl$chr[i]) &&
        !is.na(dmr.tbl$start[i]) &&
        !is.na(dmr.tbl$end[i])) {
      
      chr_i <- dmr.tbl$chr[i]
      start_i <- dmr.tbl$start[i]
      end_i <- dmr.tbl$end[i]
      
      cpgs <- anno.dt[
        chr == chr_i & pos >= start_i & pos <= end_i,
        CpG
      ]
    }
    
    cpg.list[[i]] <- unique(cpgs)
  }
  
  return(cpg.list)
}
```
`build_dmr_cpg_list()` was created for probe mapping.

#### Map 450K CpGs to the discovery DMRs
The CpG list was generated for every DMR, and the number of mapped probes was counted.
```{r}
dmr.cpg.list <- build_dmr_cpg_list(dmr.tbl, anno.dt)

dmr.tbl$nCpG_450K_annotation <- lengths(dmr.cpg.list)

summary(dmr.tbl$nCpG_450K_annotation)

sum(dmr.tbl$nCpG_450K_annotation == 0)
```
DMRs with no mapped 450K probes cannot be tested in the validation datasets.

#### Define a function to load each validation dataset

This function loads an `.Rdata` file and extracts `myLoad$beta` and `myLoad$pd`.
```{r define_load_validation_dataset}

load_validation_dataset <- function(file) {
  
  env <- new.env()
  loaded_objects <- load(file, envir = env)
  
  if (!"myLoad" %in% loaded_objects) {
    stop(paste0(
      "Object 'myLoad' was not found in file: ",
      file,
      ". Objects found: ",
      paste(loaded_objects, collapse = ", ")
    ))
  }
  
  obj <- get("myLoad", envir = env)
  
  if (!is.list(obj)) {
    stop(paste0("myLoad is not a list in file: ", file))
  }
  
  if (!"beta" %in% names(obj)) {
    stop(paste0("Cannot find myLoad$beta in file: ", file))
  }
  
  if (!"pd" %in% names(obj)) {
    stop(paste0("Cannot find myLoad$pd in file: ", file))
  }
  
  beta <- obj$beta
  pd <- obj$pd
  
  beta <- as.matrix(beta)
  mode(beta) <- "numeric"
  
  pd <- as.data.frame(pd)
  
  return(list(beta = beta, pd = pd))
}
```
`load_validation_dataset()` checks that the expected objects are available.

#### Define a function to standardize sample groups

This function converts different phenotype descriptions to the standard labels Normal and Tumor.
```{r}
normalize_group <- function(group_vector) {
  
  x <- as.character(group_vector)
  out <- rep(NA_character_, length(x))
  
  out[grepl("tumor|tumour|cancer|carcinoma|crc", x, ignore.case = TRUE)] <- "Tumor"
  out[grepl("normal|control|adjacent|mucosa|healthy", x, ignore.case = TRUE)] <- "Normal"
  
  return(factor(out, levels = c("Normal", "Tumor")))
}
```
`normalize_group()` ensures consistent group labels across datasets.

#### Define a function to prepare each validation dataset

This function aligns the phenotype table and beta matrix, identifies the group column, retains Normal and Tumor samples, and checks the sample numbers.
```{r}
prepare_validation_dataset <- function(dataset_name, file) {
  
  dat <- load_validation_dataset(file)
  
  beta <- dat$beta
  pd <- dat$pd
  
  ## Set row names of phenotype table using ID column if possible
  if ("ID" %in% colnames(pd) && setequal(as.character(pd$ID), colnames(beta))) {
    rownames(pd) <- as.character(pd$ID)}
  
  ## If sample names are the same but order is different, reorder phenotype table
  if (setequal(rownames(pd), colnames(beta))) {pd <- pd[colnames(beta), , drop = FALSE]}
  
  ## Stop if beta matrix and phenotype table cannot be aligned
  if (!identical(rownames(pd), colnames(beta))) {
    stop(paste0("Sample alignment failed in ",
      dataset_name,". Please check rownames(pd), pd$ID, and colnames(beta)."))}
  
  ## Find group column
  group_col <- NA_character_
  
  for (cn in c("group", "Group", "condition", "Condition", "phenotype", "Phenotype")) {
    if (cn %in% colnames(pd)) {
      group_col <- cn
      break}}
  
  if (is.na(group_col)) {
    stop(paste0("No group column was found in phenotype table of ", dataset_name))}
  
  ## Standardize group labels to Normal and Tumor
  pd$group <- normalize_group(pd[[group_col]])
  
  ## Keep only Normal and Tumor samples
  keep <- !is.na(pd$group) & pd$group %in% c("Normal", "Tumor")
  
  beta <- beta[, keep, drop = FALSE]
  pd <- pd[keep, , drop = FALSE]
  
  ## Set clean ID column
  pd$ID <- rownames(pd)
  
  ## Check sample size
  if (sum(pd$group == "Normal") < 2 || sum(pd$group == "Tumor") < 2) {
    stop(paste0("Too few samples after filtering in ",
      dataset_name,
      ". Need at least 2 Normal and 2 Tumor samples."))}
  
  ## Sort samples by group
  sample_order <- order(pd$group)
  pd <- pd[sample_order, , drop = FALSE]
  beta <- beta[, rownames(pd), drop = FALSE]
  
  ## Final strict check
  stopifnot(identical(rownames(pd), colnames(beta)))
  
  cat("\nDataset:", dataset_name, "\n")
  cat("beta dimension:\n")
  print(dim(beta))
  cat("group distribution:\n")
  print(table(pd$group, useNA = "ifany"))
  cat("alignment check:\n")
  print(identical(rownames(pd), colnames(beta)))
  
  return(list(beta = beta, pd = pd))
}
```
`prepare_validation_dataset()` returns an aligned beta matrix and phenotype table.

#### Define the DMR validation function

For each DMR, this function calculates the mean Normal and Tumor beta values, delta beta, t-test p-value, FDR, and validation status.
```{r}
validate_one_dataset <- function(dataset_name, file, dmr.tbl, dmr.cpg.list) {
  
  dat <- prepare_validation_dataset(dataset_name, file)
  
  beta <- dat$beta
  pd <- dat$pd
  
  normal_samples <- rownames(pd)[pd$group == "Normal"]
  tumor_samples <- rownames(pd)[pd$group == "Tumor"]
  
  res.list <- vector("list", nrow(dmr.tbl))
  
  for (i in seq_len(nrow(dmr.tbl))) {
    
    dmr_id <- dmr.tbl$DMR_id[i]
    
    ## Use list index instead of DMR_id name to avoid duplicated-name problems
    cpgs_all <- dmr.cpg.list[[i]]
    cpgs_present <- intersect(cpgs_all, rownames(beta))
    
    nCpG <- length(cpgs_present)
    
    if (nCpG <= min_cpg_required) {
      
      res.list[[i]] <- data.frame(
        dataset = dataset_name,
        DMR_id = dmr_id,
        chr = dmr.tbl$chr[i],
        start = dmr.tbl$start[i],
        end = dmr.tbl$end[i],
        nCpG = nCpG,
        mean_Normal = NA_real_,
        mean_Tumor = NA_real_,
        deltaBeta_validation = NA_real_,
        p.value = NA_real_,
        stringsAsFactors = FALSE
      )
      
      next
    }
    
    ## Calculate mean beta value of this DMR for each sample
    dmr_beta <- colMeans(beta[cpgs_present, , drop = FALSE], na.rm = TRUE)
    dmr_beta[is.nan(dmr_beta)] <- NA_real_
    
    normal_values <- dmr_beta[normal_samples]
    tumor_values <- dmr_beta[tumor_samples]
    
    mean_normal <- mean(normal_values, na.rm = TRUE)
    mean_tumor <- mean(tumor_values, na.rm = TRUE)
    
    delta_beta <- mean_tumor - mean_normal
    
    pval <- tryCatch(
      t.test(tumor_values, normal_values)$p.value,
      error = function(e) NA_real_
    )
    
    res.list[[i]] <- data.frame(
      dataset = dataset_name,
      DMR_id = dmr_id,
      chr = dmr.tbl$chr[i],
      start = dmr.tbl$start[i],
      end = dmr.tbl$end[i],
      nCpG = nCpG,
      mean_Normal = mean_normal,
      mean_Tumor = mean_tumor,
      deltaBeta_validation = delta_beta,
      p.value = pval,
      stringsAsFactors = FALSE
    )
  }
  
  res <- rbindlist(res.list, fill = TRUE)
  
  ## Adjust p values within each validation dataset
  res$FDR <- p.adjust(res$p.value, method = "BH")
  
  ## Main validation criterion for discovery hyper-DMRs
  res$validated_p005 <- with(
    res,
    nCpG > min_cpg_required &
      deltaBeta_validation > delta_cutoff &
      p.value < p_cutoff
  )
  
  ## More stringent FDR-based criterion
  res$validated_FDR005 <- with(
    res,
    nCpG > min_cpg_required &
      deltaBeta_validation > delta_cutoff &
      FDR < fdr_cutoff
  )
  
  return(res)
}
```
`validate_one_dataset()` returns one validation result for every DMR in a dataset.

#### Run validation across all five datasets

The DMR validation function was applied to each GEO dataset, and all results were combined.
```{r}
all.validation.list <- list()

for (dataset_name in names(dataset_files)) {
  
  cat("\n============================================================\n")
  cat("Running validation for:", dataset_name, "\n")
  cat("============================================================\n")
  
  all.validation.list[[dataset_name]] <- validate_one_dataset(
    dataset_name = dataset_name,
    file = dataset_files[[dataset_name]],
    dmr.tbl = dmr.tbl,
    dmr.cpg.list = dmr.cpg.list
  )
}

all.validation <- rbindlist(all.validation.list, fill = TRUE)

cat("\nAll validation result dimension:\n")
print(dim(all.validation))
print(head(all.validation))
```
The combined table contains one result for each DMR-dataset combination.

#### Define safe summary functions

These functions calculate the mean, maximum, and minimum while safely handling missing values.

```{r}
## Summarize validation results across datasets
safe_mean <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  return(mean(x))
}

safe_max <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  return(max(x))
}

safe_min <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  return(min(x))
}
```
The helper functions prevent invalid summary values when no result is available.

#### Define the cross-dataset DMR summary function

This function summarizes each DMR across datasets and combines its p-values using Fisher's method.
```{r}
summarize_one_dmr <- function(df) {
  
  pvals <- df$p.value
  pvals <- pvals[!is.na(pvals)]
  
  fisher_p <- NA_real_
  
  if (length(pvals) > 0) {
    pvals <- pmax(pvals, .Machine$double.xmin)
    fisher_p <- pchisq(
      -2 * sum(log(pvals)),
      df = 2 * length(pvals),
      lower.tail = FALSE
    )
  }
  
  data.frame(
    DMR_id = df$DMR_id[1],
    chr = df$chr[1],
    start = df$start[1],
    end = df$end[1],
    tested_datasets = sum(!is.na(df$p.value)),
    validated_datasets_p005 = sum(df$validated_p005, na.rm = TRUE),
    validated_datasets_FDR005 = sum(df$validated_FDR005, na.rm = TRUE),
    mean_deltaBeta = safe_mean(df$deltaBeta_validation),
    max_deltaBeta = safe_max(df$deltaBeta_validation),
    min_p.value = safe_min(pvals),
    fisher_p.value = fisher_p,
    stringsAsFactors = FALSE
  )
}
```
`summarize_one_dmr()` produces one cross-dataset summary row per DMR.

#### Summarize and rank all DMRs

All DMR results were summarized across the five datasets, adjusted for multiple testing, and ranked by validation strength.
```{r}
summary.list <- lapply(
  split(all.validation, all.validation$DMR_id),
  summarize_one_dmr
)

validation.summary <- rbindlist(summary.list, fill = TRUE)

validation.summary$meta_FDR <- p.adjust(
  validation.summary$fisher_p.value,
  method = "BH"
)

validation.summary$validated_any_p005 <- validation.summary$validated_datasets_p005 >= 1
validation.summary$validated_at_least2_p005 <- validation.summary$validated_datasets_p005 >= 2
validation.summary$validated_any_FDR005 <- validation.summary$validated_datasets_FDR005 >= 1

## Sort DMRs by validation strength
validation.summary <- validation.summary[
  order(
    -validated_datasets_p005,
    -mean_deltaBeta,
    min_p.value
  ),
]

cat("\nValidation summary dimension:\n")
print(dim(validation.summary))
print(head(validation.summary))
```
The leading rows represent the strongest independently validated hyper-DMRs.

#### Save the main validation results

The complete validation table, cross-dataset summary, and p-value-based DMR subsets were saved.
```{r}
## Save all per-dataset results
fwrite(all.validation,
  file = file.path(outdir, "validation_results_all_DMRs_all_5datasets.csv")
)

## Save summary results across all validation datasets
fwrite(validation.summary,
  file = file.path(outdir, "validation_summary_across_5datasets.csv"))

## Save DMRs validated in at least one dataset using p < 0.05 criterion
validated_p005 <- validation.summary[
  validation.summary$validated_any_p005 == TRUE,]

fwrite(validated_p005,
  file = file.path(outdir, "validated_hyper_DMRs_at_least_one_dataset_p005.csv"))

## Save DMRs validated in at least two datasets using p < 0.05 criterion
validated_at_least2_p005 <- validation.summary[
  validation.summary$validated_at_least2_p005 == TRUE,]

fwrite(validated_at_least2_p005,
  file = file.path(outdir, "validated_hyper_DMRs_at_least_two_datasets_p005.csv"))
```
These CSV files contain the main validation results.

#### Save the FDR and dataset-specific results

The FDR-based DMR subset and separate result files for each dataset were saved.
```{r}
## Save DMRs validated in at least one dataset using FDR < 0.05 criterion
validated_FDR005 <- validation.summary[
  validation.summary$validated_any_FDR005 == TRUE,]

fwrite(validated_FDR005,
  file = file.path(outdir, "validated_hyper_DMRs_at_least_one_dataset_FDR005.csv"))

## Save per-dataset results separately
for (dataset_name in names(dataset_files)) {
  
  tmp <- all.validation[all.validation$dataset == dataset_name, ]
  
  fwrite(tmp,
    file = file.path(
      outdir,
      paste0(dataset_name, "_validation_results_all_DMRs.csv")))}
```

These files support stricter filtering and dataset-level checking.

#### Create the dataset-level summary and bar plot

The number of tested and validated DMRs was summarized for each dataset, and a bar plot was saved.
```{r}
dataset.summary <- rbindlist(lapply(
  split(all.validation, all.validation$dataset),
  function(df) {
    data.frame(
      dataset = df$dataset[1],
      tested_DMRs = sum(!is.na(df$p.value)),
      validated_p005 = sum(df$validated_p005, na.rm = TRUE),
      validated_FDR005 = sum(df$validated_FDR005, na.rm = TRUE),
      stringsAsFactors = FALSE)}))

fwrite(dataset.summary,
  file = file.path(outdir, "validation_summary_by_dataset.csv"))

png(filename = file.path(outdir, "validation_summary_barplot.png"),
  width = 1600,
  height = 900,
  res = 150)

barplot(dataset.summary$validated_p005,
  names.arg = dataset.summary$dataset,
  las = 2,
  ylab = "Number of validated hyper-DMRs",
  main = "Independent validation of discovery hyper-DMRs")

dev.off()

print(dataset.summary)
```
The table and plot show the validation performance of the five datasets.

#### Export robustly validated hyper-DMRs

DMRs validated in at least three, at least four, or all five datasets were selected and saved.
```{r}
robust_validated_DMRs <- validation.summary[validation.summary$validated_datasets_p005 >= 3,]

high_confidence_DMRs <- validation.summary[validation.summary$validated_datasets_p005 >= 4,]

all5_validated_DMRs <- validation.summary[validation.summary$validated_datasets_p005 == 5,]


print(dim(robust_validated_DMRs))

print(dim(high_confidence_DMRs))

print(dim(all5_validated_DMRs))

fwrite(robust_validated_DMRs,
  file = file.path(outdir, "robust_validated_hyper_DMRs_at_least_3datasets.csv"))

fwrite(high_confidence_DMRs,
  file = file.path(outdir, "high_confidence_hyper_DMRs_at_least_4datasets.csv"))

fwrite(all5_validated_DMRs,
  file = file.path(outdir, "hyper_DMRs_validated_in_all_5datasets.csv"))
```

## session information
```{r}
sessionInfo()
```

