---
title: "step1.DMR discovery by combining 3 datasets"
author: "YC"
date: "2026-06-22"
output:
 html_document:
  toc: true
  toc_float: true
  code_folding: hide
  fig_width: 7
  fig_height: 7
  theme: cosmo
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#🍁🍁 step0.Data preprocessing-standard beta value matrix and pd group information

## dataset1: GSE199057

Here we will process methylation data from GSE199057.

The pipeline structure is given below:

GSE199057 EPIC methylation processed matrix pipeline

1. Read GEO metadata.

2. Build phenotype table pD.

3. Extract AVG_Beta columns from the processed matrix.

4. Map processed sample IDs to GSM IDs.

5. Remove plate-control samples.

6. Generate a strict beta matrix: CpG probes  ^w GSM samples.

7. Remove probes with missing beta values.

8. Align pD rows with beta matrix columns.

9. Run ChAMP filtering.

### Load required packages

This chunk clears the current R environment and loads all packages required for data preprocessing, statistical analysis, annotation, DMR calling, and table display.

```{r load-libraries}
rm(list = ls())
options(stringsAsFactors = F)
require(GEOquery)
require(Biobase)
library("impute")
library("ChAMP")
library("data.table")
library("limma")
library("metafor")
library("minfi")
library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
library("kableExtra")

```
After this chunk, all packages required by the subsequent analysis should be available.

### Read GEO metadata
This chunk downloads the GSE199057 GEO Series object. The resulting `eset` object contains the sample metadata used to construct the phenotype table.
```{r GSE199057-read-metadata}
eset <- getGEO(
  "GSE199057",
  destdir = './',
  AnnotGPL = T,
  getGPL = F)
```
The object `eset` is used in the next chunk to obtain sample-level metadata.

### Build phenotype table pD
This chunk constructs the GSE199057 phenotype table and classifies samples as `Normal` or `Tumor` according to the GEO sample description.
Here we build a clean phenotype table:
1.ID is the GSM accession
2.title is the original GEO sample description
3.pD.all contains sample-level information, including:
- GSM accession IDs
- sample titles
- clinical/sample descriptions
⭐This part might need be modified when applying the pipeline to another dataset
```{r GSE199057-build-phenotype}
pD.all <- pData(eset[[1]])
pD <- pD.all[,c(1,2)]
names(pD) <- c("group", "ID")
pD$group <- ifelse(
grepl("normal",pD$group,ignore.case = T),
  "Normal",
  "Tumor")
pd <-pD
pd <-pD
#check if the extracted groupings are correct
check_df <- data.frame(Title = pD.all$title, Group = pD$group)
head(check_df, 10)
tail(check_df,10)
table(pd$group)
```
The displayed head, tail, and group table should be inspected to confirm that the sample grouping has been extracted correctly.
### Read the processed beta matrix
#### This chunk reads the supplementary processed methylation matrix supplied for GSE199057.
For GSE199057, this matrix contains:
-TargetID
-sample_id.AVG_Beta
-sample_id.Detection Pval

```{r GSE199057-read-beta-matrix}
GSE199057_DATA="GSE199057_processed_methlyation_matrix.txt.gz"
GSE199057_URL="https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE199057&format=file&file=GSE199057%5Fprocessed%5Fmethlyation%5Fmatrix%2Etxt%2Egz"
if ( ! file.exists(GSE199057_DATA) ) {options(timeout = 300)download.file(GSE199057_URL,destfile=GSE199057_DATA)}
beta.m <- fread("GSE199057_processed_methlyation_matrix.txt.gz",data.table = F)
dim(beta.m)
```
The reported dimensions show the number of rows and columns in the unprocessed input table.
#### This chunk identifies columns ending in `.AVG_Beta` and extracts the corresponding internal sample identifiers.
```{r GSE199057-extract-beta-columns}
beta_cols <- grep(
  "\\.AVG_Beta$", 
  colnames(beta.m), 
  value = TRUE)
## Example:"1018133.AVG_Beta" becomes "1018133"
beta_sample_id <- sub("\\.AVG_Beta$", "", beta_cols)
head(beta_sample_id)
```
The displayed identifiers should correspond to the sample identifiers embedded in the GEO titles.
### Map internal sample IDs to GSM accessions
#### This chunk creates a mapping between processed-matrix sample IDs and GEO GSM accessions.
```{r GSE199057-build-sample-map}
sample_map <- data.frame(
  GSM = pD.all$geo_accession,
  title = pD.all$title,
  stringsAsFactors = FALSE)

sample_map$SampleID <- sub(".*\\[([^]]+)\\].*", "\\1", sample_map$title)
head(sample_map)
## Match processed matrix sample IDs to GEO metadata sample IDs.
idx <- match(beta_sample_id, sample_map$SampleID)

beta_to_GSM <- data.frame(
  beta_column = beta_cols,
  SampleID = beta_sample_id,
  GSM = sample_map$GSM[idx],
  title = sample_map$title[idx],
  stringsAsFactors = FALSE)

head(beta_to_GSM)
#View(beta_to_GSM)
## Check unmatched samples and NA value
sum(is.na(beta_to_GSM$GSM))
beta_to_GSM[is.na(beta_to_GSM$GSM), ]
```
The unmatched-sample output identifies any matrix columns that could not initially be linked to GEO metadata.

#### This chunk applies the two known manual corrections for technical replicate identifiers whose punctuation differs between the processed matrix and GEO metadata.
In GSE199057, two technical replicates use hyphens in the processed matrix:
-T1003588-rep1
-T1003588-rep2
But GEO metadata title uses spaces:
-T1003588 rep1
-T1003588 rep2
Therefore, manually assign their GSM IDs.

```{r gse199057-fix-known-id-mismatches}
sample_map[grepl("T1003588", sample_map$title, ignore.case = TRUE),
                       c("GSM", "SampleID", "title")]
beta_to_GSM$GSM[beta_to_GSM$SampleID == "T1003588-rep1"] <- "GSM5963372"
beta_to_GSM$title[beta_to_GSM$SampleID == "T1003588-rep1"] <- 
  "colon tumor sample from CRC patient [T1003588 rep1]"

beta_to_GSM$GSM[beta_to_GSM$SampleID == "T1003588-rep2"] <- "GSM5963373"
beta_to_GSM$title[beta_to_GSM$SampleID == "T1003588-rep2"] <- 
  "colon tumor sample from CRC patient [T1003588 rep2]"
## Recheck unmatched samples after manual correction.
sum(is.na(beta_to_GSM$GSM))

beta_to_GSM[is.na(beta_to_GSM$GSM), ]
```
After correction, the number of missing GSM mappings should be checked again.
#### This chunk removes plate-control samples before biological Tumor-versus-Normal analysis.
Plate-control samples are technical controls, which should not be included in Tumor vs NAT / healthy_colon analysis.
```{r gse199057-remove-plate-controls}
beta_to_GSM_clean <- beta_to_GSM[
  !grepl("Plate-Control", beta_to_GSM$SampleID, ignore.case = TRUE),
]
## Confirm that all retained samples have GSM IDs.
sum(is.na(beta_to_GSM_clean))
```
The remaining mapping table should contain only biological samples with valid GSM identifiers.
### Build and align the beta matrix
This chunk extracts the beta-value columns, converts them to a numeric matrix, and assigns CpG probe IDs as row names and GSM accessions as column names.
```{r gse199057-build-standard-beta-matrix}
## Generate strict beta matrix
## Extract only AVG_Beta columns and convert to a numeric matrix.
beta <- as.matrix(beta.m[, beta_to_GSM_clean$beta_column, drop = FALSE])
mode(beta) <- "numeric"
rownames(beta) <- beta.m$TargetID## Set CpG probe IDs as row names.
colnames(beta) <- beta_to_GSM_clean$GSM## Set GSM IDs as column names.

## Check beta matrix.
## Expected:
##   rows = CpG probes
##   columns = biological GSM samples
##   values = beta values between 0 and 1
dim(beta) #224 samples
beta.m1 <- beta
```
The matrix dimensions report the number of CpG probes and retained biological samples.
#### This chunk removes phenotype records that are not represented in the beta matrix and reorders the phenotype table to match the beta-matrix columns.
```{r gse199057-align-phenotype}
## Align pD with beta.m1
## Find samples present in pd but not in beta.m1
samples_to_remove <- setdiff(pd$ID, colnames(beta.m1)); samples_to_remove
length(samples_to_remove)

## Check the metadata of these samples
pd[pd$ID %in% samples_to_remove, ]
## Remove these samples from pd
pd <- pd[!pd$ID %in% samples_to_remove, , drop = FALSE]
## Reorder pd to match beta.m1 column order
pd <- pd[match(colnames(beta.m1), pd$ID), , drop = FALSE]
## Set row names as GSM IDs
rownames(pd) <- pd$ID

## Final checks
setequal(colnames(beta.m1), pd$ID)
identical(colnames(beta.m1), pd$ID)
identical(colnames(beta.m1), rownames(pd))

dim(beta.m1)
dim(pd)
table(pd$group)
```
The equality checks should confirm that beta-matrix columns and phenotype rows contain the same samples in the same order.
#### This chunk orders the samples by group and applies the same order to the phenotype table and beta matrix.
```{r gse199057-sort-samples}
# Set group order
pd$group <- factor(
  pd$group,
  levels = c("Normal","Tumor"))

# Sort pd by group
sample_order <- order(pd$group)
pd <- pd[sample_order, , drop = FALSE]
# Set rownames as GSM ID
rownames(pd) <- pd$ID
# Reorder beta matrix columns according to pD
beta.m1 <- beta.m1[, rownames(pd), drop = FALSE]
# Final check
identical(rownames(pd), colnames(beta.m1))
table(pd$group)
pd1 <- pd
```
The final identity check should be `TRUE`, confirming strict sample alignment before filtering.
### Apply ChAMP filtering
This chunk applies `champ.filter()` to the aligned GSE199057 beta matrix and saves the resulting `myLoad` object.
```{r gse199057-champ-filter}
## Run ChAMP filtering using the cleaned beta matrix and aligned pd.
myLoad <- champ.filter(beta = beta.m1, pd = pd1, arraytype = "EPIC")
## Save the ChAMP object for downstream analysis.
save(myLoad,file = 'step0.1-GSE199057_output.Rdata')

```
For all downstream analyses, simply use this myLoad variable.
champ.filter() performs initial quality control:
(1).Filters out probes with low detection p-value.
(2).Removes probes with missing values.
(3).Filters sex chromosomes or SNP-associated probes if needed.
(4).The output myLoad is an object used for all downstream ChAMP analyses 
(normalization, DMP, DMR, pathway analysis).

## dataset2: GSE149282
The GSE149282 workflow obtains the processed beta matrix directly from the GEO Series object, constructs the Tumor/Normal phenotype table, aligns samples, and applies ChAMP filtering.
### Read data and metadata
#### This chunk clears the workspace, loads the required packages, and reads the GSE149282 beta matrix and GEO metadata object.
```{r gse149282-load-and-read}
rm(list = ls())
options(stringsAsFactors = F)
require(GEOquery)
require(Biobase)
# Download/read GEO series matrix metadata
eset <- getGEO("GSE149282",destdir = './',AnnotGPL = T,getGPL = F)
beta.m <- exprs(eset[[1]])
class(beta.m)
dim(beta.m)
```
The class and dimensions confirm the structure and size of the beta matrix.
#### This chunk constructs the GSE149282 phenotype table and assigns samples to `Normal` or `Tumor` groups.
```{r gse149282-build-phenotype}
pD.all <- pData(eset[[1]])
pD <- pD.all[,c(1,2)]
names(pD) <- c("group", "ID")
pD$group <-ifelse(
  grepl("normal",pD$group,ignore.case = T),
  "Normal",
  "Tumor")
# check if the extracted groupings are correct
check_df <- data.frame(Title = pD.all$title, Group = pD$group)
head(check_df, 10)
tail(check_df,10)
table(pD$group)
```
The displayed sample titles and group counts should be reviewed to verify the classification.
### Align samples and filter probes

This chunk checks sample correspondence, removes ordering differences, sets GSM accessions as phenotype row names, and applies a common group order.
```{r gse149282-align-samples}
stopifnot(setequal(colnames(beta.m), pD$ID))
stopifnot(identical(rownames(pD),colnames(beta.m)))##This must be TRUE before champ.filter().
# Confirm that there are no duplicated sample names
sum(duplicated(colnames(beta.m)))
sum(duplicated(pD$ID))

# Reorder pD to match the column names of beta.m1
pD <- pD[match(colnames(beta.m), pD$ID), , drop = FALSE]

# Set pD row names as GSM IDs.
rownames(pD) <- pD$ID
## Final strict alignment check.
dim(beta.m)
dim(pD)

# Set group order
pD$group <- factor(
  pD$group,
  levels = c("Normal", "Tumor"))

# Sort pD by group
sample_order <- order(pD$group)

pD <- pD[sample_order, , drop = FALSE]

# Reorder beta matrix columns according to pD
beta.m <- beta.m[, rownames(pD), drop = FALSE]
```
The beta matrix and phenotype table should now describe the same samples in the same order.
#### This chunk applies ChAMP filtering and saves the processed GSE149282 object.
```{r gse149282-champ-filter}
myLoad <- champ.filter(beta = beta.m, pd = pD, arraytype = "EPIC")

save(myLoad,file = 'step0.2-GSE149282_output.Rdata')
```
The output file-step0.2-GSE149282_output.Rdata is used in the combined analysis.

## dataset3：GSE119526
The GSE119526 workflow reads the GEO beta matrix, prepares the Tumor/Normal phenotype information, orders the samples consistently, and applies ChAMP filtering.
### Read data and define groups
This chunk clears the workspace, loads GEO packages, and reads the GSE119526 beta matrix.
```{r gse119526-read-data}
rm(list = ls())
options(stringsAsFactors = F)

require(GEOquery)
require(Biobase)
eset <- getGEO("GSE119526", destdir = './', AnnotGPL = T, getGPL = F)
beta.m <- exprs(eset[[1]])
dim(beta.m)
```
The matrix dimensions show the number of CpG probes and samples available before filtering.
#### This chunk constructs the GSE119526 phenotype table and classifies samples as Normal or Tumor.
```{r gse119526-build-phenotype}
pD.all <- pData(eset[[1]])
pD <- pD.all[, c(1, 2)]
names(pD) <- c("group", "ID")
pD$group <- ifelse(
  grepl("normal", pD$group, ignore.case = T), "Normal",
  "Tumor")
# Check if the grouping is correct.
check_df <- data.frame(Title = pD.all$title, Group = pD$group)
head(check_df)
# head(check_df, 10)
table(pD$group)
```
The displayed sample records and group counts should be inspected to confirm the grouping.
#### This chunk checks sample correspondence, orders the phenotype table by group, and reorders the beta matrix using the same sample order.
```{r gse119526-align-samples}
stopifnot(setequal(colnames(beta.m), pD$ID))
stopifnot(identical(colnames(beta.m), pD$ID))
# Set group order: Normal first, Tumor second
pD$group <- factor(pD$group, levels = c("Normal", "Tumor"))

# Get sorted sample order
sample_order <- order(pD$group)

# Sort phenotype table
pD <- pD[sample_order, , drop = FALSE]

# Reorder beta matrix columns according to sorted pD
beta.m <- beta.m[, pD$ID, drop = FALSE]

stopifnot(identical(colnames(beta.m), pD$ID))
# Check group order
table(pD$group)
```
The identity check should confirm that beta-matrix columns and phenotype IDs are aligned.
#### This chunk applies ChAMP filtering and saves the processed GSE119526 object.
```{r gse119526-champ-filter}
myLoad <- champ.filter(beta = beta.m, pd = pD, arraytype = "EPIC")
#save as .Rdata
save(myLoad, file = 'step0.3-GSE119526_output.Rdata')
```


# 🍁🍁CpG-level analysis across the three discovery datasets
The overall strategy is:

1. Analyse GSE199057, GSE149282, and GSE119526 separately using limma.
2. Extract the CpG-level effect estimate and standard error from each dataset.
3. Combine the three effects using inverse-variance fixed-effect meta-analysis.
4. Use the resulting CpG-level meta-analysis signal for DMR calling (DMRcate).
## Load the three processed datasets
This chunk loads the three saved ChAMP objects and assigns their beta matrices and phenotype tables to dataset-specific objects.
```{r load-three-discovery-datasets}
rm(list = ls())
options(timeout = 100000) 
options(scipen = 20)


load("step0.1-GSE199057_output.Rdata")
g_GSE199057 <- myLoad$beta
g_GSE199057_group <- myLoad$pd
rm(myLoad)
g1 <- g_GSE199057
pd1 <- g_GSE199057_group

load("step0.2-GSE149282_output.Rdata")
g_GSE149282 <- myLoad$beta
g_GSE149282_group <- myLoad$pd
rm(myLoad)
g2 <-g_GSE149282
pd2 <- g_GSE149282_group

load("step0.3-GSE119526_output.Rdata")
g_GSE119526 <- myLoad$beta
g_GSE119526_group <- myLoad$pd
rm(myLoad)
g3 <-g_GSE119526
pd3 <- g_GSE119526_group

dim(g1)
dim(g2)
dim(g3)

dim(pd1)
dim(pd2)
dim(pd3)

colnames(pd1)
colnames(pd2)
colnames(pd3)

head(pd1);tail(pd1)
head(pd2);tail(pd2)
head(pd3);tail(pd3)
```
The reported dimensions and sample records allow the three inputs to be checked before combination.
#### This chunk restricts all three beta matrices to the CpG probes shared by every discovery dataset.
```{r restrict-to-common-cpgs}
# unify CpG space
common_cpg <- Reduce(intersect, list(
    rownames(g1),
    rownames(g2),
    rownames(g3)
  ))

g1 <- g1[common_cpg, ]
g2 <- g2[common_cpg, ]
g3 <- g3[common_cpg, ]

all(colnames(g1) == pd1$ID)
all(colnames(g2) == pd2$ID)
all(colnames(g3) == pd3$ID)
```
After subsetting, each dataset contains the same CpG feature space for the meta-analysis.
#### This chunk creates consistently ordered Normal and Tumor factors for the three datasets.
```{r define-discovery-groups}
# build phenotype
group1 <- factor(pd1$group, levels = c("Normal", "Tumor"))
group2 <- factor(pd2$group, levels = c("Normal", "Tumor"))
group3 <- factor(pd3$group, levels = c("Normal", "Tumor"))
```
These factors define the direction of all subsequent comparisons as Tumor minus Normal.
##  Run limma separately in each dataset
#### This chunk defines the reusable run_limma() function. It converts beta values to M values, fits the Tumor-versus-Normal contrast, extracts the limma statistics, calculates standard errors, and retains delta beta for biological interpretation.
```{r define-limma-function}
## LIMMA per dataset
run_limma <- function(beta, group){
  beta <- as.matrix(beta)
  # Fixed effect direction: Tumor - Normal
  group <- factor(group, levels = c("Normal", "Tumor"))
  stopifnot(ncol(beta) == length(group))
  stopifnot(!any(is.na(group)))
  # Prevent beta values of 0 or 1 from producing infinite M values
  beta <- pmin(pmax(beta, 1e-6), 1 - 1e-6)
  # Convert beta values to M values
  mval <- log2(beta / (1 - beta))
  # Use a no-intercept design to explicitly define Tumor - Normal
  design <- model.matrix(~ 0 + group)
  colnames(design) <- levels(group)
  contrast.matrix <- makeContrasts(
    Tumor_vs_Normal = Tumor - Normal,
    levels = design
  )
  fit <- lmFit(mval, design)
  fit <- contrasts.fit(fit, contrast.matrix)
  fit <- eBayes(fit)
  res <- topTable(
    fit,
    coef = "Tumor_vs_Normal",
    number = Inf,
    sort.by = "none"
  )
  # CpG ID
  res$CpG <- rownames(res)
  # Manually calculate standard error
  se_vec <- sqrt(fit$s2.post) * fit$stdev.unscaled[, "Tumor_vs_Normal"]
  names(se_vec) <- rownames(fit$coefficients)
  res$SE <- se_vec[res$CpG]
  # Retain delta beta for biological interpretation
  res$deltaBeta <- rowMeans(beta[, group == "Tumor", drop = FALSE]) -
    rowMeans(beta[, group == "Normal", drop = FALSE])
  return(res)
}
```
The function returns one CpG-level result table for each dataset. Positive delta beta values indicate higher methylation in Tumor samples.
#### This chunk applies the same limma workflow independently to all three discovery datasets.
```{r run-limma-three-datasets}
res1 <- run_limma(g1, group1)
res2 <- run_limma(g2, group2)
res3 <- run_limma(g3, group3)
```
The objects res1, res2, and res3 contain the CpG-level results used in the meta-analysis.
#### This chunk counts significant hypermethylated and hypomethylated CpGs in each individual dataset.
```{r count-significant-cpgs}
# check no significant probes
nrow(subset(res1,logFC>0 & adj.P.Val<0.05))
nrow(subset(res1,logFC<0 & adj.P.Val<0.05))

nrow(subset(res2,logFC>0 & adj.P.Val<0.05))
nrow(subset(res2,logFC<0 & adj.P.Val<0.05))

nrow(subset(res3,logFC>0 & adj.P.Val<0.05))
nrow(subset(res3,logFC<0 & adj.P.Val<0.05))
```
These counts provide a basic summary of the direction and scale of differential methylation.
#### This chunk calculates pairwise correlations between the three sets of CpG log-fold changes.
```{r compare-effect-correlations}
# check similarity between
cor(res1$logFC,res2$logFC,use="complete.obs")
cor(res1$logFC,res3$logFC,use="complete.obs")
cor(res2$logFC,res3$logFC,use="complete.obs")
```
Higher positive correlations indicate stronger agreement in methylation effects between datasets.
## ixed-effect CpG meta-analysis
This chunk assembles the three dataset-specific effect estimates and standard errors into one CpG-level meta-analysis table.
```{r assemble-meta-analysis-input}
# CpG meta-analysis
meta_df <- data.frame(
  CpG = res1$CpG,

  beta1 = res1$logFC,
  se1   = res1$SE,

  beta2 = res2$logFC,
  se2   = res2$SE,

  beta3 = res3$logFC,
  se3   = res3$SE
)
## meta-analysis per CpG
## fixed-effect meta-analysis per CpG
```
Each row now contains the information required to combine the same CpG across all three datasets.

This chunk performs inverse-variance fixed-effect meta-analysis and calculates the pooled effect, standard error, z statistic, two-sided p value, and BH-adjusted FDR.
```{r fixed-effect-meta-analysis}
b1 <- meta_df$beta1
b2 <- meta_df$beta2
b3 <- meta_df$beta3

s1 <- meta_df$se1
s2 <- meta_df$se2
s3 <- meta_df$se3

## inverse-variance weights
w1 <- 1 / s1^2
w2 <- 1 / s2^2
w3 <- 1 / s3^2

## fixed-effect pooled beta
beta_meta <- (w1 * b1 + w2 * b2 + w3 * b3) / (w1 + w2 + w3)

## pooled standard error
se_meta <- sqrt(1 / (w1 + w2 + w3))

## z statistic
z_meta <- beta_meta / se_meta

## two-sided p value
p_meta <- 2 * pnorm(abs(z_meta), lower.tail = FALSE)

## output
meta_res_fixed <- data.frame(
  CpG = meta_df$CpG,
  beta = beta_meta,
  se = se_meta,
  z = z_meta,
  p = p_meta
)

## FDR correction
meta_res_fixed$FDR <- p.adjust(meta_res_fixed$p, method = "BH")
```
The object meta_res_fixed contains the final CpG-level meta-analysis statistics.
#🍁🍁 DMR calling using DMRcate from the meta-analysis CpG signal
This section annotates the meta-analysis CpGs, calculates the mean discovery delta beta, constructs the CpGannotated object, calls DMRs, and filters for candidate hypermethylated regions.
## Load packages and annotation
This chunk loads the packages required for genomic annotation, DMR calling, and table display.
```{r dmrcate-load-packages}
library(DMRcate)
library(GenomicRanges)
library(data.table)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(kableExtra)
```
The EPIC hg19 annotation package is used consistently with the discovery datasets.

This chunk creates a CpG annotation table containing probe ID, chromosome, genomic position, gene annotation, and CpG-island relation.
```{r build-epic-annotation-table}
ann <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)

ann_df <- data.frame(
  CpG = rownames(ann),
  chr = as.character(ann$chr),
  pos = as.integer(ann$pos),
  gene = as.character(ann$UCSC_RefGene_Name),
  relation_to_island = as.character(ann$Relation_to_Island),
  stringsAsFactors = FALSE
)
```
The annotation table is used to link the CpG-level meta-analysis results to genomic coordinates.
## Calculate discovery delta beta and prepare DMRcate input

This chunk merges the three dataset-specific delta-beta estimates and calculates their mean for each CpG:
-deltaBeta direction: Tumor - Normal
-Positive value means hypermethylated in Tumor
```{r calculate-mean-discovery-delta-beta}
delta_df <- Reduce(function(x, y) merge(x, y, by = "CpG", all = FALSE), list(
  data.frame(CpG = res1$CpG, deltaBeta1 = res1$deltaBeta),
  data.frame(CpG = res2$CpG, deltaBeta2 = res2$deltaBeta),
  data.frame(CpG = res3$CpG, deltaBeta3 = res3$deltaBeta)
))

delta_df$mean_deltaBeta <- rowMeans(
  delta_df[, c("deltaBeta1", "deltaBeta2", "deltaBeta3")],
  na.rm = TRUE
)
```
Positive mean_deltaBeta values indicate higher average methylation in Tumor samples across the discovery datasets.

This chunk merges the meta-analysis statistics, EPIC annotation, and mean delta beta; removes invalid records; retains autosomal CpGs; and orders them by genomic position.
meta_res_fixed contains:
CpG, beta, se, z, p, FDR
beta here is the meta-analysis effect size in M-value/logFC scale.
mean_deltaBeta is retained for biological interpretation and filtering.

```{r merge-meta-annotation-delta}
dat <- merge(meta_res_fixed, ann_df, by = "CpG")
dat <- merge(dat, delta_df[, c("CpG", "mean_deltaBeta")], by = "CpG")

dat <- dat[
  is.finite(dat$beta) &
    is.finite(dat$se) &
    is.finite(dat$z) &
    is.finite(dat$p) &
    is.finite(dat$FDR) &
    is.finite(dat$mean_deltaBeta) &
    !is.na(dat$chr) &
    !is.na(dat$pos),]
dat <- dat[dat$chr %in% paste0("chr", 1:22), ]

## Sort by chromosome and position
dat$chr_order <- match(dat$chr, paste0("chr", 1:22))
dat <- dat[order(dat$chr_order, dat$pos), ]

head(dat, 10) |>
  kbl(caption = "CpG data used for DMRcate") |>
  kable_paper("hover", full_width = FALSE)
```
The displayed table is the final CpG-level dataset supplied to DMRcate.

This chunk converts the CpG-level meta-analysis results into the CpGannotated object required by DMRcate.
DMRcate normally uses cpg.annotate() from M-values + design.
Here, because we already performed CpG-level meta-analysis,
we manually construct a CpGannotated object using meta statistics.
Required fields:
stat    = meta z statistic
rawpval = meta p value
diff    = mean delta beta, Tumor - Normal
ind.fdr = meta FDR
is.sig  = whether the CpG is significant at CpG level
```{r build-cpgannotated-object}
gr <- GRanges(
  seqnames = dat$chr,
  ranges = IRanges(start = dat$pos, end = dat$pos),
  stat = dat$z,
  rawpval = dat$p,
  diff = dat$mean_deltaBeta,
  ind.fdr = dat$FDR,
  is.sig = dat$FDR < 0.05
)

names(gr) <- dat$CpG

myannotation <- new("CpGannotated", ranges = gr)

myannotation
```
The object stores the meta-analysis z statistic, p value, FDR, and mean discovery delta beta for each CpG.
## Call and extract DMRs
This chunk runs DMRcate using the existing parameters `lambda = 1000`, `C = 2`, `pcutoff = "fdr"`, and `min.cpgs = 6`.
lambda = 1000: CpGs separated by >= 1000 bp tend to be split into different regions.
C = 2:Recommended commonly used setting for array data
min.cpgs = 6:previous criterion was no.cpgs > 5, so here this becomes at least 6 CpGs.
pcutoff = "fdr": Use the FDR indexing stored in the CpGannotated object.
```{r run-dmrcate}
dmrcoutput <- dmrcate(
  myannotation,
  lambda = 1000,
  C = 2,
  pcutoff = "fdr",
  min.cpgs = 6)

dmrcoutput
```

The object `dmrcoutput` contains the smoothed regional methylation results.

This chunk extracts hg19 genomic ranges from the DMRcate result.
```{r extract-dmr-ranges}
dmr_ranges <- extractRanges(dmrcoutput, genome = "hg19")

dmr_ranges
```
The resulting `dmr_ranges` object contains the genomic intervals identified as DMRs.

This chunk converts the genomic ranges to a data frame, creates a readable coordinate column, and reorders the main result columns.
```{r format-dmrcate-results}
## Convert DMRcate result to data frame
dmr_dmrcate <- as.data.frame(dmr_ranges)

## Make coordinates easier to read
dmr_dmrcate$coord <- paste0(
  dmr_dmrcate$seqnames, ":",
  dmr_dmrcate$start, "-",
  dmr_dmrcate$end
)

## Reorder columns
first_cols <- c("coord", "seqnames", "start", "end", "width", "no.cpgs")
other_cols <- setdiff(colnames(dmr_dmrcate), first_cols)
dmr_dmrcate <- dmr_dmrcate[, c(first_cols, other_cols)]

head(dmr_dmrcate, 10) |>
  kbl(caption = "DMRcate DMR result") |>
  kable_paper("hover", full_width = FALSE)

dim(dmr_dmrcate)
```
The displayed dimensions report the total number of DMRs called before biological filtering.
## Select hypermethylated candidate DMRs and save results

This chunk retains DMRs containing at least six CpGs and having a mean Tumor-minus-Normal delta beta greater than 0.20, then orders the retained regions.
-we set diff = mean_deltaBeta above,
DMRcate's meandiff should represent mean Tumor - Normal delta beta across CpGs in each DMR.
-Keep the same biological filter as before:
1. at least 6 CpGs
2. mean delta beta > 0.2
3. hypermethylated in Tumor

```{r filter-hypermethylated-dmrs}
dmr_hyper_dmrcate <- dmr_dmrcate[
  dmr_dmrcate$no.cpgs >= 6 &
    dmr_dmrcate$meandiff > 0.2,
]

## Sort by strongest regional evidence, then largest methylation difference
## DMRcate output usually contains min_smoothed_fdr.
if ("min_smoothed_fdr" %in% colnames(dmr_hyper_dmrcate)) {
  dmr_hyper_dmrcate <- dmr_hyper_dmrcate[
    order(dmr_hyper_dmrcate$min_smoothed_fdr, -dmr_hyper_dmrcate$meandiff),
  ]
} else {
  dmr_hyper_dmrcate <- dmr_hyper_dmrcate[
    order(-dmr_hyper_dmrcate$meandiff),
  ]
}

head(dmr_hyper_dmrcate, 10) |>
  kbl(caption = "Hypermethylated DMRs called by DMRcate") |>
  kable_paper("hover", full_width = FALSE)

dim(dmr_hyper_dmrcate)
```
The resulting object `dmr_hyper_dmrcate` contains the discovery-stage candidate hypermethylated DMRs.

This chunk saves the complete R objects and exports the filtered hypermethylated DMR table as CSV.
```{r save-discovery-dmr-results}
save(dmr_hyper_dmrcate,
  dmr_dmrcate,
  dmr_ranges,
  dmrcoutput,
  myannotation,
  file = "dmr_hyper_dmrcate_meta.Rdata")

write.csv(dmr_hyper_dmrcate,
  file = "dmr_hyper_dmrcate_meta.csv",
  row.names = FALSE)
```

The files `dmr_hyper_dmrcate_meta.Rdata` and `dmr_hyper_dmrcate_meta.csv` are the main outputs of Step 1.

## Session information

The session information below records the R version, operating system, and loaded package versions used to generate this report.

```{r sessioninfo}

sessionInfo()

```
