---
title: "SHAP-based prioritisation of colorectal cancer DMRs (nsim = 1000)"
author: "YC"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This report prioritises the final 110 DMR candidates for wet-lab validation using the radial SVM model and the TCGA full cohort. 
- SHAP values are calculated for all 110 DMRs before candidate ranking.
- The final priority score combines 60% scaled SHAP importance and 40% scaled discovery-stage regional delta beta.

The discovery delta beta used here is matched from the Step 1 DMRcate `meandiff` column by exact chromosome, start, and end coordinates. The previous value that had been copied from the validation-stage `mean_deltaBeta` is retained only for auditing.

## Analysis workflow

1. Load the trained radial SVM model and TCGA full-cohort feature matrix.
2. Restore the true discovery-stage delta beta for all 110 DMRs.
3. Calculate SHAP values with `nsim = 1000`.
4. Calculate TCGA Tumor-minus-Normal delta beta as supporting evidence within the TCGA prioritisation cohort.
5. Calculate the combined priority score and select the final top 10 DMRs.
6. Generate summary tables and visualisations.

# TCGA full-cohort SHAP prioritisation

The analysis begins from a clean R environment, sets numerical display options, and fixes the random seed.

```{r initialise-analysis}
rm(list = ls())

options(stringsAsFactors = FALSE)
options(scipen = 20)

set.seed(123)
```

Using the same seed makes the Monte Carlo SHAP calculation reproducible for the specified software environment.

This chunk loads the packages used for data handling, model prediction, SHAP estimation, statistical evaluation, and plotting.

```{r load-packages}
library(data.table)
library(caret)
library(fastshap)
library(ggplot2)
library(ggbeeswarm)
library(pROC)
library(kernlab)
```

All subsequent analysis steps depend on these packages being loaded successfully.

This chunk defines the four required input files and a new output directory so that the corrected results do not overwrite earlier analyses.

```{r define-inputs}
## Expected analysis dimensions
expected_n_dmrs <- 110L
expected_top_n <- 10L

## Trained ML models from GSE101764
training_model_file <- "GSE101764_ML_training_3models_110DMRs.Rdata"

## TCGA full-cohort testing results
tcga_testing_file <- "TCGA_full_testing_results_3models.Rdata"

## Final DMRs after validation and background filtering
final_dmr_file <- "final_DMRs_after_GSE196696_whole_blood_filter.csv"

## Step 1 DMRcate discovery results
discovery_dmr_file <- "dmr_hyper_dmrcate_meta.csv"

## Corrected outputs are written separately
outdir <- "step5.SHAP_selection_nsim1000"

required_input_files <- c(
  training_model_file,
  tcga_testing_file,
  final_dmr_file,
  discovery_dmr_file
)

missing_input_files <- required_input_files[
  !file.exists(required_input_files)
]

if (length(missing_input_files) > 0) {
  stop(
    "Missing required input file(s): ",
    paste(missing_input_files, collapse = ", ")
  )
}

if (!dir.exists(outdir)) {
  dir.create(outdir, recursive = TRUE)
}
```

## Load analysis objects

This chunk loads the trained machine-learning objects,TCGA full-cohort testing dataframe, the final 110-DMR table, and the Step 1 DMRcate discovery table.

```{r load-analysis-objects}
load(training_model_file)

## Load the training RData file and record its object names
loaded_training_objects <- load(training_model_file)

cat(
  "Objects loaded from the training RData file:\n"
)

print(loaded_training_objects)

required_training_objects <- c(
  "model_svm",
  "feature_name_map",
  "ml_df"
)

## Check object names directly against the output of load()
missing_training_objects <- setdiff(
  required_training_objects,
  loaded_training_objects
)

if (length(missing_training_objects) > 0) {
  stop(
    "Missing object(s) in the training RData file: ",
    paste(missing_training_objects, collapse = ", "),
    "\nObjects actually found: ",
    paste(loaded_training_objects, collapse = ", ")
  )
}

cat(
  "All required training objects were loaded successfully.\n"
)
loaded_tcga_objects <- load(tcga_testing_file)

cat(
  "Objects loaded from the TCGA testing RData file:\n"
)

print(loaded_tcga_objects)

if (!"test_df_full" %in% loaded_tcga_objects) {
  stop(
    "test_df_full was not found in the TCGA testing RData file.",
    "\nObjects actually found: ",
    paste(loaded_tcga_objects, collapse = ", ")
  )
}

test_df_full <- as.data.frame(test_df_full)
final_dmrs <- fread(final_dmr_file)
discovery_dmrs <- fread(discovery_dmr_file)

dim(test_df_full)
table(test_df_full$group)
dim(final_dmrs)
dim(discovery_dmrs)
```

The dimensions and group counts provide a basic check that the intended datasets were loaded.

## Restore the true discovery-stage delta beta


This chunk verifies the expected row counts, grouping variable, and required columns before any values are renamed or matched.

```{r validate-input-structure}
if (nrow(final_dmrs) != expected_n_dmrs) {
  stop(
    "Expected ", expected_n_dmrs,
    " final DMRs, but found ", nrow(final_dmrs), "."
  )
}

required_final_columns <- c(
  "DMR_ID", "chr", "start", "end",
  "discovery_deltaBeta", "mean_deltaBeta"
)

required_discovery_columns <- c(
  "coord", "seqnames", "start", "end",
  "meandiff", "no.cpgs"
)

missing_final_columns <- setdiff(
  required_final_columns,
  names(final_dmrs)
)

missing_discovery_columns <- setdiff(
  required_discovery_columns,
  names(discovery_dmrs)
)

if (length(missing_final_columns) > 0) {
  stop(
    "Missing final-DMR column(s): ",
    paste(missing_final_columns, collapse = ", ")
  )
}

if (length(missing_discovery_columns) > 0) {
  stop(
    "Missing discovery-DMR column(s): ",
    paste(missing_discovery_columns, collapse = ", ")
  )
}

if (!"group" %in% names(test_df_full)) {
  stop("The TCGA testing dataframe does not contain a group column.")
}
```

The report stops here if the files are incompatible with the corrected Step 5 workflow.


This chunk standardises chromosome labels, preserves the original row order, retains the legacy delta-beta value, explicitly labels the validation mean delta beta, and constructs coordinate keys.

```{r prepare-coordinate-keys}
fix_chr <- function(x) {
  x <- as.character(x)
  x <- gsub("^chr", "", x, ignore.case = TRUE)
  paste0("chr", x)
}

if (anyNA(final_dmrs[, .(chr, start, end)])) {
  stop("Missing genomic coordinates were found in final_dmrs.")
}

if (anyNA(discovery_dmrs[, .(seqnames, start, end)])) {
  stop("Missing genomic coordinates were found in discovery_dmrs.")
}

## Preserve the original order for restoration after merge
final_dmrs[, original_row_order := seq_len(.N)]

## Preserve the old, incorrectly labelled value for auditing
if ("legacy_discovery_deltaBeta" %in% names(final_dmrs)) {
  stop("legacy_discovery_deltaBeta already exists in the input table.")
}

setnames(
  final_dmrs,
  old = "discovery_deltaBeta",
  new = "legacy_discovery_deltaBeta"
)

## Give the Step 2 validation value an explicit name
if ("validation_mean_deltaBeta" %in% names(final_dmrs)) {
  stop("validation_mean_deltaBeta already exists in the input table.")
}

setnames(
  final_dmrs,
  old = "mean_deltaBeta",
  new = "validation_mean_deltaBeta"
)

final_dmrs[, coordinate_key := paste0(
  fix_chr(chr), ":", as.integer(start), "-", as.integer(end)
)]

discovery_dmrs[, coordinate_key := paste0(
  fix_chr(seqnames), ":", as.integer(start), "-", as.integer(end)
)]
```

The coordinate key provides the stable link between the later DMR identifiers and the original Step 1 DMRcate regions.

This chunk checks coordinate uniqueness, builds the discovery mapping table, and merges the Step 1 `meandiff` and `no.cpgs` values into all final DMRs.

```{r build-discovery-map}
if (anyDuplicated(final_dmrs$DMR_ID) > 0) {
  stop("Duplicated DMR_ID values were found in the final DMR table.")
}

if (anyDuplicated(final_dmrs$coordinate_key) > 0) {
  stop("Duplicated coordinates were found in the final DMR table.")
}

if (anyDuplicated(discovery_dmrs$coordinate_key) > 0) {
  stop("Duplicated coordinates were found in the Step 1 DMRcate table.")
}

discovery_delta_map <- discovery_dmrs[
  ,
  .(
    coordinate_key,
    discovery_coord = coord,
    discovery_deltaBeta = as.numeric(meandiff),
    discovery_no_cpgs = as.integer(no.cpgs)
  )
]

if (anyNA(discovery_delta_map$discovery_deltaBeta)) {
  stop("Missing meandiff values were found in the Step 1 DMRcate table.")
}

if (anyNA(discovery_delta_map$discovery_no_cpgs)) {
  stop("Missing no.cpgs values were found in the Step 1 DMRcate table.")
}
```

The mapping table contains one unique discovery coordinate, its true regional delta beta, and its discovery-stage probe count.

This chunk performs the left join and confirms that all 110 final DMRs were matched exactly once.

```{r merge-and-check-discovery}
n_final_before_merge <- nrow(final_dmrs)

final_dmrs <- merge(
  final_dmrs,
  discovery_delta_map,
  by = "coordinate_key",
  all.x = TRUE,
  sort = FALSE
)

if (nrow(final_dmrs) != n_final_before_merge) {
  stop("The discovery merge changed the number of final DMR rows.")
}

setorder(final_dmrs, original_row_order)
final_dmrs[, original_row_order := NULL]

unmatched_dmrs <- final_dmrs[
  is.na(discovery_deltaBeta) |
    is.na(discovery_no_cpgs)
]

if (nrow(unmatched_dmrs) > 0) {
  print(
    unmatched_dmrs[
      ,
      .(DMR_ID, chr, start, end, coordinate_key)
    ]
  )
  stop("Some final DMRs could not be matched to Step 1.")
}

if (any(final_dmrs$discovery_deltaBeta <= 0.20)) {
  stop("At least one matched discovery DMR has meandiff <= 0.20.")
}

if (any(final_dmrs$discovery_no_cpgs < 6L)) {
  stop("At least one matched discovery DMR contains fewer than six probes.")
}

cat("All", nrow(final_dmrs), "final DMRs matched to Step 1.
")
```

All final candidates must retain the Step 1 hypermethylation and minimum-CpG criteria.

This chunk checks optional upstream filtering fields when they are available in the final DMR table.

```{r check-upstream-filter-status}
assert_all_true <- function(x, label) {
  x <- as.logical(x)

  if (anyNA(x) || !all(x)) {
    stop(label, " is not TRUE for every final DMR.")
  }
}

if ("background_pass_all" %in% names(final_dmrs)) {
  assert_all_true(
    final_dmrs$background_pass_all,
    "background_pass_all"
  )
}

if ("GSE196696_pass" %in% names(final_dmrs)) {
  assert_all_true(
    final_dmrs$GSE196696_pass,
    "GSE196696_pass"
  )
}

if ("validated_datasets_p005" %in% names(final_dmrs)) {
  validation_count <- as.numeric(
    final_dmrs$validated_datasets_p005
  )

  if (anyNA(validation_count) || any(validation_count < 3)) {
    stop("Some final DMRs were validated in fewer than three datasets.")
  }
}
```

These conditional checks confirm that the final table still satisfies its stated upstream validation and background-filtering criteria.

This chunk exports an audit table and proves that the corrected discovery value is no longer simply the validation mean copied under another name.

```{r audit-delta-sources}
delta_source_audit <- final_dmrs[
  ,
  .(
    DMR_ID,
    coordinate_key,
    validation_mean_deltaBeta,
    legacy_discovery_deltaBeta,
    discovery_deltaBeta,
    discovery_no_cpgs,
    legacy_minus_validation =
      legacy_discovery_deltaBeta - validation_mean_deltaBeta,
    corrected_minus_validation =
      discovery_deltaBeta - validation_mean_deltaBeta
  )
]

legacy_matches_validation <- isTRUE(
  all.equal(
    delta_source_audit$legacy_discovery_deltaBeta,
    delta_source_audit$validation_mean_deltaBeta,
    tolerance = 1e-8
  )
)

corrected_matches_validation <- isTRUE(
  all.equal(
    delta_source_audit$discovery_deltaBeta,
    delta_source_audit$validation_mean_deltaBeta,
    tolerance = 1e-8
  )
)

cat("Legacy value equals validation mean:", legacy_matches_validation, "
")
cat("Corrected discovery value equals validation mean:",
    corrected_matches_validation, "
")

if (corrected_matches_validation) {
  stop("The discovery-delta correction did not change the scoring input.")
}

fwrite(
  delta_source_audit,
  file.path(outdir, "all_110_DMRs_deltaBeta_source_audit_nsim1000.csv")
)
```

After the merge, `discovery_deltaBeta` represents the true Step 1 regional delta beta and `discovery_no_cpgs` records the corresponding discovery probe count.

## Select the final model and prepare TCGA predictors

The radial SVM is selected as the final model, and the exact predictor names used during training are extracted.

```{r select-svm-features}
final_model <- model_svm
final_model_name <- "Radial SVM"

if (is.null(final_model$trainingData)) {
  stop("The radial SVM model does not contain trainingData.")
}

print(final_model$bestTune)

features_svm <- setdiff(
  colnames(final_model$trainingData),
  ".outcome"
)

if (length(features_svm) != expected_n_dmrs) {
  stop(
    "Expected ", expected_n_dmrs,
    " SVM predictors, but found ", length(features_svm), "."
  )
}

if (anyDuplicated(features_svm) > 0) {
  stop("Duplicated predictor names were found in the radial SVM.")
}
```

This chunk confirms that the model predictors, final DMR identifiers, and TCGA feature columns represent the same 110 DMRs.

```{r check-model-feature-identity}
missing_from_final_dmrs <- setdiff(
  features_svm,
  final_dmrs$DMR_ID
)

extra_in_final_dmrs <- setdiff(
  final_dmrs$DMR_ID,
  features_svm
)

missing_from_tcga <- setdiff(
  features_svm,
  colnames(test_df_full)
)

if (length(missing_from_final_dmrs) > 0 ||
    length(extra_in_final_dmrs) > 0) {
  print(missing_from_final_dmrs)
  print(extra_in_final_dmrs)
  stop("The SVM predictors and final DMR_ID values are not identical.")
}

if (length(missing_from_tcga) > 0) {
  stop(
    "SVM features missing from TCGA: ",
    paste(missing_from_tcga, collapse = ", ")
  )
}

cat("The 110 SVM predictors exactly match the final DMR IDs.
")
```

All extracted features must be present in the TCGA testing dataframe before SHAP values can be calculated.

This chunk creates the TCGA feature matrix used for SHAP explanation and standardises the group order for downstream summaries.

```{r prepare-tcga-groups}
test_df_full$group <- factor(
  as.character(test_df_full$group),
  levels = c("Normal", "Tumor")
)

if (anyNA(test_df_full$group)) {
  stop("TCGA group contains missing or unsupported labels.")
}

group_counts <- table(test_df_full$group)

if (any(group_counts == 0)) {
  stop("Both Normal and Tumor samples are required.")
}

if (is.null(rownames(test_df_full)) ||
    any(rownames(test_df_full) == "")) {
  rownames(test_df_full) <- paste0(
    "Sample_", seq_len(nrow(test_df_full))
  )
}

if (anyDuplicated(rownames(test_df_full)) > 0) {
  stop("Duplicated TCGA sample identifiers were found.")
}

print(group_counts)
```

This chunk creates the predictor matrix and checks that every methylation feature is numeric, finite, and within the expected beta-value range.

```{r prepare-shap-data}
x_explain <- as.data.frame(
  test_df_full[, features_svm, drop = FALSE]
)

numeric_features <- vapply(
  x_explain,
  is.numeric,
  logical(1)
)

if (!all(numeric_features)) {
  stop("At least one TCGA DMR feature is not numeric.")
}

x_matrix <- as.matrix(x_explain)

if (any(!is.finite(x_matrix))) {
  stop("TCGA DMR features contain NA, NaN, or infinite values.")
}

if (any(x_matrix < -1e-8 | x_matrix > 1 + 1e-8)) {
  stop("TCGA DMR beta values fall outside the expected 0-1 range.")
}

print(dim(x_explain))
```

The resulting matrix should contain TCGA samples in rows and all radial-SVM DMR predictors in columns.

SHAP requires a prediction wrapper that returns the model-predicted probability of Tumor for each sample.

```{r define-prediction-wrapper}
pred_wrapper_svm <- function(object, newdata) {newdata <- as.data.frame(newdata)
  newdata <- newdata[, features_svm, drop = FALSE]
  
  predict(object,newdata = newdata,type = "prob")[, "Tumor"]}
```

Positive SHAP values will therefore indicate contributions that increase the predicted Tumor probability.

## Calculate SHAP values for all 110 DMRs

This chunk calculates approximate SHAP values for every TCGA sample and every SVM feature using 1000 Monte Carlo simulations.

```{r run-shap-nsim1000}
nsim_shap <- 1000L
set.seed(123)

shap_svm <- fastshap::explain(
  object = final_model,
  X = x_explain,
  pred_wrapper = pred_wrapper_svm,
  nsim = nsim_shap,
  adjust = TRUE
)

shap_svm <- as.data.frame(shap_svm)

save(
  shap_svm,
  file = file.path(
    outdir,
    "radial_SVM_SHAP_values_all_110_DMRs_TCGA_full__nsim1000.Rdata"
  )
)
```

This chunk confirms that the SHAP matrix has the expected samples, features, names, and finite values.

```{r validate-shap-output}
if (nrow(shap_svm) != nrow(x_explain)) {
  stop("The SHAP matrix has an unexpected number of sample rows.")
}

if (ncol(shap_svm) != length(features_svm)) {
  stop("The SHAP matrix has an unexpected number of feature columns.")
}

if (!identical(colnames(shap_svm), features_svm)) {
  stop("The SHAP feature order differs from the SVM feature order.")
}

if (any(!is.finite(as.matrix(shap_svm)))) {
  stop("The SHAP matrix contains NA, NaN, or infinite values.")
}

cat(
  "Validated SHAP matrix:",
  nrow(shap_svm), "samples x",
  ncol(shap_svm), "DMRs.
"
)
```

The saved matrix contains sample-level SHAP contributions and is the basis for all subsequent SHAP summaries.

This chunk calculates the mean absolute SHAP value and mean signed SHAP value for each DMR, then ranks all 110 features.

```{r summarise-shap-importance}
shap_importance <- data.table(
  Feature = colnames(shap_svm),
  mean_abs_SHAP = colMeans(abs(shap_svm)),
  mean_SHAP = colMeans(shap_svm)
)

if (nrow(shap_importance) != expected_n_dmrs) {
  stop("The SHAP importance table does not contain 110 DMRs.")
}

if (anyDuplicated(shap_importance$Feature) > 0) {
  stop("Duplicated Feature values were found in SHAP importance.")
}

if (any(!is.finite(shap_importance$mean_abs_SHAP))) {
  stop("Non-finite mean absolute SHAP values were found.")
}

setorder(shap_importance, -mean_abs_SHAP)
shap_importance[, SHAP_rank := seq_len(.N)]

fwrite(
  shap_importance,
  file.path(
    outdir,
    "radial_SVM_SHAP_importance_all_110_DMRs_nsim1000.csv"
  )
)

head(shap_importance, 10)
```

A higher mean absolute SHAP value indicates a larger average contribution to model predictions.

The SHAP ranking is merged with the corrected final DMR table using the model feature name and `DMR_ID`.

```{r merge-shap-with-dmr-evidence}
final_priority_table <- merge(
  shap_importance,
  final_dmrs,
  by.x = "Feature",
  by.y = "DMR_ID",
  all.x = TRUE,
  sort = FALSE
)

setorder(final_priority_table, SHAP_rank)

if (nrow(final_priority_table) != expected_n_dmrs) {
  stop("The SHAP-to-DMR merge changed the expected row count.")
}

if (anyDuplicated(final_priority_table$Feature) > 0) {
  stop("Duplicated features were created by the SHAP-to-DMR merge.")
}

if (anyNA(final_priority_table$discovery_deltaBeta) ||
    anyNA(final_priority_table$discovery_no_cpgs)) {
  stop("Discovery information is missing after the SHAP-to-DMR merge.")
}

dim(final_priority_table)
head(final_priority_table)
```

The merged table now contains model importance, discovery delta beta, validation evidence, and background-filtering information.

## Calculate TCGA delta beta

This function calculates the TCGA mean beta value in Normal and Tumor samples, their difference, and a Wilcoxon p value for one DMR.

```{r define-tcga-delta-function}
calc_tcga_delta <- function(feature) {
  
  values <- as.numeric(test_df_full[[feature]])
  group <- test_df_full$group
  
  normal_values <- values[group == "Normal"]
  tumor_values <- values[group == "Tumor"]
  
  mean_normal <- mean(normal_values, na.rm = TRUE)
  mean_tumor <- mean(tumor_values, na.rm = TRUE)
  delta_beta <- mean_tumor - mean_normal
  
  p_value <- tryCatch(wilcox.test(tumor_values, normal_values)$p.value,
    error = function(e) NA_real_)
  
  data.table(
    Feature = feature,
    TCGA_mean_Normal = mean_normal,
    TCGA_mean_Tumor = mean_tumor,
    TCGA_deltaBeta = delta_beta,
    TCGA_wilcox_p = p_value)}
```

TCGA delta beta is retained as external confirmation and is not included in the 60%–40% priority score.

This chunk applies the TCGA delta-beta function to all SVM features and performs BH multiple-testing correction.

```{r calculate-tcga-delta-all-dmrs}
tcga_delta_table <- rbindlist(
  lapply(features_svm, calc_tcga_delta),
  fill = TRUE
)

tcga_delta_table[
  ,
  TCGA_FDR := p.adjust(TCGA_wilcox_p, method = "BH")
]

if (nrow(tcga_delta_table) != expected_n_dmrs) {
  stop("The TCGA delta-beta table does not contain 110 DMRs.")
}

if (anyDuplicated(tcga_delta_table$Feature) > 0) {
  stop("Duplicated Feature values were found in the TCGA table.")
}

tcga_numeric_cols <- c(
  "TCGA_mean_Normal", "TCGA_mean_Tumor",
  "TCGA_deltaBeta", "TCGA_wilcox_p", "TCGA_FDR"
)

if (any(!is.finite(as.matrix(
  tcga_delta_table[, ..tcga_numeric_cols]
)))) {
  stop("The TCGA delta-beta table contains non-finite values.")
}

n_nonpositive_tcga <- sum(
  tcga_delta_table$TCGA_deltaBeta <= 0
)

if (n_nonpositive_tcga > 0) {
  warning(
    n_nonpositive_tcga,
    " DMR(s) have non-positive TCGA delta beta."
  )
}

fwrite(
  tcga_delta_table,
  file.path(outdir, "TCGA_full_deltaBeta_all_110_DMRs.csv")
)

head(tcga_delta_table)
```

The exported table summarises external TCGA methylation differences for all 110 DMRs.

This chunk merges SHAP, TCGA, discovery, validation, and background evidence into one table and moves the key columns to the front.

```{r assemble-priority-evidence-table}
priority_table <- merge(
  final_priority_table,
  tcga_delta_table,
  by = "Feature",
  all.x = TRUE,
  sort = FALSE
)

if (nrow(priority_table) != expected_n_dmrs) {
  stop("The priority evidence merge changed the expected row count.")
}

if (anyDuplicated(priority_table$Feature) > 0) {
  stop("Duplicated features were found in the priority table.")
}

required_priority_values <- c(
  "mean_abs_SHAP",
  "discovery_deltaBeta",
  "TCGA_deltaBeta"
)

if (anyNA(priority_table[, ..required_priority_values])) {
  stop("Required values are missing from the priority table.")
}
```

This chunk moves the key evidence fields to the front and exports the complete unscored priority table.

```{r order-and-export-priority-evidence}
key_cols <- intersect(
  c(
    "SHAP_rank", "Feature", "mean_abs_SHAP",
    "TCGA_mean_Normal", "TCGA_mean_Tumor",
    "TCGA_deltaBeta", "TCGA_wilcox_p", "TCGA_FDR",
    "discovery_deltaBeta", "validation_mean_deltaBeta",
    "legacy_discovery_deltaBeta", "discovery_no_cpgs",
    "discovery_coord", "DMR_coordinate",
    "chr", "start", "end", "DMR_length"
  ),
  names(priority_table)
)

setcolorder(
  priority_table,
  c(key_cols, setdiff(names(priority_table), key_cols))
)

fwrite(
  priority_table,
  file.path(
    outdir,
    "SHAP_TCGA_validation_background_all_110_DMRs__nsim1000.csv"
  )
)

dim(priority_table)
head(priority_table)
```

The resulting table is an unranked evidence table; the combined score is calculated in the next section.

## Calculate the corrected priority score

The scaling function converts each scoring component to the range 0–1 while handling missing or constant inputs.

```{r define-scaling-function}
priority_table_scored <- copy(priority_table)

scale01 <- function(x, variable_name) {
  x <- as.numeric(x)

  if (any(!is.finite(x))) {
    stop(
      variable_name,
      " contains NA, NaN, or infinite values."
    )
  }

  rng <- range(x)

  if (rng[1] == rng[2]) {
    warning(
      variable_name,
      " is constant; assigning a score of 0.5."
    )
    return(rep(0.5, length(x)))
  }

  (x - rng[1]) / (rng[2] - rng[1])
}
```

Scaling allows SHAP importance and discovery delta beta to be combined despite their different numerical ranges.

This chunk scales mean absolute SHAP and the corrected Step 1 discovery delta beta, calculates the 60%–40% score, and ranks all 110 DMRs.

```{r calculate-priority-components}
priority_table_scored[
  ,
  score_SHAP := scale01(
    mean_abs_SHAP,
    "mean_abs_SHAP"
  )
]

priority_table_scored[
  ,
  score_discovery_delta := scale01(
    discovery_deltaBeta,
    "discovery_deltaBeta"
  )
]

priority_table_scored[
  ,
  priority_score :=
    0.60 * score_SHAP +
    0.40 * score_discovery_delta
]

score_columns <- c(
  "score_SHAP",
  "score_discovery_delta",
  "priority_score"
)

score_matrix <- as.matrix(
  priority_table_scored[, ..score_columns]
)

if (any(!is.finite(score_matrix))) {
  stop("The priority score table contains non-finite values.")
}

if (any(score_matrix < 0 | score_matrix > 1)) {
  stop("A scaled or combined score falls outside the 0-1 range.")
}
```

This chunk orders all 110 candidates by the corrected combined score and verifies that the rank sequence is complete.

```{r rank-priority-table}
setorder(
  priority_table_scored,
  -priority_score,
  -mean_abs_SHAP,
  -discovery_deltaBeta
)

priority_table_scored[
  ,
  priority_rank := seq_len(.N)
]

if (!identical(
  sort(priority_table_scored$priority_rank),
  seq_len(expected_n_dmrs)
)) {
  stop("The priority ranks are incomplete or duplicated.")
}
```

The final rank is a prioritisation measure rather than a p value or estimate of statistical significance.

This chunk compares the SHAP-only rank with the combined priority rank and records the direction of each rank change.

```{r compare-shap-and-combined-ranks}
ranking_comparison <- priority_table_scored[
  ,
  .(
    Feature,
    SHAP_rank,
    priority_rank,
    mean_abs_SHAP,
    discovery_deltaBeta,
    validation_mean_deltaBeta,
    legacy_discovery_deltaBeta,
    discovery_no_cpgs,
    score_SHAP,
    score_discovery_delta,
    priority_score
  )
]
## A positive value means that the DMR ranks better in the
## combined ranking than in the SHAP-only ranking.
ranking_comparison[, rank_change :=
                     SHAP_rank - priority_rank]
setorder(ranking_comparison, priority_rank)
head(ranking_comparison, 20)
```

A positive rank change means that discovery delta beta improved the DMR's position relative to SHAP-only ranking.

This chunk exports the complete evidence table, the corrected scored table, the rank comparison, and the final top 10 DMR panel.

```{r export-ranked-results}
fwrite(
  priority_table,
  file.path(
    outdir,
    "SHAP_TCGA_validation_background_all_110_DMRs__nsim1000.csv"
  )
)

fwrite(
  priority_table_scored,
  file.path(
    outdir,
    "SHAP_discovery_delta_scored_all_110_DMRs__nsim1000.csv"
  )
)

fwrite(
  ranking_comparison,
  file.path(
    outdir,
    "SHAP_vs_combined_priority_ranking_comparison__nsim1000.csv"
  )
)

top10_dmrs <- priority_table_scored[
  priority_rank <= expected_top_n
]

if (nrow(top10_dmrs) != expected_top_n) {
  stop("The final panel does not contain exactly 10 DMRs.")
}

if (anyDuplicated(top10_dmrs$Feature) > 0) {
  stop("Duplicated features were found in the final top 10.")
}

if (anyNA(top10_dmrs$discovery_deltaBeta)) {
  stop("The final top 10 contains missing discovery delta beta.")
}

fwrite(
  top10_dmrs,
  file.path(
    outdir,
    "Final_top10_DMRs_for_wetlab_validation__nsim1000.csv"
  )
)
```

This chunk records the main quality-control results in a compact CSV file.

```{r export-pipeline-qc-summary}
pipeline_qc_summary <- data.table(
  QC_item = c(
    "Final DMR count",
    "Exact discovery matches",
    "SVM feature count",
    "TCGA Normal samples",
    "TCGA Tumor samples",
    "SHAP sample count",
    "SHAP feature count",
    "Final top DMR count",
    "Legacy equals validation mean",
    "Corrected discovery equals validation mean"
  ),
  Value = c(
    nrow(final_dmrs),
    sum(!is.na(final_dmrs$discovery_deltaBeta)),
    length(features_svm),
    unname(group_counts["Normal"]),
    unname(group_counts["Tumor"]),
    nrow(shap_svm),
    ncol(shap_svm),
    nrow(top10_dmrs),
    legacy_matches_validation,
    corrected_matches_validation
  )
)

fwrite(
  pipeline_qc_summary,
  file.path(outdir, "step5_pipeline_QC_summary_nsim1000.csv")
)

print(pipeline_qc_summary)
```

The final top 10 file is the primary output for subsequent wet-lab feasibility assessment.

# Visualisation of the final top 10 DMRs

This chunk extracts the final top 10 DMRs and creates consistent rank-based labels for all figures.

```{r prepare-top10-labels}
library(data.table)
library(ggplot2)
library(ggbeeswarm)
top10_dmrs <- copy(priority_table_scored[
    priority_rank <= 10])

## Arrange the selected DMRs from rank 1 to rank 10.
setorder(top10_dmrs, priority_rank)
 
## Create a consistent label for all plots
top10_dmrs[, DMR_label :=paste0("Rank ",priority_rank,": ",Feature)]

## Reverse the factor levels so that rank 1 appears at the
## top of horizontal bar plots and SHAP summary plots.

top10_dmrs[, DMR_label :=factor(DMR_label,levels = rev(DMR_label))]
```

Reversing the factor levels places Rank 1 at the top of horizontal plots.

## Combined priority score

This chunk constructs a horizontal barplot showing the combined priority score for the final top 10 DMRs.

```{r build-priority-barplot}
p_priority_bar <- ggplot(
  top10_dmrs,
  aes(x = priority_score,
    y = DMR_label)) +
  geom_col(
    width = 0.7,
    fill = "grey40"
  ) +
  geom_text(
    aes(
      label = sprintf("%.3f", priority_score)
    ),
    hjust = -0.12,
    size = 3.5
  ) +
  expand_limits(
    x = max(
      top10_dmrs$priority_score,
      na.rm = TRUE
    ) * 1.15
  ) +
  labs(
    title = "Final top 10 DMRs by combined priority score nsim1000",
    subtitle = "60% SHAP importance + 40% discovery delta beta",
    x = "Combined priority score",
    y = "DMR"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0.5
    ),
    axis.text.y = element_text(size = 9),
    panel.grid.major.y = element_blank()
  )
```

Longer bars indicate higher combined priority after integrating SHAP importance and discovery delta beta.

This chunk displays and saves the combined priority score plot in PNG and PDF formats.

```{r save-priority-barplot}
print(p_priority_bar)
ggsave(filename = file.path(outdir,
      "final_top10_combined_priority_score_barplot_nsim1000.png"),
  plot = p_priority_bar,
  width = 10,
  height = 7,
  dpi = 300)

ggsave(
  filename = file.path(outdir,
 "final_top10_combined_priority_score_barplot_nsim1000.pdf"),
  plot = p_priority_bar,
  width = 10,
  height = 7)
```

The files can be used in reports, presentations, and manuscript figures.

## Weighted score components

This chunk calculates the actual 60% SHAP and 40% discovery-delta contributions and confirms that their sum reproduces the priority score.

```{r calculate-weighted-components}
top10_dmrs[, weighted_SHAP :=0.60 * score_SHAP]

top10_dmrs[, weighted_discovery_delta :=0.40 * score_discovery_delta]

############################################################
## Confirm that the two contributions reproduce the score
############################################################
top10_dmrs[, score_check :=weighted_SHAP +weighted_discovery_delta]

cat(
  "\nMaximum difference between priority_score and score_check:\n"
)

max_score_difference <- max(
  abs(
    top10_dmrs$priority_score -
      top10_dmrs$score_check
  )
)

print(max_score_difference)

if (max_score_difference > 1e-12) {
  stop("The weighted components do not reproduce priority_score.")
}
```

The maximum difference should be zero or extremely close to zero because of floating-point precision.

This chunk converts the two weighted score components to long format and assigns presentation labels.

```{r reshape-score-components}
component_plot_df <- melt(
  top10_dmrs,
  id.vars = c(
    "Feature",
    "priority_rank",
    "DMR_label",
    "priority_score"
  ),
  measure.vars = c(
    "weighted_SHAP",
    "weighted_discovery_delta"
  ),
  variable.name = "Score_component",
  value.name = "Weighted_contribution"
)

component_plot_df[
  ,
  Score_component := factor(
    Score_component,
    levels = c(
      "weighted_SHAP",
      "weighted_discovery_delta"
    ),
    labels = c(
      "SHAP contribution (60%)",
      "Discovery delta contribution (40%)"))]
```

The long-format table is used to draw a stacked contribution plot.

This chunk creates the stacked barplot showing how much each scoring component contributes to the final score.

```{r build-component-barplot}
p_component_bar <- ggplot(
  component_plot_df,
  aes(
    x = Weighted_contribution,
    y = DMR_label,
    fill = Score_component
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    data = top10_dmrs,
    aes(
      x = priority_score,
      y = DMR_label,
      label = sprintf("%.3f", priority_score)
    ),
    inherit.aes = FALSE,
    hjust = -0.12,
    size = 3.4
  ) +
  scale_fill_manual(
    values = c(
      "SHAP contribution (60%)" = "#4C78A8",
      "Discovery delta contribution (40%)" = "#F58518"
    )
  ) +
  expand_limits(
    x = max(top10_dmrs$priority_score, na.rm = TRUE) * 1.15
  ) +
  labs(
    title = "Contribution to the final priority score nsim1000",
    subtitle = "Weighted SHAP and discovery delta-beta components",
    x = "Weighted contribution",
    y = "DMR",
    fill = "Score component"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    legend.position = "top"
  )
```

The blue and orange segments sum to the total priority score displayed at the end of each bar.

This chunk displays and saves the weighted score-component plot in PNG and PDF formats.

```{r save-component-barplot}
print(p_component_bar)
ggsave(
  filename = file.path(outdir,"final_top10_priority_score_components_nsim1000.png"),
  plot = p_component_bar,
  width = 10,
  height = 7,
  dpi = 300)

ggsave(
  filename = file.path(outdir,"final_top10_priority_score_components_nsim1000.pdf"),
  plot = p_component_bar,
  width = 10,
  height = 7)
```

The plot makes it possible to distinguish DMRs driven mainly by SHAP importance from those supported strongly by discovery effect size.

## SHAP summary beeswarm

This chunk identifies the selected top 10 features, constructs the rank-label mapping, and defines sample identifiers.

```{r prepare-shap-feature-labels}
selected_features_shap <- as.character( top10_dmrs$Feature)
missing_shap_features <- setdiff(
  selected_features_shap,
  colnames(shap_svm)
)

if (length(missing_shap_features) > 0) {
  stop(
    "Top-10 features missing from the SHAP matrix: ",
    paste(missing_shap_features, collapse = ", ")
  )
}

selected_features_shap <- selected_features_shap[
  selected_features_shap %in% colnames(shap_svm)
]

############################################################
## Create the feature-to-label mapping
############################################################

selected_label_map_shap <- top10_dmrs[Feature %in% selected_features_shap,
  .(Feature,
    priority_rank,
    DMR_label)]

selected_label_map_shap[
  ,
  DMR_label := as.character(DMR_label)]
############################################################
## Create sample identifiers
############################################################

sample_ids_shap <- rownames(x_explain)

if (is.null(sample_ids_shap)) {
  sample_ids_shap <- paste0(
    "Sample_",
    seq_len(nrow(x_explain))
  )
}
```

Only features present in the SHAP matrix are retained for the beeswarm plot.

This chunk converts the selected SHAP values from wide to long format.

```{r reshape-shap-values}
shap_dt <- as.data.table(
  shap_svm[
    ,
    selected_features_shap,
    drop = FALSE
  ]
)

shap_dt[, Sample := sample_ids_shap]

shap_long <- melt(
  shap_dt,
  id.vars = "Sample",
  measure.vars = selected_features_shap,
  variable.name = "Feature",
  value.name = "SHAP_value"
)
```

Each row of the long table represents one sample–DMR SHAP contribution.

This chunk converts the corresponding TCGA DMR beta values to long format.

```{r reshape-shap-beta-values}
beta_dt <- as.data.table(
  as.data.frame(x_explain)[
    ,
    selected_features_shap,
    drop = FALSE
  ]
)

beta_dt[, Sample := sample_ids_shap]

beta_long <- melt(
  beta_dt,
  id.vars = "Sample",
  measure.vars = selected_features_shap,
  variable.name = "Feature",
  value.name = "DMR_beta"
)
```

These beta values provide the colour scale used in the SHAP summary plot.

This chunk joins the sample-level SHAP contributions with their corresponding DMR beta values.

```{r merge-shap-and-beta}
shap_plot_df <- merge(
  shap_long,
  beta_long,
  by = c(
    "Sample",
    "Feature"
  ),
  all.x = TRUE
)
```

The merged table links model contribution and methylation level for every selected sample–DMR pair.

This chunk adds the combined-priority rank and display label to the SHAP plotting table.

```{r add-shap-rank-labels}
shap_plot_df <- merge(
  shap_plot_df,
  selected_label_map_shap[
    ,
    .(
      Feature,
      priority_rank,
      DMR_label
    )
  ],
  by = "Feature",
  all.x = TRUE
)
```

The rank label ensures that the beeswarm order follows the final corrected priority ranking.

This chunk fixes the plotting order so that Rank 1 appears at the top.

```{r order-shap-labels}
label_order_shap <- selected_label_map_shap[
  order(priority_rank),
  DMR_label
]

shap_plot_df[
  ,
  DMR_label := factor(
    DMR_label,
    levels = rev(label_order_shap)
  )
]
```

The factor order is shared across all points belonging to the same DMR.

This chunk constructs the SHAP summary beeswarm. Horizontal position represents contribution to predicted Tumor probability, and colour represents the DMR beta value.

```{r build-shap-beeswarm}
p_shap_beeswarm <- ggplot(
  shap_plot_df,
  aes(
    x = SHAP_value,
    y = DMR_label,
    color = DMR_beta
  )
) +
  ggbeeswarm::geom_quasirandom(
    groupOnX = FALSE,
    alpha = 0.75,
    size = 1.2,
    width = 0.25
  ) +
  scale_color_gradient(
    low = "#2C7BB6",
    high = "#D7191C"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.5
  ) +
  labs(
    title = "SHAP summary plot for the final top 10 DMRs nsim1000",
    subtitle = "Top 10 selected using SHAP importance and discovery delta beta",
    x = "SHAP value for predicted Tumor probability",
    y = "DMR",
    color = "DMR beta"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0.5
    ),
    axis.text.y = element_text(size = 9)
  )
```

Red points at positive SHAP values indicate high methylation values that push predictions toward Tumor.

This chunk displays and saves the SHAP summary beeswarm in PNG and PDF formats.

```{r save-shap-beeswarm}
print(p_shap_beeswarm)
ggsave(filename = file.path(outdir,"final_top10_SHAP_summary_beeswarm_nsim1000.png"),
  plot = p_shap_beeswarm,
  width = 10,
  height = 7,
  dpi = 300)

ggsave(filename = file.path(outdir,"final_top10_SHAP_summary_beeswarm_nsim1000.pdf"),
  plot = p_shap_beeswarm,
  width = 10,
  height = 7)
```

The plot provides sample-level direction and magnitude information beyond the mean absolute SHAP ranking.

## TCGA Tumor–Normal beta distributions

This chunk identifies the top 10 features present in the TCGA dataframe and creates their rank-label mapping.

```{r prepare-beta-feature-labels}
selected_features_beta <- as.character(
  top10_dmrs$Feature)

selected_features_beta <- selected_features_beta[
  selected_features_beta %in% colnames(test_df_full)]

missing_beta_features <- setdiff(
  as.character(top10_dmrs$Feature),
  selected_features_beta
)

if (length(missing_beta_features) > 0) {
  stop(
    "Top-10 features missing from the TCGA plotting data: ",
    paste(missing_beta_features, collapse = ", ")
  )
}

############################################################
## Create feature-to-label mapping
############################################################

selected_label_map_beta <- top10_dmrs[
  Feature %in% selected_features_beta,
  .(
    Feature,
    priority_rank,
    DMR_label
  )
]

selected_label_map_beta[
  ,
  DMR_label := as.character(DMR_label)
]
```

Any missing selected feature would be recorded before plotting.

This chunk prepares sample identifiers and converts the selected TCGA beta values to long format.

```{r reshape-tcga-beta-data}
test_df_plot <- as.data.frame(test_df_full)

sample_ids_beta <- rownames(test_df_plot)

if (is.null(sample_ids_beta)) {
  sample_ids_beta <- paste0(
    "Sample_",
    seq_len(nrow(test_df_plot))
  )
}

beta_selected_dt <- as.data.table(
  test_df_plot[
    ,
    c(
      "group",
      selected_features_beta
    ),
    drop = FALSE
  ]
)

beta_selected_dt[, Sample := sample_ids_beta]

## Convert TCGA beta data to long format
beta_selected_long <- melt(
  beta_selected_dt,
  id.vars = c(
    "Sample",
    "group"
  ),
  measure.vars = selected_features_beta,
  variable.name = "Feature",
  value.name = "DMR_beta"
)
```

The long table contains one DMR beta value for each sample, feature, and diagnostic group.

This chunk adds the corrected priority labels, fixes the Normal–Tumor group order, and arranges plot facets from Rank 1 to Rank 10.

```{r add-beta-rank-labels}
beta_selected_long <- merge(
  beta_selected_long,
 selected_label_map_beta[,.(
      Feature,
      priority_rank,
      DMR_label)],
  by = "Feature",
  all.x = TRUE)

## Set the group order
beta_selected_long[,group := factor(as.character(group),levels = c("Normal","Tumor"))]
## Arrange facets from priority rank 1 to rank 10
label_order_beta <- selected_label_map_beta[
  order(priority_rank),
  DMR_label]

beta_selected_long[,DMR_label := factor(DMR_label,levels = label_order_beta)]
```

The prepared dataframe is ready for direct visual comparison of methylation distributions.

This chunk constructs faceted boxplots with beeswarm points for Normal and Tumor samples in the TCGA full cohort.

```{r build-tcga-beta-plot}
p_beta_box_beeswarm <- ggplot(
  beta_selected_long,
  aes(
    x = group,
    y = DMR_beta,
    color = group,
    fill = group
  )
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.25,
    width = 0.55
  ) +
  ggbeeswarm::geom_quasirandom(
    width = 0.18,
    alpha = 0.55,
    size = 1.1
  ) +
  facet_wrap(~ DMR_label, ncol = 2, scales = "free_y") +
  scale_color_manual(
    values = c("Normal" = "#2C7BB6", "Tumor" = "#D7191C")
  ) +
  scale_fill_manual(
    values = c("Normal" = "#2C7BB6", "Tumor" = "#D7191C")
  ) +
  labs(
    title = "TCGA confirmation of the final top 10 DMRs nsim1000",
    subtitle = "Final candidates selected using SHAP and discovery delta beta",
    x = NULL,
    y = "DMR mean beta value",
    color = "Group",
    fill = "Group"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    strip.text = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    legend.position = "top"
  )
```

Clear upward shifts in the Tumor group support the external reproducibility of the selected hypermethylated DMRs.

This chunk displays and saves the TCGA beta distribution figure in PNG and PDF formats.

```{r save-tcga-beta-plot}
print(p_beta_box_beeswarm)
## Save the TCGA beta boxplot and beeswarm
ggsave(filename = file.path(outdir,"final_top10_TCGA_beta_boxplot_beeswarm_nsim1000.png"),
  plot = p_beta_box_beeswarm,
  width = 11,
  height = 12,
  dpi = 300
)

ggsave(filename = file.path(outdir,"final_top10_TCGA_beta_boxplot_beeswarm_nsim1000.pdf"),
  plot = p_beta_box_beeswarm,
  width = 11,
  height = 12)
```

The figure provides an external visual check of the Tumor–Normal separation for each final candidate.

This chunk saves the corrected ranked tables, plotting dataframes, and figure objects in one RData file.

```{r save-analysis-objects}
save(
  final_dmrs,
  delta_source_audit,
  pipeline_qc_summary,
  shap_svm,
  shap_importance,
  tcga_delta_table,
  priority_table,
  priority_table_scored,
  ranking_comparison,
  top10_dmrs,
  component_plot_df,
  shap_plot_df,
  beta_selected_long,
  p_priority_bar,
  p_component_bar,
  p_shap_beeswarm,
  p_beta_box_beeswarm,
  file = file.path(
    outdir,
    "final_top10_DMR_visualisation_results_full_nsim1000.Rdata"
  )
)
```

The saved object bundle allows the figures and result tables to be reopened without repeating the SHAP calculation.


# Interpretation note

The TCGA full cohort is used both for SHAP-based prioritisation and for visualising Tumor–Normal methylation differences. It should therefore be described as an external prioritisation cohort rather than as a fully independent validation cohort for the final top 10 panel. Independent confirmation will come from subsequent wet-lab testing or another cohort not used during candidate selection.

# Session information

The following output records the R version, operating system, and package versions used to generate the report.

```{r session-information}
sessionInfo()
```
